---
title: "scripts for variant calling"
author: "subhash"
date: "2024-06-04"
data: "whole genome sequencing of sorghum association panel - trimmed reads from #400 individuals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/work/jyanglab/subhash/variant_calling")
```



```{r}
PI <- read.table("accession_ID.txt", sep = "\t", header = TRUE)
PI_numbers = PI[, 1]
PI_numbers
```

### using  fastqc to generate quality report for all the individuals for both farward and revere end files

```{r}
template <-
'#!/bin/sh
#SBATCH --ntasks-per-node=8
#SBATCH --nodes=1
#SBATCH --mem=1G
#SBATCH --time=24:00:00
#SBATCH --job-name=fastqc_{PI}
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu
#SBATCH --mail-type=ALL
#SBATCH --error=/work/jyanglab/subhash/variant_calling/1.trimmed_data/0.1_fastqc_report/0.scripts/{PI}_fastqc.err
#SBATCH --output=/work/jyanglab/subhash/variant_calling/1.trimmed_data/0.1_fastqc_report/0.scripts/{PI}_fastqc.out

ml fastqc
fastqc -o /work/jyanglab/subhash/variant_calling/1.trimmed_data/0.1_fastqc_report/ /work/jyanglab/subhash/variant_calling/1.trimmed_data/{PI}_1.fastq.gz

fastqc -o /work/jyanglab/subhash/variant_calling/1.trimmed_data/0.1_fastqc_report/ /work/jyanglab/subhash/variant_calling/1.trimmed_data/{PI}_2.fastq.gz
'
# Initialize an empty vector to store template commands
fastq <- c()
# Function to generate SLURM script for each PI number
generate_script <- function(PI) {
  script <- gsub("\\{PI\\}", PI, template)
  file_name <- paste0("fastqc_", PI, ".sh")
  writeLines(script, con = file_name)
  # Add sbatch command to the vector
  fastq  <<- c(fastq, paste0("sbatch -p guest ", file_name))
}
# Generate scripts for each PI number
sapply(PI_numbers, generate_script)
# Write sbatch commands to a txt file - sh .txt
writeLines(fastq, con = "fastqc.txt")
```




### step - 1
#R code to generate multiple slurm files to submit multiple jobs for each sample to align trimmed read files with refrence genome using bwa aligner....

```{r}
setwd("/work/jyanglab/subhash/variant_calling/3.alignment/0.scripts")
# Define the template for the SLURM script
template <-
'#!/bin/sh
#SBATCH --ntasks-per-node=8
#SBATCH --nodes=1
#SBATCH --mem=32G
#SBATCH --time=40:00:00
#SBATCH --job-name=alignment_{PI}
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu
#SBATCH --mail-type=ALL
#SBATCH --error=/work/jyanglab/subhash/variant_calling/3.alignment/0.1.log_files/{PI}_bwa.err
#SBATCH --output=/work/jyanglab/subhash/variant_calling/3.alignment/0.1.log_files/{PI}_bwa.out

ml bwa
ml samtools
ref="/work/jyanglab/subhash/variant_calling/2.refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.fa"
reads="/work/jyanglab/subhash/variant_calling/1.trimmed_data"
out="/work/jyanglab/subhash/variant_calling/3.alignment"
bwa mem -t 8 -R "@RG\\tID:{PI}\\tPL:ILLUMINA\\tSM:{PI}" ${ref} ${reads}/{PI}_1.fastq.gz ${reads}/{PI}_2.fastq.gz|samtools view --threads 8 -bS -f 2 -q 30 -h - | samtools sort --threads 8 -o ${out}/{PI}.srt.bam -'
# Initialize an empty vector to store sbatch commands
alignment <- c()
# Function to generate SLURM script for each PI number
generate_script <- function(PI) {
  script <- gsub("\\{PI\\}", PI, template)
  file_name <- paste0("bwa_", PI, ".sh")
  writeLines(script, con = file_name)
  # Add sbatch command to the vector
  alignment <<- c(alignment, paste0("sbatch -p guest ", file_name))
}
# Generate scripts for each PI number
sapply(PI_numbers, generate_script)
# Write sbatch commands to a txt file
writeLines(alignment, con = "alignment.txt")
```

### step - 2
#marking the duplicates using picard tool....

```{r}
setwd("/work/jyanglab/subhash/variant_calling/4.picard/0.scripts")
# Define the template for the SLURM script
template <-
'#!/bin/sh
#SBATCH --ntasks-per-node=8
#SBATCH --nodes=1
#SBATCH --mem=10G
#SBATCH --time=24:00:00
#SBATCH --job-name=picard_{PI}
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu
#SBATCH --mail-type=ALL
#SBATCH --error=/work/jyanglab/subhash/variant_calling/4.picard/0.log_files/{PI}_picard.err
#SBATCH --output=/work/jyanglab/subhash/variant_calling/4.picard/0.log_files/{PI}_picard.out

ml picard/3.0
picard  MarkDuplicates \\
INPUT=/work/jyanglab/subhash/variant_calling/3.alignment/{PI}.srt.bam \\
OUTPUT=/work/jyanglab/subhash/variant_calling/4.picard/{PI}_picard_dedup.bam \\
METRICS_FILE=/work/jyanglab/subhash/variant_calling/4.picard/{PI}_picard_metrics.txt \\
CREATE_INDEX=true \\
REMOVE_DUPLICATES=true
'
# Initialize an empty vector to store sbatch commands
picard <- c()
# Function to generate SLURM script for each PI number
generate_script <- function(PI) {
  script <- gsub("\\{PI\\}", PI, template)
  file_name <- paste0("picard_", PI, ".sh")
  writeLines(script, con = file_name)
  # Add sbatch command to the vector
  picard <<- c(picard, paste0("sbatch -p guest ", file_name))
}
# Generate scripts for each PI number
sapply(PI_numbers, generate_script)
# Write sbatch commands to a txt file
writeLines(picard, con = "picard.txt")
```

#step - 3
#skipped the base recalibration quality score step since we dont have previous called variants...
### calling variants for the entire genome using GATK haplotype caller...

```{r}
#setwd("/work/jyanglab/subhash/variant_calling/5.gvcf/0.scripts")
#Define the template for the SLURM script
#10 GB is enough to do work increase the time....
template <-
'#!/bin/sh
#SBATCH --ntasks-per-node=8
#SBATCH --nodes=1
#SBATCH --mem=10G  
#SBATCH --time=50:00:00
#SBATCH --job-name=gvcf_{PI}
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu
#SBATCH --mail-type=ALL
#SBATCH --error=/work/jyanglab/subhash/variant_calling/5.gvcf/log_files/{PI}_gvcf.err
#SBATCH --output=/work/jyanglab/subhash/variant_calling/5.gvcf/log_files/{PI}_gvcf.out

ml gatk4/4.4
gatk HaplotypeCaller \\
    -R /work/jyanglab/subhash/variant_calling/2.refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.fa \\
    -I /work/jyanglab/subhash/variant_calling/4.picard/{PI}_picard_dedup.bam \\
    -O /work/jyanglab/subhash/variant_calling/5.gvcf/{PI}.g.vcf \\
    -ERC GVCF \\
    --native-pair-hmm-threads 8
'
# Initialize an empty vector to store sbatch commands
gvcf <- c()
# Function to generate SLURM script for each PI number
generate_script <- function(PI) {
  script <- gsub("\\{PI\\}", PI, template)
  file_name <- paste0("gvcf_", PI, ".sh")
  writeLines(script, con = file_name)
  # Add sbatch command to the vector
  gvcf <<- c(gvcf, paste0("sbatch -p guest ", file_name))
}
# Generate scripts for each PI number
sapply(PI_numbers, generate_script)
# Write sbatch commands to a txt file
writeLines(gvcf, con = "gvcf.txt")

```

### creating a genome database for every 5MB window for the entire genome 

```{r}
# Define the chromosome lengths - from the refrence file these are the lengths of each chromosome.
chromosome_lengths <- list(
  Chr01 = 85112863,
  Chr02 = 79114963,
  Chr03 = 80873341,
  Chr04 = 71215609,
  Chr05 = 77058072,
  Chr06 = 62713908,
  Chr07 = 68911884,
  Chr08 = 65779274,
  Chr09 = 63277606,
  Chr10 = 62870657
)
interval_size <- 5000000  # 5 MB intervals
# PI numbers
#PI_numbers <- c("PI152651", "PI17548", "PI24969", "PI276837", "PI564163")
# Initialize a vector to store the names of the generated .sh files
sh_file_names <- c()
# Function to generate SLURM and GATK commands for each interval
generate_script <- function(chrom, start, end, PI_numbers) {
  start_formatted <- format(start, scientific = FALSE)
  end_formatted <- format(end, scientific = FALSE)
  interval <- paste0(chrom, ":", start_formatted, "-", end_formatted)
  db_path <- paste0(chrom, "_", start_formatted, "_", end_formatted, ".database")
  output_vcf <- paste0(chrom, "_", start_formatted, "_", end_formatted, ".vcf")
  job_name <- paste0("db_", chrom, "_", start_formatted, "_", end_formatted)
  err_file <- paste0("/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/1.database_5mb/0.log_files/", chrom, "_", start_formatted, "_", end_formatted, "_db_vcf.err")
  out_file <- paste0("/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/1.database_5mb/0.log_files/", chrom, "_", start_formatted, "_", end_formatted, "_db_vcf.out")
  # Generate the -V arguments
  v_arguments <- paste0("-V ${gvcf_path}/", PI_numbers, ".g.vcf", collapse = " \\\n  ")
  script_content <- paste(
    "#!/bin/sh",
    "#SBATCH --ntasks-per-node=8",
    "#SBATCH --nodes=1",
    "#SBATCH --mem=150G",
    "#SBATCH --time=80:00:00",
    paste0("#SBATCH --job-name=", job_name),
    "#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu",
    "#SBATCH --mail-type=ALL",
    paste0("#SBATCH --error=", err_file),
    paste0("#SBATCH --output=", out_file),
    "",
    "gvcf_path=\"/work/jyanglab/subhash/variant_calling/5.gvcf\"",
    #"out=\"/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/2.vcf_files\"",
    "db=\"/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/1.database_5mb\"",
    "",
    "ml gatk4/4.4",
    "",
    paste0("gatk --java-options '-Xms10g -Xmx120g' GenomicsDBImport \\"),
    paste0("  ", v_arguments, " \\"),
    paste0("  --genomicsdb-workspace-path ", "${db}/", db_path, " \\"),
    paste0("  --intervals ", interval),
    sep = "\n")
  
  file_name <- paste0(chrom, "_", start_formatted, "_", end_formatted, "db.sh")
  writeLines(script_content, file_name)
  
  # Adding the sbatch command with file name to the vector
  sh_file_names <<- c(sh_file_names, paste0("sbatch -p guest ", file_name))
}

# Generate scripts for each chromosome
for (chrom in names(chromosome_lengths)) {
  length <- chromosome_lengths[[chrom]]
  for (start in seq(1, length, by = interval_size)) {
    end <- min(start + interval_size - 1, length)
    generate_script(chrom, start, end, PI_numbers)
  }
}
# Write the names of the generated .sh files to a .txt file
writeLines(sh_file_names, "db_vcf_sh_files.txt")

```

### from the created database I have called variants for every 5Mb interval....

```{r}
# Define the chromosome lengths
chromosome_lengths <- list(
  Chr01 = 85112863,
  Chr02 = 79114963,
  Chr03 = 80873341,
  Chr04 = 71215609,
  Chr05 = 77058072,
  Chr06 = 62713908,
  Chr07 = 68911884,
  Chr08 = 65779274,
  Chr09 = 63277606,
  Chr10 = 62870657
)
interval_size <- 5000000  # 5 MB intervals
# Initialize a vector to store the names of the generated .sh files
sh_file_names <- c()
# Function to generate SLURM and GATK commands for each interval
generate_script <- function(chrom, start, end, PI_numbers) {
  start_formatted <- format(start, scientific = FALSE)
  end_formatted <- format(end, scientific = FALSE)
  interval <- paste0(chrom, ":", start_formatted, "-", end_formatted)
  db_path <- paste0(chrom, "_", start_formatted, "_", end_formatted, ".database")
  output_vcf <- paste0(chrom, "_", start_formatted, "_", end_formatted, ".vcf.gz")
  job_name <- paste0("vcf_", chrom, "_", start_formatted, "_", end_formatted)
  err_file <- paste0("/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/2.vcf_files/0.log_files/", chrom, "_", start_formatted, "_", end_formatted, "_vcf_gz.err")
  out_file <- paste0("/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/2.vcf_files/0.log_files/", chrom, "_", start_formatted, "_", end_formatted, "_vcf_gz.out")
  # Generate the -V arguments
  v_arguments <- paste0("-V ${gvcf_path}/", PI_numbers, ".g.vcf", collapse = " \\\n  ")
  script_content <- paste(
    "#!/bin/sh",
    "#SBATCH --ntasks-per-node=8",
    "#SBATCH --nodes=1",
    "#SBATCH --mem=16G",
    "#SBATCH --time=80:00:00",
    paste0("#SBATCH --job-name=", job_name),
    "#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu",
    "#SBATCH --mail-type=ALL",
    paste0("#SBATCH --error=", err_file),
    paste0("#SBATCH --output=", out_file),
    "",
    #"gvcf_path=\"/work/jyanglab/subhash/variant_calling/5.gvcf\"",
    "out=\"/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/2.vcf_files\"",
    "db=\"/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/1.database_5mb\"",
    "",
    "ml gatk4/4.4",
    "",
    paste0("gatk --java-options '-Xms4g -Xmx10g' GenotypeGVCFs \\"),
    paste0("  -R /work/jyanglab/subhash/variant_calling/2.refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.fa\\"),
    paste0("  -V gendb://", "${db}/", db_path, " \\"),
    paste0("  -O ", "${out}/", output_vcf),
    sep = "\n")
  file_name <- paste0(chrom, "_", start_formatted, "_", end_formatted, "vcf.sh")
  writeLines(script_content, file_name)
  # Adding the sbatch command with file name to the vector
  sh_file_names <<- c(sh_file_names, paste0("sbatch -p guest ", file_name))
}
# Generate scripts for each chromosome
for (chrom in names(chromosome_lengths)) {
  length <- chromosome_lengths[[chrom]]
  for (start in seq(1, length, by = interval_size)) {
    end <- min(start + interval_size - 1, length)
    generate_script(chrom, start, end, PI_numbers)
  }
}
# Write the names of the generated .sh files to a .txt file
writeLines(sh_file_names, "vcf_sh_files.txt")
```

### creating RAW vcf file from each g.vcf file.

```{bash}
find $(pwd) -type f -name '*.vcf.gz' | sort -V > vcf_list.txt #creating a .txt fike with the path and names of all the vcf files
#using the .txt file
gatk --java-options "-Xmx10g" GatherVcfs \
     -I vcf_list.txt \
     -O /work/jyanglab/subhash/variant_calling/6.converting_to_vcf/3.merged_vcf/RAW_SAP.vcf.gz

```

##BQSR - Base Quality Score Recalibration creating g.vcf file

```{r}

setwd("/work/jyanglab/subhash/variant_calling/3.alignment/0.scripts")
# Define the template for the SLURM script
template <-
'#!/bin/sh
#SBATCH --ntasks-per-node=8
#SBATCH --nodes=1
#SBATCH --mem=16G
#SBATCH --time=80:00:00
#SBATCH --job-name=BQSR
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu
#SBATCH --mail-type=ALL
#SBATCH --error=/work/jyanglab/subhash/variant_calling/8.BQSR/0.log_files/{PI}_BQSR.err
#SBATCH --output=/work/jyanglab/subhash/variant_calling/8.BQSR/0.log_files/{PI}_BQSR.out
ml gatk4/4.1
ref="/work/jyanglab/subhash/variant_calling/2.refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.fa"
KS="/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/3.merged_vcf/RAW_SAP.vcf.gz"
IN="/work/jyanglab/subhash/variant_calling/4.picard/"
OUT1="/work/jyanglab/subhash/variant_calling/8.BQSR/1.data_table/"
OUT2="/work/jyanglab/subhash/variant_calling/8.BQSR/2.BQSR_bam/"
OUT3="/work/jyanglab/subhash/variant_calling/8.BQSR/3.g.vcf/"

gatk BaseRecalibrator -R ${ref} -I ${IN}{PI}_picard_dedup.bam -O ${OU1}{PI}_DT.recal_data.table --known-sites ${KS}
gatk ApplyBQSR -R ${ref} -I ${IN}{PI}_picard_dedup.bam --bqsr-recal-file ${OU1}{PI}_DT.recal_data.table -O ${OUT2}{PI}.BQSR.bam --create-output-bam-index true
gatk HaplotypeCaller -R ${ref} -I ${OUT2}{PI}.BQSR.bam -ERC GVCF -O ${OUT3}{PI}.g.vcf
'

# Initialize an empty vector to store sbatch commands
BQSR <- c()
# Function to generate SLURM script for each PI number
generate_script <- function(PI) {
  script <- gsub("\\{PI\\}", PI, template)
  file_name <- paste0("BQSR_", PI, ".sh")
  writeLines(script, con = file_name)
  # Add sbatch command to the vector
  BQSR <<- c(BQSR, paste0("sbatch -p guest ", file_name))
}
# Generate scripts for each PI number
sapply(PI_numbers, generate_script)
# Write sbatch commands to a txt file
writeLines(BQSR, con = "BQSR.txt")

```

### creating a genome database


```{r}
# Define the chromosome lengths
chromosome_lengths <- list(
  Chr01 = 85112863,
  Chr02 = 79114963,
  Chr03 = 80873341,
  Chr04 = 71215609,
  Chr05 = 77058072,
  Chr06 = 62713908,
  Chr07 = 68911884,
  Chr08 = 65779274,
  Chr09 = 63277606,
  Chr10 = 62870657
)
interval_size <- 5000000  # 5 MB intervals
# Initialize a vector to store the names of the generated .sh files
sh_file_names <- c()
# Function to generate SLURM and GATK commands for each interval
generate_script <- function(chrom, start, end, PI_numbers) {
  start_formatted <- format(start, scientific = FALSE)
  end_formatted <- format(end, scientific = FALSE)
  interval <- paste0(chrom, ":", start_formatted, "-", end_formatted)
  db_path <- paste0(chrom, "_", start_formatted, "_", end_formatted, ".database")
  output_vcf <- paste0(chrom, "_", start_formatted, "_", end_formatted, ".vcf")
  job_name <- paste0("dbBQSR_", chrom, "_", start_formatted, "_", end_formatted)
  err_file <- paste0("/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/2.database/0.log_files/", chrom, "_", start_formatted, "_", end_formatted, "_dbBQ_vcf.err")
  out_file <- paste0("/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/2.database/0.log_files/", chrom, "_", start_formatted, "_", end_formatted, "_dbBQ_vcf.out")
  # Generate the -V arguments
  v_arguments <- paste0("-V ${gvcf_path}/", PI_numbers, ".g.vcf", collapse = " \\\n  ")
  script_content <- paste(
    "#!/bin/sh",
    "#SBATCH --ntasks-per-node=8",
    "#SBATCH --nodes=1",
    "#SBATCH --mem=150G",
    "#SBATCH --time=80:00:00",
    paste0("#SBATCH --job-name=", job_name),
    "#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu",
    "#SBATCH --mail-type=ALL",
    paste0("#SBATCH --error=", err_file),
    paste0("#SBATCH --output=", out_file),
    "",
    "gvcf_path=\"/work/jyanglab/subhash/variant_calling/8.BQSR/3.g.vcf\"",
    #"out=\"/work/jyanglab/subhash/variant_calling/6.converting_to_vcf/2.vcf_files\"",
    "db=\"/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/2.database\"",
    "",
    "ml gatk4/4.4",
    "",
    paste0("gatk --java-options '-Xms10g -Xmx120g' GenomicsDBImport \\"),
    paste0("  ", v_arguments, " \\"),
    paste0("  --genomicsdb-workspace-path ", "${db}/", db_path, " \\"),
    paste0("  --intervals ", interval),
    
    #paste0("gatk --java-options '-Xms10g' GenotypeGVCFs \\"),
    #paste0("  -R /work/jyanglab/subhash/variant_calling/2.refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.f#\\"),
    #paste0("  -V gendb://", "${db}/", db_path, " \\"),
    #paste0("  -O ", "${out}/", output_vcf),
    sep = "\n")
  
  file_name <- paste0(chrom, "_", start_formatted, "_", end_formatted, "BQSRdb.sh")
  writeLines(script_content, file_name)
  
  # Append the sbatch command with file name to the vector
  sh_file_names <<- c(sh_file_names, paste0("sbatch -p guest ", file_name))
}

# Generate scripts for each chromosome
for (chrom in names(chromosome_lengths)) {
  length <- chromosome_lengths[[chrom]]
  for (start in seq(1, length, by = interval_size)) {
    end <- min(start + interval_size - 1, length)
    generate_script(chrom, start, end, PI_numbers)
  }
}
# Write the names of the generated .sh files to a .txt file
writeLines(sh_file_names, "db_BQSR_sh_files.txt")
```

### calling vcf from database

```{r}
# Define the chromosome lengths
chromosome_lengths <- list(
  Chr01 = 85112863,
  Chr02 = 79114963,
  Chr03 = 80873341,
  Chr04 = 71215609,
  Chr05 = 77058072,
  Chr06 = 62713908,
  Chr07 = 68911884,
  Chr08 = 65779274,
  Chr09 = 63277606,
  Chr10 = 62870657
)
interval_size <- 5000000  # 5 MB intervals
# Initialize a vector to store the names of the generated .sh files
sh_file_names <- c()
# Function to generate SLURM and GATK commands for each interval
generate_script <- function(chrom, start, end, PI_numbers) {
  start_formatted <- format(start, scientific = FALSE)
  end_formatted <- format(end, scientific = FALSE)
  interval <- paste0(chrom, ":", start_formatted, "-", end_formatted)
  db_path <- paste0(chrom, "_", start_formatted, "_", end_formatted, ".database")
  output_vcf <- paste0(chrom, "_", start_formatted, "_", end_formatted, ".vcf.gz")
  job_name <- paste0("vcf_", chrom, "_", start_formatted, "_", end_formatted)
  err_file <- paste0("/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/2.vcf_files/0.log_files/", chrom, "_", start_formatted, "_", end_formatted, "_vcf_gz.err")
  out_file <- paste0("/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/2.vcf_files/0.log_files/", chrom, "_", start_formatted, "_", end_formatted, "_vcf_gz.out")
  # Generate the -V arguments
  v_arguments <- paste0("-V ${gvcf_path}/", PI_numbers, ".g.vcf", collapse = " \\\n  ")
  script_content <- paste(
    "#!/bin/sh",
    "#SBATCH --ntasks-per-node=8",
    "#SBATCH --nodes=1",
    "#SBATCH --mem=16G",
    "#SBATCH --time=80:00:00",
    paste0("#SBATCH --job-name=", job_name),
    "#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu",
    "#SBATCH --mail-type=ALL",
    paste0("#SBATCH --error=", err_file),
    paste0("#SBATCH --output=", out_file),
    "",
    #"gvcf_path=\"/work/jyanglab/subhash/variant_calling/5.gvcf\"",
    "out=\"/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/2.vcf_files\"",
    "db=\"/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/2.database\"",
    "",
    "ml gatk4/4.4",
    "",
    #paste0("gatk --java-options '-Xms10g -Xmx120g' GenomicsDBImport \\"),
    #paste0("  ", v_arguments, " \\"),
    #paste0("  --genomicsdb-workspace-path ", "${db}/", db_path, " \\"),
    #paste0("  --intervals ", interval),
    
    paste0("gatk --java-options '-Xms4g -Xmx10g' GenotypeGVCFs \\"),
    paste0("  -R /work/jyanglab/subhash/variant_calling/2.refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.fa\\"),
    paste0("  -V gendb://", "${db}/", db_path, " \\"),
    paste0("  -O ", "${out}/", output_vcf),
    sep = "\n")
  
  file_name <- paste0(chrom, "_", start_formatted, "_", end_formatted, "vcf.sh")
  writeLines(script_content, file_name)
  
  # Append the sbatch command with file name to the vector
  sh_file_names <<- c(sh_file_names, paste0("sbatch -p guest ", file_name))
}

# Generate scripts for each chromosome
for (chrom in names(chromosome_lengths)) {
  length <- chromosome_lengths[[chrom]]
  for (start in seq(1, length, by = interval_size)) {
    end <- min(start + interval_size - 1, length)
    generate_script(chrom, start, end, PI_numbers)
  }
}
# Write the names of the generated .sh files to a .txt file
writeLines(sh_file_names, "vcf_sh_files.txt")
```


### creating RAW vcf file from each g.vcf file.

```{bash}
find $(pwd) -type f -name '*.vcf.gz' | sort -V > vcf_list.txt #creating a .txt fike with the path and names of all the vcf files
#using the .txt file
gatk --java-options "-Xmx10g" GatherVcfs \
     -I vcf_list.txt \
     -O /work/jyanglab/subhash/variant_calling/6.converting_to_vcf/3.merged_vcf/RAW_SAP.vcf.gz

```



```{bash}
ml bcftools
ml vcftools
ml plink2

#counting the number of variants from raw vcf file
bcftools view -H SAP.vcf.gz | wc -l

#changing ID to chrom_pos to compare manual and plink2.0 calculation
bcftools annotate --set-id +'%CHROM\_%POS' SAP.vcf.gz -Ov -o SAP_v5.vcf.gz

#calculating basic statistics
#--freq to calculate allele freequency
#--missing to calculate missing rate
#--hardy to calculate heterozygosity rate 

plink2 --bfile SAPo --freq alt1bins=0.01,0.02,0.03,0.04,0.05,0.1,0.2,0.3,0.4  --missing --hardy  --out /work/jyanglab/subhash/snp_sorghum/1.sap_genotype_files/3.basic_statistics_calculat

#filtering of vcf files using maf and missing rate from plink2.0
#--maf to filter minor allele freequency
#--geno to filter missing rate from bed- vcf file

plink2 --bfile SAP --maf 0.05 --geno 0.3 --make-bed --out /work/jyanglab/subhash/snp_sorghum/1.sap_genotype_files/4.filtered_SAP/filtered

#for filtering heterozygosity there is no direct function in plink2.0
# so I have obtained seperate file of observed heterozygosity from --hardy function using following code
#calculated heterozygosity using plink2.0 function -hardy which gives observed heterozygosity in column 8.

awk '$8 < 0.1 {print $2}' SAP.hardy > low_het_snps.txt

#now this low_het_snps.txt contains ID with <0.1 heterozygosity.
#here SAP filtered is  maf >0.05 and geno <30%

plink2 -bfile filtered -extract low_het_snps.txt -make-bed -out your_final_filtered_data

#this final filtered data contains variants with maf >/=0.05 and missing rate of <30% and heterozygosity with <0.1
```

### generating the graphs from basic statistics.

```{r}
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("data.table")
library(tidyverse)
library(ggplot2)
library(data.table)

#all the output files are otained from plink2.0    allele frequency = --freq , missing rate = --misssing  
#from PLINK2.0
#reading the data frame
var_freq <- read_delim("filt_stat.afreq", delim = "\t") 
#data
head(var_freq)
#calculating refrence allele frreequency since plink2.0 has calcuated only alternative allele
var_freq$REF_FREQS <- 1 - var_freq$ALT_FREQS
#minimum freequency of alt or ref
var_freq$maf <- pmin(var_freq$ALT_FREQS, var_freq$REF_FREQS)
#looking at the data frame
head(var_freq) 
# Confirming the minimum and maximum values of maf
min_maf <- min(var_freq$maf)
max_maf <- max(var_freq$maf)
print(paste("Minimum MAF:", min_maf))
print(paste("Maximum MAF:", max_maf))

# Using ggplot to create a histogram of minor allele frequency
maf <- ggplot(var_freq, aes(x = maf)) + 
  geom_histogram(fill = "brown3", color = "black") +
  labs(title = "Minor Allele Frequency_SAP_v5", x = "MAF", y = "Count") + 
  scale_x_continuous(limits = c(0.05, max_maf), breaks = seq(min_maf, by = 0.05)) +
  theme(axis.text.x = element_text(size = 14, face = "bold"),
        axis.text.y = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        axis.title.y = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 20, face = "bold"))
maf
ggsave("maf_bqsr_filt.png", plot = maf, width = 10, height = 8, dpi = 600)
#plotting the graph of missing rate from plink2.0
var_miss <- read_delim("filt_stat.vmiss", delim = "\t") #reading the data frame
head(var_miss)
#hist(var_miss$F_MISS,breaks = 50) #basic R plotting
vm <- ggplot(var_miss, aes(x = F_MISS)) + 
  geom_histogram(fill = "brown3", color = "black") +
  labs(title = "variant missing rate SAP_V5", x = "missing rate", y = "Count (Variants)") +
  theme(axis.text.x = element_text(size = 14, face = "bold"),
        axis.text.y = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        axis.title.y = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 20, face = "bold"))

vm
ggsave("vm_bqsr_filt.png", plot = vm, width = 10, height = 8, dpi = 600)

#plotting observed heterozygosity using plink2.0
het = fread("0.1_het.hardy")
het <- ggplot(het, aes(x = `O(HET_REF)`)) + 
  geom_histogram(fill = "brown3", color = "black") +
  labs(title = "Heterozygosity rate SAP_V5", x = "Heterozygosity rate", y = "Count (Variants)") +
  theme(axis.text.x = element_text(size = 14, face = "bold"),
        axis.text.y = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        axis.title.y = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 20, face = "bold"))
het
ggsave("het_bqsr_filt.png", plot = het, width = 10, height = 8, dpi = 600)
```



#### misc!

### to submit the jobs at a specified time

```{r}
# Define the job names
jobs <- c(
  "Chr01_1_5000000vcf.sh",
  "Chr01_5000001_10000000vcf.sh",
  "Chr01_10000001_15000000vcf.sh",
  "Chr01_15000001_20000000vcf.sh",
  "Chr01_20000001_25000000vcf.sh",
  "Chr01_25000001_30000000vcf.sh",
  "Chr01_30000001_35000000vcf.sh",
  "Chr01_35000001_40000000vcf.sh",
  "Chr01_40000001_45000000vcf.sh",
  "Chr01_45000001_50000000vcf.sh",
  "Chr01_50000001_55000000vcf.sh",
  "Chr01_55000001_60000000vcf.sh",
  "Chr01_60000001_65000000vcf.sh",
  "Chr01_65000001_70000000vcf.sh",
  "Chr01_70000001_75000000vcf.sh",
  "Chr01_75000001_80000000vcf.sh",
  "Chr01_80000001_85000000vcf.sh",
  "Chr01_85000001_85112863vcf.sh"
)

# Define the output file
output_file <- "chr01_vcf.txt"

# Open the file for writing
file_conn <- file(output_file, "w")

# Write the commands to the file
for (job in jobs) {
  command <- paste("sbatch -p guest --begin=2024-06-29T09:00:00", job)
  writeLines(command, file_conn)
}

# Close the file connection
close(file_conn)

# Print a message indicating the file has been created
cat("Commands have been written to", output_file, "\n")

```

