---
title: "Generate Geo, PCA, and Race-wise Figures"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## Purpose

This notebook is an organized entry point for regenerating:

- the Morris passport geographic distribution figure
- the PCA figure
- the combined `geo + PCA` two-panel figure
- the race-wise phenotype ridge and violin figures

It does not replace the underlying plotting scripts. It runs them in a documented order.

## Paths

```{r}
repo_root <- "/Users/subhashmahamkali/Documents/gwas_sap"

geo_pca_builder <- file.path(
  repo_root,
  "profiling/01_publication/figure_1/build_geo_pca_two_panel.R"
)

racewise_script <- file.path(
  repo_root,
  "profiling/01_publication/figure_2_racewise_phenotype/create_racewise_plots.R"
)

combined_geo_pca_pdf <- file.path(
  repo_root,
  "graphs/01_publication/figure_1/figure1_geo_pca_two_panel_A4_landscape.pdf"
)

racewise_root <- file.path(
  repo_root,
  "graphs/01_publication/figure_2_racewise_phenotype"
)

list(
  geo_pca_builder = geo_pca_builder,
  racewise_script = racewise_script,
  combined_geo_pca_pdf = combined_geo_pca_pdf,
  racewise_root = racewise_root
)
```

## 1. Build Geo + PCA Two-panel Figure

This step regenerates the combined two-panel figure:

- Panel A: Morris passport geographic distribution
- Panel B: PCA

```{r}
source(geo_pca_builder)
```

```{r}
file.exists(combined_geo_pca_pdf)
combined_geo_pca_pdf
```

## 2. Build Race-wise Ridge and Violin Figures

This step regenerates the race-wise phenotype figures, including:

- `ridge_hn_ln`
- `ridge_nr`
- `violin_hn_ln`
- `violin_nr`

```{r}
source(racewise_script)
```

## 3. Key Output Directories

```{r}
list.files(
  racewise_root,
  full.names = TRUE
)
```

## 4. Important Output Files

```{r}
important_outputs <- c(
  combined_geo_pca_pdf,
  file.path(racewise_root, "ridge_nr", "2020"),
  file.path(racewise_root, "ridge_nr", "2021"),
  file.path(racewise_root, "ridge_hn_ln", "2020"),
  file.path(racewise_root, "ridge_hn_ln", "2021")
)

important_outputs
```

## Notes

- The underlying source scripts remain the authoritative plotting code.
- This notebook is only an organized runner and record of what to execute.
- Running this notebook may also refresh individual panel outputs as a side effect of the sourced scripts.
