---
title: "genetic load in selective sweeps"
author: "subhash"
date: "2025-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('..//..//')) 
```

I overlap zero shot deleteriousness scores (ZSS) with the top 1% FSt positive selection peaks in wild, landrace, and improved sorghum. 
For each ZSS cutoff (top 1%, 5%, 10%), sites are classified as inside FSt peaks (“PEAK”) or outside (“BACKGROUND”)

```{r}

library(data.table)
library(ggplot2)

#Inputs
#ZSS_BED <- "/work/jyanglab/subhash/sorgsd/1.genetic_load/zeroscore.bed"
ZSS_BED <- "/mnt/nrdstor/jyanglab/subhash/git/GWAS_sorghum_seedling/largedata/zeroscore.bed"
#on HCC

# FST files for both comparisons (top 1% windows)
#FST_FILES <- list(
  #"wild vs landrace"   = "/work/jyanglab/subhash/sorgsd/2.pos/sorg_wild_vs_landrace.windowed.weir.fst",
  #"landrace vs improved" = "/work/jyanglab/subhash/sorgsd/2.pos/landrace_vs_improved.windowed.weir.fst")

FST_FILES <- list(
  "wild vs landrace"   = "data/2.2_positive_selec/sorg_wild_vs_landrace.windowed.weir.fst",
  "landrace vs improved" = "data/2.2_positive_selec/landrace_vs_improved.windowed.weir.fst"
)

cuts <- c("1%"=-7.016113, "5%"=-5.150879, "10%"=-3.637695)

OUTDIR <- "graphs/01_publication/3.genetic_load/"
dir.create(OUTDIR, recursive = TRUE, showWarnings = FALSE)

to_num_chr <- function(x){
  x <- sub("^chr|^Chr","",as.character(x))
  x <- sub("^0+","",x); x[nchar(x)==0] <- "0"; x}
p_to_stars <- function(p){
  if (is.na(p)) "ns" else if (p < 1e-4) "****" else if (p < 1e-3) "***"
  else if (p < 1e-2) "**" else if (p < 5e-2) "*" else "ns"}
merge_overlaps_onechr <- function(dt){
  if (nrow(dt) == 0L) return(dt)
  setorder(dt, start, end)
  out_s <- integer(); out_e <- integer()
  s <- dt$start[1]; e <- dt$end[1]
  for(i in seq_len(nrow(dt))[-1]){
    if (dt$start[i] <= e) e <- max(e, dt$end[i])
    else { out_s <- c(out_s, s); out_e <- c(out_e, e); s <- dt$start[i]; e <- dt$end[i] }
  }
  data.table(start = c(out_s, s), end = c(out_e, e))}

Z_all <- fread(ZSS_BED, header = FALSE)
setnames(Z_all, paste0("V", seq_len(ncol(Z_all))))
Z_all <- Z_all[, .(chr = to_num_chr(V1), pos = as.integer(V4), zss = as.numeric(V5))]
Z_all <- Z_all[is.finite(zss)]
setkey(Z_all, chr, pos)

pieces <- list()

for (comp in names(FST_FILES)){
  fst_path <- FST_FILES[[comp]]
  FST <- fread(fst_path)
  stopifnot(all(c("CHROM","BIN_START","BIN_END") %in% names(FST)))
  fst_col <- if ("WEIGHTED_FST" %in% names(FST)) "WEIGHTED_FST" else "WEIR_AND_COCKERHAM_FST"
  FST <- FST[, .(chr = to_num_chr(CHROM),
                 start = as.integer(BIN_START),
                 end   = as.integer(BIN_END),
                 fst   = as.numeric(get(fst_col)))]
  FST <- FST[is.finite(fst) & end > start]

  thr <- quantile(FST$fst, 0.99, na.rm = TRUE)
  peaks_raw <- FST[fst >= thr, .(chr, start, end)]
  # If 0-based: peaks_raw[, start := start + 1L]
  PEAKS <- peaks_raw[, merge_overlaps_onechr(.SD), by = chr]
  setkey(PEAKS, chr, start, end)

  Zrng <- Z_all[, .(chr, start=pos, end=pos, pos, zss)]
  setkey(Zrng, chr, start, end)
  ov <- foverlaps(Zrng, PEAKS, nomatch=0L)
  Z_tag <- copy(Z_all)
  Z_tag[, region := "BACKGROUND"]
  if (nrow(ov)) Z_tag[pos %in% ov$pos, region := "PEAK"]
  Z_tag[, region := factor(region, levels=c("BACKGROUND","PEAK"))]

  for (lab in names(cuts)){
    cutv <- cuts[[lab]]
    Z <- Z_tag[zss <= cutv]
    if (nrow(Z) < 2) next
    pval <- if (any(Z$region=="PEAK") && any(Z$region=="BACKGROUND"))
              wilcox.test(Z[region=="PEAK", zss],
                          Z[region=="BACKGROUND", zss],
                          alternative="less")$p.value else NA_real_
    Z[, `:=`(GROUP = comp, CUTOFF = lab, stars = p_to_stars(pval))]
    pieces[[paste(comp, lab, sep="_")]] <- Z
  }
}
df <- rbindlist(pieces, fill=TRUE)
stopifnot(nrow(df) > 0)

cuts_levels  <- c("1%","5%","10%")
group_levels <- names(FST_FILES)
id_levels <- as.vector(outer(cuts_levels, group_levels, paste, sep=" • "))

df[, id    := factor(paste(CUTOFF, GROUP, sep=" • "), levels=id_levels)]
df[, xnum  := as.numeric(id)]
df[, region := factor(region, levels=c("BACKGROUND","PEAK"))]

summ <- df[, .(y_star = quantile(zss, 0.995, na.rm=TRUE) + 0.15,
               stars  = unique(stars)), by=id]

block_df <- data.table(
  xmin = c(0.5, 3.5),
  xmax = c(3.5, 6.5),
  fill = c("#00000007", "#00000000"))

pop_labels <- data.table(
  x = c(2, 5),
  pop = group_levels)

y_bottom <- -12
y_top    <- max(df$zss, na.rm=TRUE)
reg_cols <- c(BACKGROUND = "#7f8c8d", PEAK = "#f7a600")


p <- ggplot(df, aes(x=id, y=zss)) +
  geom_rect(data=block_df, inherit.aes=FALSE,
            aes(xmin=xmin, xmax=xmax, ymin=-Inf, ymax=Inf),
            fill=block_df$fill, color=NA) +
  #geom_split_violin(aes(fill=region), scale="width", alpha=0.90) +
  geom_boxplot(data=df,
               aes(x=xnum + ifelse(region=="PEAK", +0.14, -0.14),
                   y=zss, group=interaction(id, region), fill=region),
               inherit.aes=FALSE, width=0.20, outlier.shape=NA,
               color="black", alpha=0.6, size=0.3) +
  stat_summary(data=df,
               aes(x=xnum + ifelse(region=="PEAK", +0.14, -0.14),
                   y=zss, color=region, group=interaction(id, region)),
               inherit.aes=FALSE, fun=median, geom="point",
               size=2, stroke=0.6) +
  geom_text(data=summ, inherit.aes=FALSE,
            aes(x=id, y=y_star, label=stars), size=5) +
  annotate("text", x=pop_labels$x, y=y_top + 0.45,
           label=pop_labels$pop, fontface=2, size=5) +
  scale_fill_manual(values=reg_cols) +
  scale_color_manual(values=reg_cols, guide="none") +
  scale_x_discrete(labels=rep(c("1%","5%","10%"), times=length(group_levels))) +
  coord_cartesian(ylim=c(y_bottom, y_top + 0.75), clip="off") +
  scale_y_continuous(expand=expansion(mult=c(0,0.03))) +
  labs(x=NULL, y="Zero-shot score (ZSS)") +
  theme_bw(base_size=12) +
  theme(
    legend.position="right",
    legend.title=element_blank(),
    panel.grid=element_blank(),
    axis.text.x=element_text(vjust=0.9),
    plot.margin=margin(6,18,4,6)
  )

ggsave(file.path(OUTDIR, "genetic_load_in_positive_selection_sorghum.tiff"),
       p, width=12, height=4, dpi=600, compression="lzw")

ggsave(
  file.path(OUTDIR, "genetic_load_in_positive_selection_sorghum.pdf"),
  p,
  width  = 32,
  height = 8,
  units  = "cm"
)
```


I overlap zero shot deleteriousness scores (ZSS) with the top 1% B2 balancing selection peaks in wild, landrace, and improved sorghum. 
For each ZSS cutoff (top 1%, 5%, 10%), sites are classified as inside B2 peaks (“PEAK”) or outside (“BACKGROUND”)



```{r}
library(data.table)
library(ggplot2)

#Inputs
#ZSS_BED <- "/work/jyanglab/subhash/sorgsd/1.genetic_load/zeroscore.bed"
ZSS_BED <- "/mnt/nrdstor/jyanglab/subhash/git/GWAS_sorghum_seedling/largedata/zeroscore.bed"

#B2_files <- list(
#wild     = "/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/wild.top1.ext5k.bed",
#landrace = "/mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.ext5k.bed",
#improved = "/work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.top1.ext5k.bed")

B2_files <- list(
wild     = "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/wild.top1.bed",
landrace = "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/land.top1.bed",
improved = "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/imp.top1.bed")

cuts   <- c("1%"=-7.016113, "5%"=-5.150879, "10%"=-3.637695)
OUTDIR <- "graphs/01_publication/3.genetic_load/"
dir.create(OUTDIR, showWarnings = FALSE, recursive = TRUE)

to_num_chr <- function(x){
  x <- sub("^chr|^Chr","",as.character(x))
  x <- sub("^0+","",x); x[nchar(x)==0] <- "0"; x
}
p_to_stars <- function(p){
  if (is.na(p)) "ns" else if (p < 1e-4) "****" else if (p < 1e-3) "***"
  else if (p < 1e-2) "**" else if (p < 5e-2) "*" else "ns"
}
Z_all <- fread(ZSS_BED, header = FALSE)[, .(
  chr = to_num_chr(V1), pos = as.integer(V4), zss = as.numeric(V5)
)]
setkey(Z_all, chr, pos)

ok <- vapply(B2_files, file.exists, logical(1))
if (!all(ok)) {
  message("Skipping missing files: ", paste(names(B2_files)[!ok], collapse=", "))
  B2_files <- B2_files[ok]
}
stopifnot(length(B2_files) > 0)

pieces <- list()
for (GROUP in names(B2_files)){
  B2 <- fread(B2_files[[GROUP]], header=FALSE)[, .(
    chr = to_num_chr(V1), start = as.integer(V2), end = as.integer(V3)
  )]
  setkey(B2, chr, start, end)
  for (lab in names(cuts)){
    cutv <- cuts[[lab]]
    Z <- Z_all[zss <= cutv]
    if (nrow(Z) < 2) next
    Zrng <- Z[, .(chr, start=pos, end=pos, pos, zss)]; setkey(Zrng, chr, start, end)
    ov <- foverlaps(Zrng, B2, nomatch=0L)
    
    Z[, region := "BACKGROUND"]
    if (nrow(ov)) Z[pos %in% ov$pos, region := "PEAK"]
    
    pval <- if (any(Z$region=="PEAK") && any(Z$region=="BACKGROUND"))
      wilcox.test(Z[region=="PEAK", zss], Z[region=="BACKGROUND", zss],
                  alternative="less")$p.value else NA_real_
    
    Z[, `:=`(GROUP = GROUP, CUTOFF = lab, stars = p_to_stars(pval))]
    pieces[[paste(GROUP, lab, sep="_")]] <- Z
    }}
df <- rbindlist(pieces, fill=TRUE)
stopifnot(nrow(df) > 0)
cuts_levels  <- c("1%","5%","10%")
group_levels <- c("wild","landrace","improved")
id_levels    <- as.vector(outer(cuts_levels, group_levels, paste, sep=" • "))

df[, id    := factor(paste(CUTOFF, GROUP, sep=" • "), levels = id_levels)]
df[, xnum  := as.numeric(id)]
df[, region := factor(region, levels = c("BACKGROUND","PEAK"))]

summ <- df[, .(
  y_star = quantile(zss, 0.995, na.rm=TRUE) + 0.15,
  stars  = unique(stars)
), by = id]

block_df <- data.table(
  xmin = c(0.5, 3.5, 6.5),
  xmax = c(3.5, 6.5, 9.5),
  fill = c("#00000007", "#00000000", "#00000007"))

pop_labels <- data.table(
  x = c(2, 5, 8),
  pop = c("wild", "landrace", "improved"))

reg_cols <- c(BACKGROUND = "#7f8c8d", PEAK = "#f7a600")
y_bottom <- -12                     
y_top    <- max(df$zss, na.rm=TRUE)

p <- ggplot(df, aes(x = id, y = zss)) +
    geom_rect(data = block_df, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            fill = block_df$fill, color = NA) +
 # geom_split_violin(aes(fill = region), scale = "width",
                    #color = "black", size = 0.3, alpha = 0.90) +
  
  geom_boxplot(
    data = df,
    aes(x = xnum + ifelse(region=="PEAK", +0.14, -0.14),
        y = zss, group = interaction(id, region), fill = region),
    inherit.aes = FALSE, width = 0.20, outlier.shape = NA,
    color = "black", alpha = 0.60, size = 0.3
  ) +
    stat_summary(
    data = df,
    aes(x = xnum + ifelse(region=="PEAK", +0.14, -0.14),
        y = zss, color = region, group = interaction(id, region)),
    inherit.aes = FALSE, fun = median, geom = "point",
    size = 2.1, stroke = 0.6
  ) +
  
  geom_text(data = summ, inherit.aes = FALSE,
            aes(x = id, y = y_star, label = stars),
            size = 5) +
  annotate("text", x = pop_labels$x, y = y_top + 0.45,
           label = pop_labels$pop, fontface = 2, size = 5) +
  scale_fill_manual(values = reg_cols) +
  scale_color_manual(values = reg_cols, guide = "none") +

  scale_x_discrete(labels = rep(c("1%","5%","10%"), times = length(group_levels))) +

  coord_cartesian(ylim = c(y_bottom, y_top + 0.75), clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03))) +
  labs(x = NULL, y = "Zero-shot score (ZSS)") +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(vjust = 0.9),
    plot.margin = margin(6, 18, 4, 6)
  )

ggsave(file.path(OUTDIR, "genetic_load_in_balancing_sweeps.tiff"),
       p, width = 16, height = 4, dpi = 600, compression = "lzw")

ggsave(
  file.path(OUTDIR, "genetic_load_in_balancing_sweeps.pdf"),
  p,                 # plot with 3 conditions
  width  = 48,       # cm
  height = 8,
  units  = "cm"
)
```

