---
title: "genetic load in selective sweeps"
author: "subhash"
date: "2025-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('..//..//')) 
```

I overlap zero shot deleteriousness scores (ZSS) with the top 1% FSt positive selection peaks in wild, landrace, and improved sorghum. 
For each ZSS cutoff (top 1%, 5%, 10%), sites are classified as inside FSt peaks (“PEAK”) or outside (“BACKGROUND”)

```{r}

library(data.table)
library(ggplot2)

#Inputs
#ZSS_BED <- "/work/jyanglab/subhash/sorgsd/1.genetic_load/zeroscore.bed"
ZSS_BED <- "/mnt/nrdstor/jyanglab/subhash/git/GWAS_sorghum_seedling/largedata/zeroscore.bed"
#on HCC

# FST files for both comparisons (top 1% windows)
#FST_FILES <- list(
  #"wild vs landrace"   = "/work/jyanglab/subhash/sorgsd/2.pos/sorg_wild_vs_landrace.windowed.weir.fst",
  #"landrace vs improved" = "/work/jyanglab/subhash/sorgsd/2.pos/landrace_vs_improved.windowed.weir.fst")

FST_FILES <- list(
  "wild vs landrace"   = "data/2.2_positive_selec/sorg_wild_vs_landrace.windowed.weir.fst",
  "landrace vs improved" = "data/2.2_positive_selec/landrace_vs_improved.windowed.weir.fst"
)

cuts <- c("1%"=-7.016113, "5%"=-5.150879, "10%"=-3.637695)

OUTDIR <- "graphs/01_publication/3.genetic_load/"
dir.create(OUTDIR, recursive = TRUE, showWarnings = FALSE)

to_num_chr <- function(x){
  x <- sub("^chr|^Chr","",as.character(x))
  x <- sub("^0+","",x); x[nchar(x)==0] <- "0"; x}
p_to_stars <- function(p){
  if (is.na(p)) "ns" else if (p < 1e-4) "****" else if (p < 1e-3) "***"
  else if (p < 1e-2) "**" else if (p < 5e-2) "*" else "ns"}
merge_overlaps_onechr <- function(dt){
  if (nrow(dt) == 0L) return(dt)
  setorder(dt, start, end)
  out_s <- integer(); out_e <- integer()
  s <- dt$start[1]; e <- dt$end[1]
  for(i in seq_len(nrow(dt))[-1]){
    if (dt$start[i] <= e) e <- max(e, dt$end[i])
    else { out_s <- c(out_s, s); out_e <- c(out_e, e); s <- dt$start[i]; e <- dt$end[i] }
  }
  data.table(start = c(out_s, s), end = c(out_e, e))}

Z_all <- fread(ZSS_BED, header = FALSE)
setnames(Z_all, paste0("V", seq_len(ncol(Z_all))))
Z_all <- Z_all[, .(chr = to_num_chr(V1), pos = as.integer(V4), zss = as.numeric(V5))]
Z_all <- Z_all[is.finite(zss)]
setkey(Z_all, chr, pos)

pieces <- list()

for (comp in names(FST_FILES)){
  fst_path <- FST_FILES[[comp]]
  FST <- fread(fst_path)
  stopifnot(all(c("CHROM","BIN_START","BIN_END") %in% names(FST)))
  fst_col <- if ("WEIGHTED_FST" %in% names(FST)) "WEIGHTED_FST" else "WEIR_AND_COCKERHAM_FST"
  FST <- FST[, .(chr = to_num_chr(CHROM),
                 start = as.integer(BIN_START),
                 end   = as.integer(BIN_END),
                 fst   = as.numeric(get(fst_col)))]
  FST <- FST[is.finite(fst) & end > start]

  thr <- quantile(FST$fst, 0.99, na.rm = TRUE)
  peaks_raw <- FST[fst >= thr, .(chr, start, end)]
  # If 0-based: peaks_raw[, start := start + 1L]
  PEAKS <- peaks_raw[, merge_overlaps_onechr(.SD), by = chr]
  setkey(PEAKS, chr, start, end)

  Zrng <- Z_all[, .(chr, start=pos, end=pos, pos, zss)]
  setkey(Zrng, chr, start, end)
  ov <- foverlaps(Zrng, PEAKS, nomatch=0L)
  Z_tag <- copy(Z_all)
  Z_tag[, region := "BACKGROUND"]
  if (nrow(ov)) Z_tag[pos %in% ov$pos, region := "PEAK"]
  Z_tag[, region := factor(region, levels=c("BACKGROUND","PEAK"))]

  for (lab in names(cuts)){
    cutv <- cuts[[lab]]
    Z <- Z_tag[zss <= cutv]
    if (nrow(Z) < 2) next
    pval <- if (any(Z$region=="PEAK") && any(Z$region=="BACKGROUND"))
              wilcox.test(Z[region=="PEAK", zss],
                          Z[region=="BACKGROUND", zss],
                          alternative="less")$p.value else NA_real_
    Z[, `:=`(GROUP = comp, CUTOFF = lab, stars = p_to_stars(pval))]
    pieces[[paste(comp, lab, sep="_")]] <- Z
  }
}
df <- rbindlist(pieces, fill=TRUE)
stopifnot(nrow(df) > 0)

cuts_levels  <- c("1%","5%","10%")
group_levels <- names(FST_FILES)
id_levels <- as.vector(outer(cuts_levels, group_levels, paste, sep=" • "))

df[, id    := factor(paste(CUTOFF, GROUP, sep=" • "), levels=id_levels)]
df[, xnum  := as.numeric(id)]
df[, region := factor(region, levels=c("BACKGROUND","PEAK"))]

summ <- df[, .(y_star = quantile(zss, 0.995, na.rm=TRUE) + 0.15,
               stars  = unique(stars)), by=id]

block_df <- data.table(
  xmin = c(0.5, 3.5),
  xmax = c(3.5, 6.5),
  fill = c("#00000007", "#00000000"))

pop_labels <- data.table(
  x = c(2, 5),
  pop = group_levels)

y_bottom <- -12
y_top    <- max(df$zss, na.rm=TRUE)
reg_cols <- c(BACKGROUND = "#7f8c8d", PEAK = "#f7a600")


p <- ggplot(df, aes(x=id, y=zss)) +
  geom_rect(data=block_df, inherit.aes=FALSE,
            aes(xmin=xmin, xmax=xmax, ymin=-Inf, ymax=Inf),
            fill=block_df$fill, color=NA) +
  #geom_split_violin(aes(fill=region), scale="width", alpha=0.90) +
  geom_boxplot(data=df,
               aes(x=xnum + ifelse(region=="PEAK", +0.14, -0.14),
                   y=zss, group=interaction(id, region), fill=region),
               inherit.aes=FALSE, width=0.20, outlier.shape=NA,
               color="black", alpha=0.6, size=0.3) +
  stat_summary(data=df,
               aes(x=xnum + ifelse(region=="PEAK", +0.14, -0.14),
                   y=zss, color=region, group=interaction(id, region)),
               inherit.aes=FALSE, fun=median, geom="point",
               size=2, stroke=0.6) +
  geom_text(data=summ, inherit.aes=FALSE,
            aes(x=id, y=y_star, label=stars), size=5) +
  annotate("text", x=pop_labels$x, y=y_top + 0.45,
           label=pop_labels$pop, fontface=2, size=5) +
  scale_fill_manual(values=reg_cols) +
  scale_color_manual(values=reg_cols, guide="none") +
  scale_x_discrete(labels=rep(c("1%","5%","10%"), times=length(group_levels))) +
  coord_cartesian(ylim=c(y_bottom, y_top + 0.75), clip="off") +
  scale_y_continuous(expand=expansion(mult=c(0,0.03))) +
  labs(x=NULL, y="Zero-shot score (ZSS)") +
  theme_bw(base_size=12) +
  theme(
    legend.position="right",
    legend.title=element_blank(),
    panel.grid=element_blank(),
    axis.text.x=element_text(vjust=0.9),
    plot.margin=margin(6,18,4,6)
  )

ggsave(file.path(OUTDIR, "genetic_load_in_positive_selection_sorghum.tiff"),
       p, width=12, height=4, dpi=600, compression="lzw")

ggsave(
  file.path(OUTDIR, "genetic_load_in_positive_selection_sorghum.pdf"),
  p,
  width  = 32,
  height = 8,
  units  = "cm"
)
```


I overlap zero shot deleteriousness scores (ZSS) with the top 1% B2 balancing selection peaks in wild, landrace, and improved sorghum. 
For each ZSS cutoff (top 1%, 5%, 10%), sites are classified as inside B2 peaks (“PEAK”) or outside (“BACKGROUND”)



```{r}
library(data.table)
library(ggplot2)

#Inputs
#ZSS_BED <- "/work/jyanglab/subhash/sorgsd/1.genetic_load/zeroscore.bed"
ZSS_BED <- "/mnt/nrdstor/jyanglab/subhash/git/GWAS_sorghum_seedling/largedata/zeroscore.bed"

#B2_files <- list(
#wild     = "/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/wild.top1.ext5k.bed",
#landrace = "/mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.ext5k.bed",
#improved = "/work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.top1.ext5k.bed")

B2_files <- list(
wild     = "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/wild.top1.bed",
landrace = "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/land.top1.bed",
improved = "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/imp.top1.bed")

cuts   <- c("1%"=-7.016113, "5%"=-5.150879, "10%"=-3.637695)
OUTDIR <- "graphs/01_publication/3.genetic_load/"
dir.create(OUTDIR, showWarnings = FALSE, recursive = TRUE)

to_num_chr <- function(x){
  x <- sub("^chr|^Chr","",as.character(x))
  x <- sub("^0+","",x); x[nchar(x)==0] <- "0"; x
}
p_to_stars <- function(p){
  if (is.na(p)) "ns" else if (p < 1e-4) "****" else if (p < 1e-3) "***"
  else if (p < 1e-2) "**" else if (p < 5e-2) "*" else "ns"
}
Z_all <- fread(ZSS_BED, header = FALSE)[, .(
  chr = to_num_chr(V1), pos = as.integer(V4), zss = as.numeric(V5)
)]
setkey(Z_all, chr, pos)

ok <- vapply(B2_files, file.exists, logical(1))
if (!all(ok)) {
  message("Skipping missing files: ", paste(names(B2_files)[!ok], collapse=", "))
  B2_files <- B2_files[ok]
}
stopifnot(length(B2_files) > 0)

pieces <- list()
for (GROUP in names(B2_files)){
  B2 <- fread(B2_files[[GROUP]], header=FALSE)[, .(
    chr = to_num_chr(V1), start = as.integer(V2), end = as.integer(V3)
  )]
  setkey(B2, chr, start, end)
  for (lab in names(cuts)){
    cutv <- cuts[[lab]]
    Z <- Z_all[zss <= cutv]
    if (nrow(Z) < 2) next
    Zrng <- Z[, .(chr, start=pos, end=pos, pos, zss)]; setkey(Zrng, chr, start, end)
    ov <- foverlaps(Zrng, B2, nomatch=0L)
    
    Z[, region := "BACKGROUND"]
    if (nrow(ov)) Z[pos %in% ov$pos, region := "PEAK"]
    
    pval <- if (any(Z$region=="PEAK") && any(Z$region=="BACKGROUND"))
      wilcox.test(Z[region=="PEAK", zss], Z[region=="BACKGROUND", zss],
                  alternative="less")$p.value else NA_real_
    
    Z[, `:=`(GROUP = GROUP, CUTOFF = lab, stars = p_to_stars(pval))]
    pieces[[paste(GROUP, lab, sep="_")]] <- Z
    }}
df <- rbindlist(pieces, fill=TRUE)
stopifnot(nrow(df) > 0)
cuts_levels  <- c("1%","5%","10%")
group_levels <- c("wild","landrace","improved")
id_levels    <- as.vector(outer(cuts_levels, group_levels, paste, sep=" • "))

df[, id    := factor(paste(CUTOFF, GROUP, sep=" • "), levels = id_levels)]
df[, xnum  := as.numeric(id)]
df[, region := factor(region, levels = c("BACKGROUND","PEAK"))]

summ <- df[, .(
  y_star = quantile(zss, 0.995, na.rm=TRUE) + 0.15,
  stars  = unique(stars)
), by = id]

block_df <- data.table(
  xmin = c(0.5, 3.5, 6.5),
  xmax = c(3.5, 6.5, 9.5),
  fill = c("#00000007", "#00000000", "#00000007"))

pop_labels <- data.table(
  x = c(2, 5, 8),
  pop = c("wild", "landrace", "improved"))

reg_cols <- c(BACKGROUND = "#7f8c8d", PEAK = "#f7a600")
y_bottom <- -12                     
y_top    <- max(df$zss, na.rm=TRUE)

p <- ggplot(df, aes(x = id, y = zss)) +
    geom_rect(data = block_df, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            fill = block_df$fill, color = NA) +
 # geom_split_violin(aes(fill = region), scale = "width",
                    #color = "black", size = 0.3, alpha = 0.90) +
  
  geom_boxplot(
    data = df,
    aes(x = xnum + ifelse(region=="PEAK", +0.14, -0.14),
        y = zss, group = interaction(id, region), fill = region),
    inherit.aes = FALSE, width = 0.20, outlier.shape = NA,
    color = "black", alpha = 0.60, size = 0.3
  ) +
    stat_summary(
    data = df,
    aes(x = xnum + ifelse(region=="PEAK", +0.14, -0.14),
        y = zss, color = region, group = interaction(id, region)),
    inherit.aes = FALSE, fun = median, geom = "point",
    size = 2.1, stroke = 0.6
  ) +
  
  geom_text(data = summ, inherit.aes = FALSE,
            aes(x = id, y = y_star, label = stars),
            size = 5) +
  annotate("text", x = pop_labels$x, y = y_top + 0.45,
           label = pop_labels$pop, fontface = 2, size = 5) +
  scale_fill_manual(values = reg_cols) +
  scale_color_manual(values = reg_cols, guide = "none") +

  scale_x_discrete(labels = rep(c("1%","5%","10%"), times = length(group_levels))) +

  coord_cartesian(ylim = c(y_bottom, y_top + 0.75), clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03))) +
  labs(x = NULL, y = "Zero-shot score (ZSS)") +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(vjust = 0.9),
    plot.margin = margin(6, 18, 4, 6)
  )

ggsave(file.path(OUTDIR, "genetic_load_in_balancing_sweeps.tiff"),
       p, width = 16, height = 4, dpi = 600, compression = "lzw")

ggsave(
  file.path(OUTDIR, "genetic_load_in_balancing_sweeps.pdf"),
  p,                 # plot with 3 conditions
  width  = 48,       # cm
  height = 8,
  units  = "cm"
)
```


#Violin plot for both in 5% of ZSS in balancing selection (b2) and positive selection(fst and XP-CLR)

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(vioplot)
})

# ==========================================================================
# Absolute deleterious score (|ZSS|) in Peak vs Background regions
# One-row plot: B2 (wild, landrace, improved) + Positive Selection (FST ∪ XPCLR)
# ZSS cutoff: top 5% (≤ -5.150879)
# Positive selection peaks = union of FST top 1% + XPCLR top 1% windows
# ==========================================================================

# --------------------------------------------------------------------------
# 0. Paths
# --------------------------------------------------------------------------
BASE <- "/mnt/nrdstor/jyanglab/subhash/git/GWAS_sorghum_seedling"

ZSS_BED <- file.path(BASE, "largedata/zeroscore.bed")

# B2 balancing selection peak BEDs (already top 1%)
B2_FILES <- list(
  wild     = file.path(BASE, "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/wild.top1.bed"),
  landrace = file.path(BASE, "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/land.top1.bed"),
  improved = file.path(BASE, "data/3.1.1_sorghum_genetic_load_in_selection_sweeps/imp.top1.bed")
)

# FST windowed files (full — we take top 1%)
FST_FILES <- list(
  "wild vs landrace"     = file.path(BASE, "data/2.2_positive_selec/sorg_wild_vs_landrace.windowed.weir.fst"),
  "landrace vs improved" = file.path(BASE, "data/2.2_positive_selec/landrace_vs_improved.windowed.weir.fst")
)

# XPCLR windowed files (full — we take top 1% by xpclr column)
XPCLR_FILES <- list(
  "wild vs landrace"     = file.path(BASE, "largedata/figure_2/domestication_landrace_vs_wild.all_chr.tsv"),
  "landrace vs improved" = file.path(BASE, "largedata/figure_2/breeding_improved_vs_landrace.all_chr.tsv")
)

OUTDIR <- file.path(BASE, "graphs/01_publication/3.genetic_load/")
dir.create(OUTDIR, recursive = TRUE, showWarnings = FALSE)

# --------------------------------------------------------------------------
# 1. Settings
# --------------------------------------------------------------------------
ZSS_CUTOFF     <- -5.150879   # top 5%
cutoff_to_plot <- "5%"

b2_groups  <- c("wild", "landrace", "improved")
pos_groups <- c("wild vs landrace", "landrace vs improved")
all_groups <- c(b2_groups, pos_groups)

col_bg   <- adjustcolor("#7f8c8d", alpha.f = 0.85)
col_peak <- adjustcolor("#f7a600", alpha.f = 0.85)
cols     <- c(col_bg, col_peak)

fmt_axis <- function(x) formatC(x, format = "fg", digits = 3)

# --------------------------------------------------------------------------
# 2. Helper functions
# --------------------------------------------------------------------------
to_num_chr <- function(x) {
  x <- sub("^chr|^Chr", "", as.character(x))
  x <- sub("^0+", "", x)
  x[nchar(x) == 0] <- "0"
  x
}

merge_overlaps <- function(dt) {
  # Merge overlapping/adjacent intervals per chromosome
  if (nrow(dt) == 0L) return(dt)
  setorder(dt, start, end)
  out_s <- integer(); out_e <- integer()
  s <- dt$start[1]; e <- dt$end[1]
  for (i in seq_len(nrow(dt))[-1]) {
    if (dt$start[i] <= e) {
      e <- max(e, dt$end[i])
    } else {
      out_s <- c(out_s, s); out_e <- c(out_e, e)
      s <- dt$start[i]; e <- dt$end[i]
    }
  }
  data.table(start = c(out_s, s), end = c(out_e, e))
}

draw_bracket <- function(x1, x2, y, p_val, tick, label_offset) {
  sig <- if      (p_val < 0.0001) "****"
         else if (p_val < 0.001)  "***"
         else if (p_val < 0.01)   "**"
         else if (p_val < 0.05)   "*"
         else                     "ns"
  segments(x1, y, x2, y, xpd = NA)
  segments(x1, y - tick, x1, y, xpd = NA)
  segments(x2, y - tick, x2, y, xpd = NA)
  text((x1 + x2) / 2, y + label_offset, labels = sig, cex = 1.0, xpd = NA)
}

# --------------------------------------------------------------------------
# 3. Load ZSS
# --------------------------------------------------------------------------
message("Loading ZSS BED ...")
Z_all <- fread(ZSS_BED, header = FALSE)
setnames(Z_all, paste0("V", seq_len(ncol(Z_all))))
Z_all <- Z_all[, .(chr = to_num_chr(V1), pos = as.integer(V4), zss = as.numeric(V5))]
Z_all <- Z_all[is.finite(zss)]
setkey(Z_all, chr, pos)

# Filter to top 5% deleterious
Z_del <- Z_all[zss <= ZSS_CUTOFF]
Z_del[, abs_zss := abs(zss)]
message(sprintf("  Total ZSS sites: %s | Top 5%% deleterious (≤ %.3f): %s",
                formatC(nrow(Z_all), big.mark = ","),
                ZSS_CUTOFF,
                formatC(nrow(Z_del), big.mark = ",")))

# --------------------------------------------------------------------------
# 4. B2 balancing selection: tag PEAK vs BACKGROUND
# --------------------------------------------------------------------------
message("Processing B2 peaks ...")

b2_results <- list()

for (grp in b2_groups) {
  B2 <- fread(B2_FILES[[grp]], header = FALSE)[, .(
    chr   = to_num_chr(V1),
    start = as.integer(V2),
    end   = as.integer(V3)
  )]
  setkey(B2, chr, start, end)

  # Overlap
  Zrng <- Z_del[, .(chr, start = pos, end = pos, pos, abs_zss)]
  setkey(Zrng, chr, start, end)
  ov <- foverlaps(Zrng, B2, nomatch = 0L)

  Z_tag <- copy(Z_del)
  Z_tag[, region := "BACKGROUND"]
  if (nrow(ov)) Z_tag[pos %in% ov$pos, region := "PEAK"]

  Z_tag[, GROUP  := grp]
  Z_tag[, CUTOFF := cutoff_to_plot]
  b2_results[[grp]] <- Z_tag

  message(sprintf("  B2 %s: %d peaks, %d PEAK sites, %d BG sites",
                  grp, nrow(B2),
                  sum(Z_tag$region == "PEAK"),
                  sum(Z_tag$region == "BACKGROUND")))
}

df_b2 <- rbindlist(b2_results)

# --------------------------------------------------------------------------
# 5. Positive selection: FST top 1% ∪ XPCLR top 1% → merged PEAK regions
# --------------------------------------------------------------------------
message("Processing positive selection (FST ∪ XPCLR) peaks ...")

pos_results <- list()

for (comp in pos_groups) {

  # --- FST top 1% windows ---
  fst_raw <- fread(FST_FILES[[comp]])
  fst_col <- if ("WEIGHTED_FST" %in% names(fst_raw)) "WEIGHTED_FST" else "WEIR_AND_COCKERHAM_FST"
  fst_dt <- fst_raw[, .(
    chr   = to_num_chr(CHROM),
    start = as.integer(BIN_START),
    end   = as.integer(BIN_END),
    val   = as.numeric(get(fst_col))
  )]
  fst_dt <- fst_dt[is.finite(val) & end > start]
  fst_thr <- quantile(fst_dt$val, 0.99, na.rm = TRUE)
  fst_peaks <- fst_dt[val >= fst_thr, .(chr, start, end)]
  message(sprintf("  %s — FST top 1%% threshold: %.4f → %d windows",
                  comp, fst_thr, nrow(fst_peaks)))

  # --- XPCLR top 1% windows ---
  xpclr_raw <- fread(XPCLR_FILES[[comp]])
  xpclr_dt <- xpclr_raw[, .(
    chr   = to_num_chr(chrom),
    start = as.integer(start),
    end   = as.integer(stop),
    val   = as.numeric(xpclr)
  )]
  xpclr_dt <- xpclr_dt[is.finite(val) & end > start]
  xpclr_thr <- quantile(xpclr_dt$val, 0.99, na.rm = TRUE)
  xpclr_peaks <- xpclr_dt[val >= xpclr_thr, .(chr, start, end)]
  message(sprintf("  %s — XPCLR top 1%% threshold: %.4f → %d windows",
                  comp, xpclr_thr, nrow(xpclr_peaks)))

  # --- Union of FST + XPCLR peaks, then merge overlapping intervals ---
  all_peaks <- rbind(fst_peaks, xpclr_peaks)
  merged_peaks <- all_peaks[, merge_overlaps(.SD), by = chr]
  setkey(merged_peaks, chr, start, end)
  message(sprintf("  %s — Merged (FST ∪ XPCLR) peak regions: %d",
                  comp, nrow(merged_peaks)))

  # --- Overlap with deleterious ZSS sites ---
  Zrng <- Z_del[, .(chr, start = pos, end = pos, pos, abs_zss)]
  setkey(Zrng, chr, start, end)
  ov <- foverlaps(Zrng, merged_peaks, nomatch = 0L)

  Z_tag <- copy(Z_del)
  Z_tag[, region := "BACKGROUND"]
  if (nrow(ov)) Z_tag[pos %in% ov$pos, region := "PEAK"]

  Z_tag[, GROUP  := comp]
  Z_tag[, CUTOFF := cutoff_to_plot]
  pos_results[[comp]] <- Z_tag

  message(sprintf("  %s — PEAK sites: %d, BG sites: %d",
                  comp,
                  sum(Z_tag$region == "PEAK"),
                  sum(Z_tag$region == "BACKGROUND")))
}

df_pos <- rbindlist(pos_results)

# --------------------------------------------------------------------------
# 6. Build plotting vectors (same structure as original one-row code)
# --------------------------------------------------------------------------
df_b2[,  region := factor(region, levels = c("BACKGROUND", "PEAK"))]
df_pos[, region := factor(region, levels = c("BACKGROUND", "PEAK"))]

get_pair <- function(dat, grp) {
  bg_v <- dat[GROUP == grp & region == "BACKGROUND", abs_zss]
  pk_v <- dat[GROUP == grp & region == "PEAK",       abs_zss]

  # Safety: vioplot needs ≥2 values
  if (length(bg_v) < 2) bg_v <- rep(median(bg_v, na.rm = TRUE), 2)
  if (length(pk_v) < 2) pk_v <- rep(median(pk_v, na.rm = TRUE), 2)

  list(bg = bg_v, peak = pk_v)
}

pair_list <- list()
for (grp in b2_groups)  pair_list[[grp]] <- get_pair(df_b2, grp)
for (grp in pos_groups) pair_list[[grp]] <- get_pair(df_pos, grp)

# Flatten for y-axis limits
all_vals <- unlist(lapply(pair_list, function(z) c(z$bg, z$peak)))
ylo   <- min(all_vals, na.rm = TRUE) * 0.92
stepy <- (max(all_vals, na.rm = TRUE) - min(all_vals, na.rm = TRUE)) * 0.10
yhi   <- max(all_vals, na.rm = TRUE) + stepy * 1.8
ticks <- pretty(c(ylo, yhi), n = 5)

# x positions: 5 pairs
# B2: 1-2, 4-5, 7-8
# Positive selection (FST ∪ XPCLR): 11-12, 14-15
x_pos <- c(1, 2, 4, 5, 7, 8, 11, 12, 14, 15)

plot_list <- list(
  pair_list[["wild"]]$bg,
  pair_list[["wild"]]$peak,
  pair_list[["landrace"]]$bg,
  pair_list[["landrace"]]$peak,
  pair_list[["improved"]]$bg,
  pair_list[["improved"]]$peak,
  pair_list[["wild vs landrace"]]$bg,
  pair_list[["wild vs landrace"]]$peak,
  pair_list[["landrace vs improved"]]$bg,
  pair_list[["landrace vs improved"]]$peak
)

col_vec <- rep(cols, 5)

# --------------------------------------------------------------------------
# 6b. Print median summary
# --------------------------------------------------------------------------
message("\n=== Median |ZSS| summary ===")
message(sprintf("%-25s %10s %10s %10s", "Group", "BG median", "PK median", "Diff (BG-PK)"))
message(paste(rep("-", 60), collapse = ""))

for (grp in all_groups) {
  med_bg <- median(pair_list[[grp]]$bg, na.rm = TRUE)
  med_pk <- median(pair_list[[grp]]$peak, na.rm = TRUE)
  message(sprintf("%-25s %10.4f %10.4f %10.4f", grp, med_bg, med_pk, med_bg - med_pk))
}

# --------------------------------------------------------------------------
# 7. Main plotting function
# --------------------------------------------------------------------------
make_one_row_plot <- function() {
  par(mar = c(6, 6, 4, 1), oma = c(0, 0, 0, 0))

  vioplot(plot_list[[1]], plot_list[[2]], plot_list[[3]], plot_list[[4]],
          plot_list[[5]], plot_list[[6]], plot_list[[7]], plot_list[[8]],
          plot_list[[9]], plot_list[[10]],
          at       = x_pos,
          col      = col_vec,
          border   = col_vec,
          rectCol  = NA,
          lineCol  = NA,
          colMed   = NA,
          ylim     = c(ylo, yhi),
          xaxt     = "n",
          yaxt     = "n",
          las      = 1,
          tck      = -0.03,
          ylab     = "Absolute deleterious score (|ZSS|)",
          main     = "Absolute deleterious score: Background vs Peak (Top 5% ZSS)",
          cex.lab  = 1.1,
          cex.axis = 0.9,
          cex.main = 1.2,
          font.main = 2)

  # Horizontal grid
  grid(nx = NA, ny = NULL, col = "gray70", lty = "solid", lwd = 0.6)

  # Re-draw violins on top of grid
  vioplot(plot_list[[1]], plot_list[[2]], plot_list[[3]], plot_list[[4]],
          plot_list[[5]], plot_list[[6]], plot_list[[7]], plot_list[[8]],
          plot_list[[9]], plot_list[[10]],
          at       = x_pos,
          col      = col_vec,
          border   = col_vec,
          rectCol  = NA,
          lineCol  = NA,
          colMed   = NA,
          add      = TRUE)

  # Boxplots inside violins
  for (i in seq_along(plot_list)) {
    boxplot(plot_list[[i]],
            at        = x_pos[i],
            add       = TRUE,
            boxwex    = 0.22,
            outline   = FALSE,
            col       = adjustcolor("white", alpha.f = 0.6),
            border    = "black",
            staplewex = 0.5,
            medcol    = "black",
            medlwd    = 2,
            whisklty  = 1,
            xaxt      = "n",
            yaxt      = "n")
  }

  # Y axis
  axis(2, at = ticks, labels = fmt_axis(ticks), las = 1, cex.axis = 0.9)

  # Group labels on x-axis
  axis(1,
       at     = c(1.5, 4.5, 7.5, 11.5, 14.5),
       labels = c("wild", "landrace", "improved",
                  "wild vs\nlandrace", "landrace vs\nimproved"),
       tick   = FALSE, line = 1, cex.axis = 0.85)

  # BG / PK sub-labels
  text(x = x_pos,
       y = par("usr")[3] - 0.04 * diff(par("usr")[3:4]),
       labels = rep(c("BG", "PK"), 5),
       xpd = NA, cex = 0.85)

  # Section headers
  text(x = mean(c(1.5, 7.5)),
       y = yhi + 0.04 * diff(c(ylo, yhi)),
       labels = "Balancing selection (B2)",
       xpd = NA, cex = 1.1, font = 2)

  text(x = mean(c(11.5, 14.5)),
       y = yhi + 0.04 * diff(c(ylo, yhi)),
       labels = expression(bold("Positive selection (FST" ~ union() ~ "XPCLR)")),
       xpd = NA, cex = 1.1)

  # Separator

  abline(v = 9.5, lty = 2, col = "gray50")

  # Significance brackets (Wilcoxon test for each pair)
  tick_h <- stepy * 0.12
  loff   <- stepy * 0.10

  for (k in 1:5) {
    i1 <- 2 * k - 1
    i2 <- 2 * k
    wt <- suppressWarnings(wilcox.test(plot_list[[i1]], plot_list[[i2]]))
    pair_max <- max(c(plot_list[[i1]], plot_list[[i2]]), na.rm = TRUE)
    bracket_y <- pair_max + stepy * 0.18
    draw_bracket(x_pos[i1], x_pos[i2], bracket_y, wt$p.value, tick_h, loff)
  }

  # Legend
  legend("topright",
         legend = c("Background", "Peak"),
         fill   = cols,
         border = cols,
         bty    = "n",
         cex    = 0.95)
}

# --------------------------------------------------------------------------
# 8. Save
# --------------------------------------------------------------------------
pdf(file.path(OUTDIR, "absZSS_5pct_one_row_B2_FST_XPCLR.pdf"),
    width = 16, height = 5.8)
make_one_row_plot()
dev.off()

png(file.path(OUTDIR, "absZSS_5pct_one_row_B2_FST_XPCLR.png"),
    width = 16, height = 5.8, units = "in", res = 350, bg = "white")
make_one_row_plot()
dev.off()

message("\n=== Done! Output saved to: ===")
message(file.path(OUTDIR, "absZSS_5pct_one_row_B2_FST_XPCLR.pdf"))
message(file.path(OUTDIR, "absZSS_5pct_one_row_B2_FST_XPCLR.png"))
```

