---
title: "genetic load"
author: "subhash"
date: "2025-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('/Users/subhashmahamkali/Documents/gwas_sap')) 
```


#remember more negative values menas meore deleterious load
#we have used Plant CAD DNA language model (zero shot score) to calulate deleteriousness

# plotting genetic load and number of deleterious in sorghum between wild, landrace and improved across different threholds with missing rate filters
thresholds: 5%, 10%, 25%
missing filters: 30%, 50%

```{r}
library(data.table)
library(ggplot2)

#outdir <- "/work/jyanglab/subhash/sorgsd/out_raw_full"
outdir <- "data/3.1_genetic_load_sorghum/"
miss_levels <- c(30, 50)  
for (m in miss_levels) {
  files <- list.files(outdir, pattern = sprintf("miss%02d_chr", m), full.names = TRUE)
  if (!length(files)) next
  df <- rbindlist(lapply(files, fread), use.names = TRUE, fill = TRUE)
  df[, cutoff := factor(cutoff, levels = c(0.05, 0.10, 0.25))]
  df <- df[!is.na(cutoff)]
  fsize <- 16

  ## ---------- Plots ----------
  pN <- ggplot(df, aes(x = sub, y = N_del, fill = cutoff)) +
    geom_boxplot(outlier.size = 0.5) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      x = "",
      y = "Number of deleterious alleles",
      title = sprintf("Missing ≤ %d%%", m)
    ) +
    theme(
      axis.text   = element_text(size = fsize),
      axis.title  = element_text(size = fsize, face = "bold"),
      legend.title = element_text(size = fsize, face = "bold"),
      legend.text  = element_text(size = fsize)
    )

  pW <- ggplot(df, aes(x = sub, y = Genetic_load, fill = cutoff)) +
    geom_boxplot(outlier.size = 0.5) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      x = "",
      y = "Genetic load",
      title = sprintf("Missing ≤ %d%%", m)
    ) +
    theme(
      axis.text   = element_text(size = fsize),
      axis.title  = element_text(size = fsize, face = "bold"),
      legend.title = element_text(size = fsize, face = "bold"),
      legend.text  = element_text(size = fsize)
    )

ggsave(sprintf("graphs/01_publication/3.genetic_load/Ndel_sorghum_miss%02d.png", m), plot   = pN, width  = 10, height = 7, dpi    = 200)
ggsave(sprintf("graphs/01_publication/3.genetic_load/GenLoad_sorghum_miss%02d.png", m),plot   = pW, width  = 10, height = 7, dpi    = 200)

#Wilcoxon tests
  cat("\n== Wilcoxon tests (missing <=", m, "%) ==\n")
  for (c in levels(df$cutoff)) {d <- df[cutoff == c]
    if (nrow(d) == 0) next
    cat("cutoff", c, "\n")
    pair_tests <- list(
      "wild vs landrace" = list(
        x = d[sub == "wild", Genetic_load],
        y = d[sub == "landrace", Genetic_load]),
      "wild vs improved" = list(
        x = d[sub == "wild", Genetic_load],
        y = d[sub == "improved", Genetic_load]),
      "landrace vs improved" = list(
        x = d[sub == "landrace", Genetic_load],
        y = d[sub == "improved", Genetic_load]
      ))
    for (nm in names(pair_tests)) {
      x <- pair_tests[[nm]]$x
      y <- pair_tests[[nm]]$y
      x <- x[!is.na(x)]
      y <- y[!is.na(y)]
      if (length(x) < 2 || length(y) < 2) {
        cat("  ", nm, ": skipped (not enough observations)\n")
      } else {cat("  ", nm, ":\n")
        print(wilcox.test(x, y))
      }}}}
```

#genetic load in top 1% cutoff only in sorghum between wild, landrace and improved

```{r}
library(data.table)
library(ggplot2)

outdir <- "data/3.1_genetic_load_sorghum/"

read_group <- function(outdir, group, miss = 50) {
  pats <- c(
    file.path(outdir, sprintf("raw_%s_miss%d_ALL.csv", group, miss)),
    file.path(outdir, sprintf("raw_%s_miss%d_*.csv", group, miss))   # per-chr
  )
  files <- unique(unlist(lapply(pats, Sys.glob)))
  if (length(files) == 0)
    stop(sprintf("No files found for group=%s miss=%d in %s", group, miss, outdir))
  rbindlist(lapply(files, fread), use.names = TRUE, fill = TRUE)
}

miss_level <- 50
df_impr <- read_group(outdir, "improved", miss_level)
df_wild <- read_group(outdir, "wild",     miss_level)
df_land <- read_group(outdir, "landrace", miss_level)

df <- rbindlist(list(df_impr, df_wild, df_land), use.names = TRUE, fill = TRUE)
## cutoff = 0.01 top 1%
df[, cutoff := as.numeric(as.character(cutoff))]
df001 <- df[cutoff == 0.01]
message("Rows per group at cutoff 0.01:")
print(df001[, .N, by = sub])

#plot for genetic load at 1% cutoff
p <- ggplot(df001, aes(x = sub, y = Genetic_load)) +
  geom_boxplot(
    width        = 0.7,
    outlier.size = 0.7,
    fill         = NA,
    color        = "black"
  ) +
  labs(
    title = sprintf("Genetic load at top 1%,missing ≤ %d%%", miss_level),
    x = "",
    y = "Genetic load"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

ggsave(sprintf("graphs/01_publication/3.genetic_load/GenLoad_sorghum_cut001_miss%02d.png", miss_level), plot   = p, width  = 10, height = 7, dpi    = 200)

```

