---
title: "genetic_load in sunflower"
author: "subhash"
date: "2025-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('..//..//')) 
```


#remember more negative values menas meore deleterious load
#we have used Plant CAD DNA language model (zero shot score) to calulate deleteriousness

# plotting genetic load and number of deleterious in sunflower between wild, landrace and moderen across different threholds with missing rate filters
thresholds: 5%, 10%, 25%
missing filters: 30%, 50%

```{r}
library(data.table)
library(ggplot2)

outdir      <- "data/3.2_genetic_load_sunflower/"
miss_level  <- 50
groups      <- c("wild","landrace","modern")      
cut_levels  <- c(0.25, 0.10, 0.05, 0.01)                  
cut_labels  <- c("25%","10%","5%","1%")
chr_filter  <- NULL                                      

read_group <- function(outdir, group, miss, chr = NULL) {
  pattern <- if (is.null(chr)) {
    file.path(outdir, sprintf("raw_%s_miss%02d_chr*.csv", group, miss))
  } else {
    file.path(outdir, sprintf("raw_%s_miss%02d_chr%s.csv", group, miss, chr))
  }
  files <- Sys.glob(pattern)
  if (!length(files)) return(data.table())
  DT <- rbindlist(lapply(files, function(f) {
    df <- tryCatch(fread(f), error = function(e) data.table())
    need <- c("sample","Genetic_load","N_del","cutoff","sub")
    miss_cols <- setdiff(need, names(df))
    if (length(miss_cols)) for (mc in miss_cols) df[[mc]] <- NA
    df[, cutoff := as.numeric(cutoff)]
    df
  }), use.names = TRUE, fill = TRUE)
  if (nrow(DT)) DT[, sub := factor(sub, levels = groups)]
  DT
}

dfs <- lapply(groups, read_group,
              outdir = outdir, miss = miss_level, chr = chr_filter)
df  <- rbindlist(dfs, use.names = TRUE, fill = TRUE)
stopifnot(nrow(df) > 0)
df <- df[!is.na(cutoff) & cutoff %in% cut_levels]
df[, cutoff := factor(cutoff, levels = cut_levels, labels = cut_labels)]
df[, sub    := factor(sub, levels = groups)]
keep_cuts2 <- c("5%","10%","25%")
df2 <- df[cutoff %in% keep_cuts2]
df2[, cutoff := factor(cutoff, levels = keep_cuts2)]   # x-axis order: 1, 5, 10

#N_del
df2[, Ndel_plot := N_del] 
print(df2[, .N, by = .(cutoff, sub)][order(cutoff, sub)])
df_box <- copy(df2)
df_box[, cutoff := factor(cutoff, levels = c("5%", "10%","25%"))]

fsize <- 16
# Number of deleterious alleles 
pN_sf <- ggplot(df_box, aes(x = sub, y = N_del, fill = cutoff)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x     = "",
    y     = "Number of deleterious alleles",
    title = sprintf("Sunflower – Missing \u2264 %d%%", miss_level)
  ) +
  theme_bw(base_size = fsize) +
  theme(
    axis.text    = element_text(size = fsize),
    axis.title   = element_text(size = fsize, face = "bold"),
    legend.title = element_text(size = fsize, face = "bold"),
    legend.text  = element_text(size = fsize)
  )

#Genetic load
pW_sf <- ggplot(df_box, aes(x = sub, y = Genetic_load, fill = cutoff)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x     = "",
    y     = "Genetic load",
    title = sprintf("Sunflower – Missing \u2264 %d%%", miss_level)
  ) +
  theme_bw(base_size = fsize) +
  theme(
    axis.text    = element_text(size = fsize),
    axis.title   = element_text(size = fsize, face = "bold"),
    legend.title = element_text(size = fsize, face = "bold"),
    legend.text  = element_text(size = fsize)
  )

ggsave(sprintf("graphs/01_publication/3.genetic_load/Ndel_sunflower_miss%02d.png", miss_level), plot   = pN_sf, width  = 10, height = 7, dpi    = 200)
ggsave(sprintf("graphs/01_publication/3.genetic_load/GenLoad_sunflower_miss%02d.png", miss_level), plot   = pW_sf, width  = 10, height = 7, dpi    = 200)
```


#genetic load in top 1% cutoff only in sunflower between wild, landrace and improved

```{r}
library(data.table)
library(ggplot2)

outdir <- "data/3.2_genetic_load_sunflower/"

read_group <- function(outdir, group, miss = 50) {
  pats  <- file.path(outdir, sprintf("raw_%s_miss%02d_chr*.csv", group, miss))
  files <- unique(unlist(lapply(pats, Sys.glob)))
  if (length(files) == 0)
    stop(sprintf("No files found for group=%s miss=%d in %s", group, miss, outdir))
  rbindlist(lapply(files, fread), use.names = TRUE, fill = TRUE)
}

miss_level <- 50
df_wild  <- read_group(outdir, "wild",     miss_level)
df_land  <- read_group(outdir, "landrace", miss_level)
df_mod   <- read_group(outdir, "modern",   miss_level)

df <- rbindlist(list(df_wild, df_land, df_mod), use.names = TRUE, fill = TRUE)

df[, sub := factor(
  sub,
  levels = c("wild", "landrace", "modern"),
  labels = c("wild", "landrace", "improved")
)]
## cutoff = 0.01 top 1%
df[, cutoff := as.numeric(as.character(cutoff))]
df001 <- df[cutoff == 0.01]

message("Rows per group at cutoff 0.01:")
print(df001[, .N, by = sub])

# plot for genetic load at 1% cutoff
p <- ggplot(df001, aes(x = sub, y = Genetic_load)) +
  geom_boxplot(
    width        = 0.7,
    outlier.size = 0.7,
    fill         = NA,
    color        = "black"
  ) +
  labs(
    title = sprintf("Sunflower – Genetic load at top 1%%, missing \u2264 %d%%", miss_level),
    x = "",
    y = "Genetic load"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")

ggsave(sprintf("graphs/01_publication/3.genetic_load/GenLoad_sunflower_cut001_miss%02d.png", miss_level), plot   = p, width  = 10, height = 7, dpi    = 200)

```

