---
title: "GWAS and selection integration"
author: "subhash"
date: "2025-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('..//..//'))   
```

# here we visualizes how GWAS ancestor allele effect sizes are distributed for SNPs inside high B2 balancing-selection sweeps, for each trait category and population (Improved, Landrace, Wild). 

 we have 4 different trait categories: Developmental, Architecture, Panicle, Seed under nitrogen resposne.
 
```{r}
#GWAS B2 integration
library(data.table)
library(ggplot2)

#BASE_DIR <- "/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/1.plink/1.stat_filter/1.het_filter/0.rmvp_full"
#OUTDIR   <- file.path(BASE_DIR, "4B_integration")

BASE_DIR <- "data/4.GWAS_selection/A.balancing_selection"
OUTDIR   <- BASE_DIR
cats <- c("Developmental", "Architecture", "Panicle", "Seed")
cols <- c(
  Wild     = "#1b9e77",  
  Landrace = "#d95f02",  
  Improved = "#7570b3"   
)

for (cat in cats) {
  prefix <- file.path(OUTDIR, paste0("NR_", cat, "_B2_top1"))
  curves_file <- paste0(prefix, "_pooled_curves.tsv")
  if (!file.exists(curves_file)) {
    warning("Missing file: ", curves_file)
    next
  }
  
  d <- fread(curves_file)
  d[, population := factor(population, levels = c("Improved","Landrace","Wild"))]
  
  p <- ggplot(d, aes(x = anc_effect, y = rank, color = population)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.6, alpha = 0.6) +
    scale_color_manual(values = cols) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    theme_bw(base_size = 13) +
    labs(
      title = paste0("B2_NR ", cat),
      x     = "Ancestorallele effect size",
      y     = "SNPs within B2 sweeps",
      color = "Population"
    )
 out_pdf <- file.path(
    "graphs/01_publication/4.GWAS_selection_integration",
    paste0("NR_", cat, "_B2_top1_plot.pdf")
  )
  ggsave(out_pdf, p, width = 4.3, height = 3.5, dpi = 300)
  message("Saved: ", out_pdf)
}
p
```


#how GWAS ancestor allele effect sizes are distributed for SNPs inside high positive selection sweeps, for each trait category.


```{r}
library(data.table)
library(ggplot2)

## Folder where those PosSel files live
BASE_DIR  <- "data/4.GWAS_selection/B.positive_selection"  
OUT_FIGS  <- "graphs/01_publication/4.GWAS_selection_integration"
dir.create(OUT_FIGS, recursive = TRUE, showWarnings = FALSE)

# Trait categories (must match the "NR_<cat>_PosSel_top1_..." naming)
cats <- c("Developmental", "Architecture", "Panicle", "Seed")

# Colors for the two populations in PosSel
cols <- c(
  Landrace = "#E69F00",
  Improved = "#009E73"
)

for (CATEGORY in cats) {

  # ---------- 1) Read pooled curve file ----------
  prefix      <- file.path(BASE_DIR, paste0("NR_", CATEGORY, "_PosSel_top1"))
  curves_file <- paste0(prefix, "_pooled_curves.tsv")

  if (!file.exists(curves_file)) {
    warning("Missing file: ", curves_file)
    next
  }

  curve <- fread(curves_file)

  # (optional) set population factor order
  curve[, population := factor(population, levels = c("Landrace", "Improved"))]

  # ---------- 2) Plot (same as in your generation script) ----------
  p <- ggplot(curve, aes(x = anc_effect, y = rank, color = population)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.6, alpha = 0.6) +
    scale_color_manual(values = cols) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    theme_bw(base_size = 13) +
    labs(
      title = paste0("Positive Selection (top1%) – NR ", CATEGORY),
      x     = "Ancestral-allele effect size (beta)",
      y     = "SNPs within selection sweeps",
      color = "Population"
    )

  # ---------- 3) Save figure ----------
  out_pdf <- file.path(
    OUT_FIGS,
    paste0("NR_", CATEGORY, "_PosSel_top1_plot.pdf")
  )

  ggsave(out_pdf, p, width = 4.3, height = 3.5, dpi = 300)
  message("Saved: ", out_pdf)
}

```


# violin plots showing the distribution of ancestral-allele effect sizes for SNPs within balancing selection sweeps (B2) and positive selection sweeps (PosSel), faceted by trait category.



```{r}
library(data.table)
library(ggplot2)

read_curves <- function(base_dir, sweep_tag, cats) {
  rbindlist(lapply(cats, function(cat) {
    f <- file.path(base_dir, paste0("NR_", cat, "_", sweep_tag, "_top1_pooled_curves.tsv"))
    if (!file.exists(f)) return(NULL)
    d <- fread(f)
    d[, category := cat]
    d[, sweep := sweep_tag]
    d
  }), fill = TRUE)
}

cats <- c("Developmental","Architecture","Panicle","Seed")

# -------- Balancing selection (B2) --------
b2_dir <- "data/4.GWAS_selection/A.balancing_selection"
b2 <- read_curves(b2_dir, "B2", cats)

# Example: magnitude view (tails)
b2[, abs_eff := anc_effect]
b2[, population := factor(population, levels = c("Wild","Landrace","Improved"))]

p_b2 <- ggplot(b2, aes(x = population, y = abs_eff, fill = population)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  facet_wrap(~category, scales = "free_y", nrow = 1) +
  theme_bw(base_size = 12) +
  labs(x = NULL, y = "|Ancestral-allele effect|", title = "Balancing selection (B2): effect magnitude in sweep SNPs")

ggsave("graphs/01_publication/4.GWAS_selection_integration/B2_violin_abs_effect.pdf",
       p_b2, width = 12, height = 3.5, dpi = 300)

# -------- Positive selection (PosSel) --------
ps_dir <- "data/4.GWAS_selection/B.positive_selection"
ps <- read_curves(ps_dir, "PosSel", cats)

ps[, abs_eff := anc_effect]
ps[, population := factor(population, levels = c("Wd_vs_Ld","Ld_vs_Imp","Landrace","Improved"))] 
# ^ pick the correct levels based on what "population" actually contains in your PosSel files

p_ps <- ggplot(ps, aes(x = population, y = abs_eff, fill = population)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  facet_wrap(~category, scales = "free_y", nrow = 1) +
  theme_bw(base_size = 12) +
  labs(x = NULL, y = "|Ancestral-allele effect|", title = "Positive selection: effect magnitude in sweep SNPs")

ggsave("graphs/01_publication/4.GWAS_selection_integration/PosSel_violin_abs_effect.pdf",
       p_ps, width = 12, height = 3.5, dpi = 300)


```

