---
title: "negative_selection"
author: "subhash"
date: "2025-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{bash,eval=F}
module purge
module load anaconda
mamba env create -f environment.yml -p $NRDSTOR/PlantCaduceus_1
conda activate $NRDSTOR/PlantCaduceus_1
#pip cache purge'
srun --nodes=1 --mem=16G --ntasks=1 --time=3:00:00 --pty bash
mamba install --no-update-deps cudatoolkit
pip install -r requirements.txt --no-build-isolation
python
import torch
from mamba_ssm import Mamba

srun --partition=gpu --gres=gpu:1 --time=00:30:00 --pty bash
python /work/jyanglab/subhash/PC/git/PlantCaduceus/src/zero_shot_score.py -input /work/jyanglab/subhash/PC/git/PlantCaduceus/examples/example_snp.tsv -output output.tsv -model 'kuleshov-group/PlantCaduceus_l32' -device 'cuda:1'

```


since UNL's HCC has limited GPU resource, I planed to do the above plant CAD analysis by chunks of .vcf file.
Insted of running plant CAD on all the 29 million variants I have subsetted them in to 10K vriants in each file.

```{bash,eval=F}
#!/usr/bin/bash
#SBATCH --job-name=chunk_VCF
#SBATCH --ntasks=1
#SBATCH --mem-per-cpu=10G
#SBATCH --time=10:00:00
#SBATCH --partition=jclarke,batch,guest
#SBATCH --output=/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.script/chunk_out.txt
#SBATCH --error=/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.script/chunk_err.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

module load bcftools
module load htslib

INPUT_VCF="/work/jyanglab/subhash/sorgsd/filterfed_vcf/SorGSD.289snp.miss05.vcf.gz"    # Input VCF
OUTDIR="/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/"         # Output directory
CHUNK_SIZE=10000                # SNPs per chunk
#mkdir -p $OUTDIR
# Step 1: Extract header
bcftools view -h $INPUT_VCF > $OUTDIR/header.vcf

# Step 2: Get total SNP count
TOTAL_SNPS=$(bcftools view -H $INPUT_VCF | wc -l)
CHUNKS=$(( (TOTAL_SNPS + CHUNK_SIZE - 1) / CHUNK_SIZE ))

# Step 3: Split into body chunks (no header)
bcftools view -H $INPUT_VCF | split -l $CHUNK_SIZE -d - $OUTDIR/body_chunk_

# Step 4: Combine header, rename chromosomes, keep GT only
# Rename chromosome 1-9 → Chr01–Chr09, 10 → Chr10
# Skip chromosomes outside 1–10
# Keep only GT field in sample columns
for f in $OUTDIR/body_chunk_*; do
  chunk_num=$(basename $f | cut -d'_' -f3)
  outvcf="$OUTDIR/chunk_${chunk_num}.vcf"

  cat $OUTDIR/header.vcf "$f" | \
  awk 'BEGIN{OFS="\t"}
    /^#/ {print; next}
    {
      
      if ($1 ~ /^[0-9]+$/ && $1 <= 10) {
        if ($1 < 10) $1 = "Chr0"$1;
        else $1 = "Chr"$1;
      } else {
        next 
      }
      for (i=10; i<=NF; i++) {
        split($i, a, ":"); $i = a[1]
      }
      print
    }' > "$outvcf"

  # Compress and index
  bgzip -f "$outvcf"
  tabix -p vcf "${outvcf}.gz"

  # Remove raw chunk
  rm "$f"
done
rm chunk_9919{36..74}.vcf.gz
rm chunk_9919{36..74}.vcf.gz.tbi

```


below code helps to create the input files from the vcf 10k variants chunks for the plant CAD model.

```{bash,eval=F}
#!/usr/bin/bash
#SBATCH --job-name=vcf_to_tsv
#SBATCH --ntasks=1
#SBATCH --mem=16G
#SBATCH --time=12:00:00
#SBATCH --partition=jclarke,batch,guest
#SBATCH --output=/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/script/vcf_to_tsv_out.txt
#SBATCH --error=/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/script/vcf_to_tsv_err.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

# Load required module
module load bedtools

# Set genome fasta path
genomeFA="/work/jyanglab/subhash/0.assignment_agro_932/0..refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.fa"
# Input directory with plain .vcf chunks
VCF_DIR="/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped"
# Output directory for .tsv files
OUT_DIR="${VCF_DIR}/tsv_out"

# Loop through each uncompressed VCF file
for inputVCF in $VCF_DIR/chunk_*.vcf; do
    base=$(basename $inputVCF .vcf)
    output="${OUT_DIR}/${base}.tsv"
    echo "Processing: $base"
    # Step 2: Create BED file
    grep -v '^#' ${inputVCF} | awk -v OFS="\t" '{print $1,$2-256,$2+256}' > ${inputVCF}.bed
    # Step 3: Extract sequence
    bedtools getfasta -tab -fi $genomeFA -bed ${inputVCF}.bed -fo ${inputVCF}.seq.tsv
    # Step 4: Combine SNP info with sequences
    awk -v OFS="\t" '$0 !~ /^#/ {print $1,$2-256,$2+256,$2,$4,$5}' ${inputVCF} | \
    paste - <(cut -f2 ${inputVCF}.seq.tsv) > ${output}_tmp
    # Step 5: Finalize
    echo -e "chr\tstart\tend\tpos\tref\talt\tsequences" > ${output}
    cat ${output}_tmp >> ${output}
    # Step 6: Cleanup
    rm ${inputVCF}.bed ${inputVCF}.seq.tsv ${output}_tmp
done
```

Running plant CAD 
I have run each gpu job on 300k varinatson each job

```{bash}
#!/bin/bash
#SBATCH --partition=gpu
#SBATCH --gres=gpu:1
#SBATCH --constraint=gpu_v100
#SBATCH --time=03:30:00
#SBATCH --mem=5G
#SBATCH --output=/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/0.zero_shot/script/batch_%A_%a.out
#SBATCH --error=/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/0.zero_shot/script/batch_%A_%a.err
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

# Activate environment
ml anaconda
conda activate $NRDSTOR/PlantCaduceus_1

# Use batch number from argument
BATCH_NUM=$1
BATCH_FILE="/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/0.zero_shot/script/0.file_names_batch/file_batch_${BATCH_NUM}"

echo "Starting Batch $BATCH_NUM using $BATCH_FILE"
while read FILE; do
    echo "Scoring $FILE"
    OUTFILE=$(basename $FILE .tsv)_score.tsv
    python /work/jyanglab/subhash/PC/git/PlantCaduceus/src/zero_shot_score.py \
        -input "$FILE" \
        -output "/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/0.zero_shot/$OUTFILE" \
        -model 'kuleshov-group/PlantCaduceus_l32' \
        -device 'cuda:0'
done < "$BATCH_FILE"

# Submit the next batch
NEXT_BATCH=$((BATCH_NUM + 1))
if [ -f "/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/0.zero_shot/script/0.file_names_batch/file_batch_${NEXT_BATCH}" ]; then
    echo "Submitting next batch: $NEXT_BATCH"
    sbatch --job-name=zero_shot_batch${NEXT_BATCH} batch_runner.slurm ${NEXT_BATCH}
else
    echo "All batches completed."
fi
```


```{bash,eval=FALSE}
#!/usr/bin/bash
#SBATCH --job-name=gene_load_summary_percent
#SBATCH --ntasks=1
#SBATCH --mem=160G
#SBATCH --time=25:00:00
#SBATCH --partition=jclarke,batch,guest
#SBATCH --output=/work/jyanglab/subhash/sorgsd/filterfed_vcf/glsout_percent.txt
#SBATCH --error=/work/jyanglab/subhash/sorgsd/filterfed_vcf/glserr_percent.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml R
#Rscript /work/jyanglab/subhash/sorgsd/filterfed_vcf/gene_load_summary.R
Rscript /work/jyanglab/subhash/sorgsd/filterfed_vcf/gl_percent.R
```



```{r,eval=FALSE}
library(data.table)
library(dplyr)

# Load data
individual_data <- fread("/work/jyanglab/subhash/sorgsd/filterfed_vcf/genotypes_fixed.txt", sep = "\t")
individual_data$V1 <- ifelse(individual_data$V1 %in% 1:9, paste0("Chr0", individual_data$V1), paste0("Chr", individual_data$V1))
setnames(individual_data, c("chr", "pos", "ref", "alt", paste0("indiv", 1:(ncol(individual_data) - 4))))

overlapping_data <- fread("/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/0.zero_shot/1.bed_files/zeroshot_in_gene.bed")
setnames(overlapping_data, c("chr", "start", "end", "pos", "zeroshotscore", "chr_1", "start_1", "end_1", "geneID", "annot", "string", "int"))

# Merge
merged_data <- merge(overlapping_data, individual_data, by = c("chr", "pos"))
d <- merged_data[, c("chr", "pos", "zeroshotscore", "start_1", "end_1", "geneID", paste0("indiv", 1:289))]

# Filter for deleterious SNPs
subset_df <- d[zeroshotscore < 0, ]

# --- VECTORIZE GENETIC LOAD ---

# Function to convert genotype to multiplier
geno_to_multiplier <- function(x) {
  ifelse(x == "./.", NA,
    ifelse(x == "1/1", 2,
      ifelse(x %in% c("0/1", "1/0"), 1,
        ifelse(x == "0/0", 0, NA)
      )
    )
  )
}

# Apply to all genotype columns at once
genotype_cols <- paste0("indiv", 1:289)
genotype_matrix <- subset_df[, ..genotype_cols]

# Convert all genotypes to multipliers
multiplier_matrix <- as.data.table(lapply(genotype_matrix, geno_to_multiplier))

# Multiply by zeroshotscore
for (i in seq_along(multiplier_matrix)) {
  multiplier_matrix[[i]] <- multiplier_matrix[[i]] * subset_df$zeroshotscore
}

# Combine gene info + load info
subset_df_final <- cbind(subset_df[, .(chr, pos, start_1, end_1, geneID)], multiplier_matrix)

# --- SUMMARIZE ---

# Create genetic load columns names
genetic_load_cols <- names(multiplier_matrix)

# Summarize per gene
gene_load_summary <- subset_df_final[, c(
  .(chr = first(chr), start = min(start_1), end = max(end_1)),
  lapply(.SD, function(x) sum(abs(x), na.rm = TRUE))
), by = geneID, .SDcols = genetic_load_cols]

# Save output
fwrite(gene_load_summary, "/work/jyanglab/subhash/sorgsd/filterfed_vcf/gene_load_summary.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```


```{r,eval=FALSE}
library(data.table)
library(dplyr)
# Load individual genotype data
individual_data <- fread("/work/jyanglab/subhash/sorgsd/filterfed_vcf/genotypes_fixed.txt", sep = "\t")
# Format chromosome names
individual_data$V1 <- ifelse(individual_data$V1 %in% 1:9, paste0("Chr0", individual_data$V1), paste0("Chr", individual_data$V1))
colnames(individual_data) <- c("chr", "pos", "ref", "alt", paste0("indiv", 1:(ncol(individual_data) - 4)))

# Load overlapping SNPs with gene and zero-shot scores
overlapping_data <- fread("/work/jyanglab/subhash/sorgsd/filterfed_vcf/0.vcf_chunk/0.unziped/tsv_out/0.zero_shot/1.bed_files/zeroshot_in_gene.bed")
colnames(overlapping_data) <- c("chr", "start", "end", "pos", "zeroshotscore", "chr_1", "start_1", "end_1", "geneID", "annot", "string", "int")

# Merge data
merged_data <- merge(overlapping_data, individual_data, by = c("chr", "pos"))

# Filter for deleterious SNPs only (zeroshotscore < 0)
d <- merged_data[, c("chr", "pos", "zeroshotscore", "start_1", "end_1", "geneID", paste0("indiv", 1:289))]
subset_df <- d[d$zeroshotscore < 0, ]
# Compute quantile cutoffs
quantiles <- quantile(subset_df$zeroshotscore, probs = c(0.5, 0.9, 0.99, 0.999), na.rm = TRUE)

# Subsets by percentile
top_50  <- subset_df[subset_df$zeroshotscore <= quantiles[1], ]
top_10  <- subset_df[subset_df$zeroshotscore <= quantiles[2], ]
top_1   <- subset_df[subset_df$zeroshotscore <= quantiles[3], ]
top_0.1 <- subset_df[subset_df$zeroshotscore <= quantiles[4], ]
# Define genetic load calculation
compute_genetic_load <- function(genotype, zeroshotscore) {
  if (genotype == "./.") {
    return(NA)
  } else if (genotype == "1/1") {
    return(2 * zeroshotscore)
  } else if (genotype == "0/0") {
    return(0)
  } else if (genotype %in% c("0/1", "1/0")) {
    return(zeroshotscore)
  } else {
    return(NA)
  }
}
# Generalized function for genetic load calculation and export
calculate_gene_load <- function(df, label) {
  for (i in 1:289) {
    indiv_col <- paste0("indiv", i)
    new_col <- paste0("genetic_load_", i)
    df[[new_col]] <- mapply(compute_genetic_load, df[[indiv_col]], df$zeroshotscore)
  }

  genetic_load_cols <- paste0("genetic_load_", 1:289)
  gene_load_summary <- df %>%
    group_by(geneID) %>%
    summarise(
      chr = first(chr),
      start = min(start_1),
      end = max(end_1),
      across(all_of(genetic_load_cols), ~sum(abs(.), na.rm = TRUE))
    )
  fwrite(
    gene_load_summary,
    paste0("/work/jyanglab/subhash/sorgsd/filterfed_vcf/gene_load_summary_top_", label, ".txt"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}
# Run for all subsets
calculate_gene_load(top_50, "50pct")
calculate_gene_load(top_10, "10pct")
calculate_gene_load(top_1, "1pct")
calculate_gene_load(top_0.1, "0.1pct")
```

