---
title: "race_wise_sap_deleterious"
author: "Subhash"
date: "2026-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This notebook documents the SAP-specific deleterious mutation workflow using PlantCaduceus zero-shot scores and the SAP race assignments used elsewhere in the project.

The workflow is:

1. prepare the PlantCaduceus environment on HCC
2. split the SAP VCF into 10K-variant chunks
3. convert chunk VCFs into PlantCaduceus TSV input
4. run zero-shot scoring in sequential GPU batches
5. merge all chunk score files
6. overlap scored variants with gene models
7. export a SAP genotype matrix from the VCF
8. calculate accession-level and gene-level deleterious load
9. merge deleterious load with SAP race labels and generate race-wise summaries

# Key inputs

```{r}
repo_root <- "/Users/subhashmahamkali/Documents/gwas_sap"
cluster_file <- file.path(repo_root, "data/0.SAP/sorted_cluster_data.txt")

sap_vcf <- "/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/2.filtered_vcf/sap_v5.vcf.gz"
plantcad_root <- "/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad"
reference_fa <- "/work/jyanglab/subhash/0.assignment_agro_932/0..refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.fa"

chunk_dir <- file.path(plantcad_root, "1.chunk_vcf")
tsv_dir <- file.path(plantcad_root, "2.tsv")
batch_dir <- file.path(plantcad_root, "3.batches")
score_dir <- file.path(plantcad_root, "4.scores")
merge_dir <- file.path(plantcad_root, "5.merged")
log_dir <- file.path(plantcad_root, "logs")
```

# PlantCaduceus environment setup

```{bash, eval=FALSE}
module purge
module load anaconda
mamba env create -f environment.yml -p $NRDSTOR/PlantCaduceus_1
conda activate $NRDSTOR/PlantCaduceus_1
mamba install --no-update-deps cudatoolkit
pip install -r requirements.txt --no-build-isolation

python
import torch
from mamba_ssm import Mamba

srun --partition=gpu --gres=gpu:1 --time=00:30:00 --pty bash
python /work/jyanglab/subhash/PC/git/PlantCaduceus/src/zero_shot_score.py \
  -input /work/jyanglab/subhash/PC/git/PlantCaduceus/examples/example_snp.tsv \
  -output output.tsv \
  -model 'kuleshov-group/PlantCaduceus_l32' \
  -device 'cuda:0'
```

# 1. Split the SAP VCF into 10K-variant chunks

The SAP VCF contains 4,703,244 variants, which gives 471 chunk files when chunked into 10,000 variants per file.

```{bash, eval=FALSE}
#!/usr/bin/bash
#SBATCH --job-name=sap_chunk_vcf
#SBATCH --ntasks=1
#SBATCH --mem=16G
#SBATCH --time=08:00:00
#SBATCH --partition=jclarke,batch,guest
#SBATCH --output=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/00_chunk_vcf.out
#SBATCH --error=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/00_chunk_vcf.err
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

set -euo pipefail

module load bcftools
module load htslib

INPUT_VCF="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/2.filtered_vcf/sap_v5.vcf.gz"
OUTDIR="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/1.chunk_vcf"
CHUNK_SIZE=10000

mkdir -p "$OUTDIR"
bcftools view -h "$INPUT_VCF" > "$OUTDIR/header.vcf"
bcftools view -H "$INPUT_VCF" | split -l $CHUNK_SIZE -d -a 5 - "$OUTDIR/body_chunk_"

for f in "$OUTDIR"/body_chunk_*; do
    chunk_num=$(basename "$f" | sed 's/body_chunk_//')
    outvcf="$OUTDIR/chunk_${chunk_num}.vcf"

    cat "$OUTDIR/header.vcf" "$f" | \
    awk 'BEGIN{OFS="\t"}
      /^#/ {print; next}
      {
        if ($1 ~ /^[0-9]+$/ && $1 <= 10) {
          if ($1 < 10) $1 = "Chr0"$1
          else $1 = "Chr"$1
        } else {
          next
        }
        print
      }' > "$outvcf"

    bgzip -f "$outvcf"
    tabix -f -p vcf "${outvcf}.gz"
done
```

# 2. Convert chunk VCFs into PlantCaduceus TSV input

The FASTA and `.fai` index were verified, and the chromosome naming matches the chunk VCFs (`Chr01` to `Chr10`).

```{bash, eval=FALSE}
#!/usr/bin/bash
#SBATCH --job-name=sap_vcf_to_tsv
#SBATCH --ntasks=1
#SBATCH --mem=24G
#SBATCH --time=12:00:00
#SBATCH --partition=jclarke,batch,guest
#SBATCH --output=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/01_vcf_to_tsv.out
#SBATCH --error=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/01_vcf_to_tsv.err
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

set -euo pipefail

module load bedtools
module load htslib

GENOME_FA="/work/jyanglab/subhash/0.assignment_agro_932/0..refrence/Phytozome/PhytozomeV13/Sbicolor/v5.1/assembly/Sbicolor_730_v5.0.fa"
VCF_DIR="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/1.chunk_vcf"
OUT_DIR="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/2.tsv"

mkdir -p "$OUT_DIR"

for inputVCF in "$VCF_DIR"/chunk_*.vcf.gz; do
    base=$(basename "$inputVCF" .vcf.gz)
    tmpvcf="$OUT_DIR/${base}.vcf"
    output="$OUT_DIR/${base}.tsv"

    zcat "$inputVCF" > "$tmpvcf"
    grep -v '^#' "$tmpvcf" | awk -v OFS="\t" '{print $1,$2-256,$2+256}' > "${tmpvcf}.bed"
    bedtools getfasta -tab -fi "$GENOME_FA" -bed "${tmpvcf}.bed" -fo "${tmpvcf}.seq.tsv"

    awk -v OFS="\t" '$0 !~ /^#/ {print $1,$2-256,$2+256,$2,$4,$5}' "$tmpvcf" | \
    paste - <(cut -f2 "${tmpvcf}.seq.tsv") > "${output}.tmp"

    echo -e "chr\tstart\tend\tpos\tref\talt\tsequences" > "$output"
    cat "${output}.tmp" >> "$output"

    rm -f "$tmpvcf" "${tmpvcf}.bed" "${tmpvcf}.seq.tsv" "${output}.tmp"
done
```

# 3. Build GPU batch lists

Thirty TSV chunk files per GPU job corresponds to roughly 300,000 variants per batch.

```{bash, eval=FALSE}
#!/usr/bin/bash
set -euo pipefail

TSV_DIR="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/2.tsv"
BATCH_DIR="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/3.batches"
BATCH_SIZE=30

mkdir -p "$BATCH_DIR"
find "$TSV_DIR" -name 'chunk_*.tsv' | sort > "$BATCH_DIR/all_tsv_files.txt"
split -l $BATCH_SIZE -d -a 3 "$BATCH_DIR/all_tsv_files.txt" "$BATCH_DIR/file_batch_"
```

Expected result:

- `471` TSV files total
- `16` batch files
- `15` full batches of 30 TSVs
- `1` final batch with 21 TSVs

# 4. Run PlantCaduceus zero-shot scoring on GPU

This script is sequential by design. Submitting batch `000` starts the first GPU job, and that job submits the next batch when it finishes.

```{bash, eval=FALSE}
#!/usr/bin/bash
#SBATCH --job-name=sap_plantcad
#SBATCH --partition=gpu
#SBATCH --gres=gpu:1
#SBATCH --constraint=gpu_v100
#SBATCH --time=04:00:00
#SBATCH --mem=8G
#SBATCH --output=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/03_plantcad_%A.out
#SBATCH --error=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/03_plantcad_%A.err
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

set -euo pipefail

ml anaconda
set +u
conda activate $NRDSTOR/PlantCaduceus_1
set -u

BATCH_NUM=$1
BATCH_FILE="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/3.batches/file_batch_${BATCH_NUM}"
OUT_DIR="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/4.scores"
MODEL="kuleshov-group/PlantCaduceus_l32"

mkdir -p "$OUT_DIR"

echo "Starting batch ${BATCH_NUM}"
while read -r FILE; do
    [ -z "$FILE" ] && continue
    base=$(basename "$FILE" .tsv)
    outfile="$OUT_DIR/${base}_score.tsv"

    if [ -s "$outfile" ]; then
        echo "Skipping existing $outfile"
        continue
    fi

    python /work/jyanglab/subhash/PC/git/PlantCaduceus/src/zero_shot_score.py \
        -input "$FILE" \
        -output "$outfile" \
        -model "$MODEL" \
        -device "cuda:0"
done < "$BATCH_FILE"

NEXT_BATCH=$(printf "%03d" $((10#$BATCH_NUM + 1)))
if [ -f "/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/3.batches/file_batch_${NEXT_BATCH}" ]; then
    sbatch /mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/0.scripts/03_run_plantcad_batch.slurm "$NEXT_BATCH"
else
    echo "All batches completed."
fi
```

Start the run with:

```{bash, eval=FALSE}
sbatch /mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/0.scripts/03_run_plantcad_batch.slurm 000
```

# 5. Merge all chunk score files

```{bash, eval=FALSE}
#!/usr/bin/bash
#SBATCH --job-name=sap_merge_scores
#SBATCH --ntasks=1
#SBATCH --mem=8G
#SBATCH --time=02:00:00
#SBATCH --partition=jclarke,batch,guest
#SBATCH --output=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/04_merge_scores.out
#SBATCH --error=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/04_merge_scores.err

set -euo pipefail

SCORE_DIR="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/4.scores"
OUT_DIR="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/5.merged"
OUT_FILE="$OUT_DIR/sap_v5_plantcad_scores_merged.tsv"

mkdir -p "$OUT_DIR"

first=1
for f in $(find "$SCORE_DIR" -name 'chunk_*_score.tsv' | sort); do
    if [ $first -eq 1 ]; then
        cat "$f" > "$OUT_FILE"
        first=0
    else
        tail -n +2 "$f" >> "$OUT_FILE"
    fi
done
```

# 6. Export the SAP genotype matrix from the VCF

The deleterious-load calculation needs genotype calls aligned to the scored variant positions.

```{bash, eval=FALSE}
#!/usr/bin/bash
#SBATCH --job-name=sap_export_gt
#SBATCH --ntasks=1
#SBATCH --mem=24G
#SBATCH --time=08:00:00
#SBATCH --partition=jclarke,batch,guest
#SBATCH --output=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/05_export_gt.out
#SBATCH --error=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/05_export_gt.err

set -euo pipefail

module load bcftools

INPUT_VCF="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/2.filtered_vcf/sap_v5.vcf.gz"
OUT_FILE="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/5.merged/sap_v5_genotypes_fixed.txt"

bcftools query -f '%CHROM\t%POS\t%REF\t%ALT[\t%GT]\n' "$INPUT_VCF" > "$OUT_FILE"
```

# 7. Overlap scored variants with gene models

This step was implicit in the old workflow because the downstream load script reads a precomputed `zeroshot_in_gene.bed`. The same logic should be applied for SAP.

Assumptions:

- `sap_v5_plantcad_scores_merged.tsv` contains at least `chr`, `start`, `end`, `pos`, and `zeroShotScore`
- a sorghum gene annotation BED file is available

```{bash, eval=FALSE}
#!/usr/bin/bash
#SBATCH --job-name=sap_score_gene_overlap
#SBATCH --ntasks=1
#SBATCH --mem=16G
#SBATCH --time=04:00:00
#SBATCH --partition=jclarke,batch,guest
#SBATCH --output=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/06_gene_overlap.out
#SBATCH --error=/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/logs/06_gene_overlap.err

set -euo pipefail

module load bedtools

SCORE_TSV="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/5.merged/sap_v5_plantcad_scores_merged.tsv"
GENE_BED="/work/jyanglab/subhash/annotations/Sbicolor_v5.1_genes.bed"
OUT_BED="/mnt/nrdstor/jyanglab/subhash/datasets/1.SAP_SNPs/3.plantcad/5.merged/sap_v5_zeroshot_in_gene.bed"

bedtools intersect -wa -wb -a "$SCORE_TSV" -b "$GENE_BED" > "$OUT_BED"
```

Update `GENE_BED` above to the actual annotation BED used on HCC.

# 8. Accession-level and gene-level deleterious load

This section adapts the older SorGSD gene-load code to SAP.

```{r}
library(data.table)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)

sap_gt_file <- file.path(merge_dir, "sap_v5_genotypes_fixed.txt")
sap_overlap_file <- file.path(merge_dir, "sap_v5_zeroshot_in_gene.bed")

# Update these column names if the overlap bed uses a different structure.
overlap_colnames <- c(
  "chr", "start", "end", "pos", "ref", "alt", "sequences", "zeroShotScore",
  "gene_chr", "gene_start", "gene_end", "geneID", "annot"
)
```

```{r, eval=FALSE}
individual_data <- fread(sap_gt_file, sep = "\t", header = FALSE)
setnames(
  individual_data,
  c("chr", "pos", "ref", "alt", paste0("indiv", seq_len(ncol(individual_data) - 4)))
)

overlapping_data <- fread(sap_overlap_file, sep = "\t", header = FALSE)
setnames(overlapping_data, overlap_colnames[seq_len(ncol(overlapping_data))])

merged_data <- merge(overlapping_data, individual_data, by = c("chr", "pos", "ref", "alt"))

d <- merged_data[, c("chr", "pos", "zeroShotScore", "gene_start", "gene_end", "geneID", paste0("indiv", 1:400)), with = FALSE]
subset_df <- d[zeroShotScore < 0, ]
```

```{r}
geno_to_multiplier <- function(x) {
  ifelse(x == "./.", NA,
    ifelse(x == "1/1", 2,
      ifelse(x %in% c("0/1", "1/0"), 1,
        ifelse(x == "0/0", 0, NA)
      )
    )
  )
}
```

```{r, eval=FALSE}
genotype_cols <- grep("^indiv", names(subset_df), value = TRUE)
genotype_matrix <- subset_df[, ..genotype_cols]
multiplier_matrix <- as.data.table(lapply(genotype_matrix, geno_to_multiplier))

for (i in seq_along(multiplier_matrix)) {
  multiplier_matrix[[i]] <- multiplier_matrix[[i]] * subset_df$zeroShotScore
}

subset_df_final <- cbind(
  subset_df[, .(chr, pos, gene_start, gene_end, geneID)],
  multiplier_matrix
)

# Accession-level deleterious load
accession_load <- data.table(
  genotype = genotype_cols,
  deleterious_load = sapply(multiplier_matrix, function(x) sum(abs(x), na.rm = TRUE)),
  deleterious_sites = sapply(multiplier_matrix, function(x) sum(!is.na(x) & x != 0)),
  mean_load_per_site = sapply(multiplier_matrix, function(x) mean(abs(x[x != 0]), na.rm = TRUE))
)

# Gene-level deleterious load
raw_gene_load_cols <- names(multiplier_matrix)
gene_load_summary <- subset_df_final[, c(
  .(chr = first(chr), start = min(gene_start), end = max(gene_end)),
  lapply(.SD, function(x) sum(abs(x), na.rm = TRUE))
), by = geneID, .SDcols = raw_gene_load_cols]

fwrite(accession_load, file.path(merge_dir, "sap_accession_deleterious_load.txt"), sep = "\t")
fwrite(gene_load_summary, file.path(merge_dir, "sap_gene_load_summary.txt"), sep = "\t")
```

# 9. Merge accession-level load with SAP race labels

```{r}
cluster_names <- c(
  "1" = "durra",
  "2" = "kafir",
  "3" = "caudatum",
  "4" = "milo/durra-bicolor",
  "5" = "mixed/bicolor",
  "6" = "guinea"
)
```

```{r, eval=FALSE}
cluster <- read.table(cluster_file, header = TRUE, stringsAsFactors = FALSE) %>%
  mutate(
    Cluster = as.character(Cluster),
    ClusterName = cluster_names[Cluster],
    genotype = str_replace_all(PI_numbers, "\\s+", "")
  ) %>%
  select(genotype, Cluster, ClusterName)

accession_load <- fread(file.path(merge_dir, "sap_accession_deleterious_load.txt")) %>%
  mutate(genotype = str_remove(genotype, "^indiv"))

# If accession_load already contains PI IDs instead of indiv numbers, drop the line above.

race_load <- inner_join(accession_load, cluster, by = "genotype")
```

The key point above is ID alignment. If the accession load table is indexed as `indiv1..indiv400`, map those columns back to SAP genotype IDs first. If the exported genotype matrix already preserves PI IDs in the header, the join can be done directly.

# 10. Race-wise summaries and plots

```{r}
race_colors <- c(
  "durra" = "#FF7F00",
  "kafir" = "#E69F00",
  "caudatum" = "#984EA3",
  "milo/durra-bicolor" = "#F781BF",
  "mixed/bicolor" = "#A6A6A6",
  "guinea" = "#56B4E9"
)
```

```{r, eval=FALSE}
race_summary <- race_load %>%
  group_by(ClusterName) %>%
  summarise(
    n = n(),
    mean_load = mean(deleterious_load, na.rm = TRUE),
    median_load = median(deleterious_load, na.rm = TRUE),
    sd_load = sd(deleterious_load, na.rm = TRUE),
    mean_sites = mean(deleterious_sites, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(median_load))

race_summary
```

```{r, eval=FALSE}
ggplot(
  race_load %>% mutate(ClusterName = reorder(ClusterName, deleterious_load, FUN = median, na.rm = TRUE)),
  aes(x = ClusterName, y = deleterious_load, fill = ClusterName)
) +
  geom_violin(alpha = 0.8, trim = FALSE, color = "grey30") +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  scale_fill_manual(values = race_colors) +
  labs(x = NULL, y = "Deleterious load", title = "Race-wise SAP deleterious load") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1))
```

```{r, eval=FALSE}
ggplot(
  race_load %>% mutate(ClusterName = reorder(ClusterName, deleterious_sites, FUN = median, na.rm = TRUE)),
  aes(x = ClusterName, y = deleterious_sites, fill = ClusterName)
) +
  geom_violin(alpha = 0.8, trim = FALSE, color = "grey30") +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  scale_fill_manual(values = race_colors) +
  labs(x = NULL, y = "Number of deleterious sites", title = "Race-wise deleterious site counts") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1))
```

# Notes

- Jensina confirmed that the phenotype BLUE models intentionally used row and column without separate block or sub-block terms because block structure was confounded with plot coordinates. That is separate from this deleterious-load workflow.
- The HCC GPU batching here is sequential, not parallel.
- The overlap step requires the actual sorghum gene BED path used on HCC.
- The accession-ID mapping step must be checked carefully before any publication figure is made from the race-wise deleterious-load table.
