---
title: "Phenotype distribution"
author: "Jinliang Yang"
date: "07-23-2024"
output: NULL
DATA : "sorghum_seedling"
Phenotypic Traits : "plant seedling height, leaf count, shoot dry weight, shoot fresh weight"
---

### path normalization

```{r setup}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('..//..//')) 
```

### setting input files

Hi Lucas,
 
I am attaching the name conversion I did as we discussed in the morning. In the file SAP.crazynames2.csv, the first column is the original name from the vcf file downloaded from the EVA, and the second column is the new name I assign. The file SAP.matchdiy.xlsx is the is the spreadsheet being screenshot in the slides.
 
Thanks,
Hongyu

- the first column is the original name from the vcf file downloaded from the EVA
- the second column is the new name I assign

```{r}
l <- read.table("data/SAP.crazynames2.csv")
l$diff <- 0
l[l$V1 != l$V2, ]$diff <- 1
```



```{r}
library(data.table)
d1=fread("output/blup_HN_dry.txt",head=T,data.table=F)
names(d1)[2] <- "value"
d1$trait <- "Dry Weight"
d1$treatment <- "high N"

d2=fread("output/blup_HN_fresh.txt",head=T,data.table=F)
names(d2)[2] <- "value"
d2$trait <- "Fresh Weight"
d2$treatment <- "high N"

d3=fread("output/blup_HN_leaf.txt",head=T,data.table=F)
names(d3)[2] <- "value"
d3$trait <- "Leaf Number"
d3$treatment <- "high N"

d4=fread("output/blup_HN_PH.txt",head=T,data.table=F)
names(d4)[2] <- "value"
d4$trait <- "Seedling Height"
d4$treatment <- "high N"


d5=fread("output/blup_LN_dry.txt",head=T,data.table=F)
names(d5)[2] <- "value"
d5$trait <- "Dry Weight"
d5$treatment <- "low N"

d6=fread("output/blup_LN_fresh.txt",head=T,data.table=F)
names(d6)[2] <- "value"
d6$trait <- "Fresh Weight"
d6$treatment <- "low N"

d7=fread("output/blup_LN_leaf.txt",head=T,data.table=F)
names(d7)[2] <- "value"
d7$trait <- "Leaf Number"
d7$treatment <- "low N"

d8=fread("output/blup_LN_PH.txt",head=T,data.table=F)
names(d8)[2] <- "value"
d8$trait <- "Seedling Height"
d8$treatment <- "low N"

d <- rbind(d1, d2, d3, d4, d5, d6, d7, d8)
```

### creating the histogram and density plot and saving it as png


```{r}
library(ggplot2)
library(tidyr)
library(plyr)

dt <- gather(p, key=trait, value=value, 3:10)
dt <- dt[!is.na(dt$value),]


mu <- ddply(d, .(trait, treatment), summarise,
            grp.mean=mean(value, na.rm=T))

# Change line colors by groups
p1 <- ggplot(d, aes(x=value, color=trait, fill=treatment)) +
  facet_wrap(vars(trait), scales = "free") +
  #geom_histogram(aes(y=..density..), position="identity", alpha=0.3, bins=30)+
  geom_density(alpha=0.3)+
  geom_vline(data=mu, aes(xintercept=grp.mean, color=trait, fill=treatment),
             linetype="dashed")+
  #scale_color_manual(values=c("#E69F00", "#56B4E9", "#fe6f5e"))+
  #scale_fill_manual(values=c("#E69F00", "#56B4E9", "#fe6f5e"), labels=c("HE", "HO", "Nip"))+
  labs(title="", x="Trait value", y = "Density")+
  guides(color="none", fill=guide_legend(title="")) +
  theme_classic() +
  #scale_y_continuous(limits = c(-0.01, 1)) +
    theme(plot.title = element_text(size=20, face = "bold"), 
          axis.text=element_text(size=16, face="bold"),
          strip.text.y = element_text(size = 16, face = "bold"),
          axis.title=element_text(size=18, face="bold"),
          strip.text.x = element_text(size = 16, face="bold"),
          #axis.text.x = element_text(angle = 15, hjust=0.8),
          legend.position = c(0.85, 0.15), 
          legend.title = element_text(size=18, face="bold"),
          legend.text = element_text(size=18))

p1

pdf("graphs/pheno_stat.pdf", width=14, height=8)
p1
dev.off()

```

