---
title: "Comparison between RNA-seq and genomics SNP sets"
author: "Jinliang"
date: "2024-07-09"
output: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/work/jyanglab/subhash/variant_calling")
```


/work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode.vcf.gz

# 1. convert vcf to plink2 bed format 

```{bash}
ml bcftools
ml vcftools
ml plink2

# RNA-seq SNPs
plink2 -vcf /work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode.vcf.gz --set-all-var-ids @_#_\$1_\$2 --make-bed --out /work/jyanglab/jyang21/sbhmp/SbDiv_imputed_Filter2.recode

# SAP SNPs
plink2 -vcf /work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.gz --set-all-var-ids @_#_\$1_\$2 --make-bed --out /work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924

# --vcf: 35935778 variants scanned.
# Warning: 3,969,709 multiallelic variants skipped (not yet supported).
```

713 samples 
170608 variants

### 2. Get SNP stat

```{bash}
plink2 -bfile /work/jyanglab/jyang21/sbhmp/SbDiv_imputed_Filter2.recode --freq --missing --pca --out /work/jyanglab/jyang21/sbhmp/SbDiv_imputed_Filter2  

# --ibc (ported from GCTA) calculates three inbreeding coefficients for each sample

plink2 -bfile /work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924 --freq --missing --pca --out /work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924 
```

--freq: Allele frequencies (founders only) written to
/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.afreq .
--missing: Sample missing data report written to
/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.smiss .
--missing: Variant missing data report written to
/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.vmiss .



# RNA-seq SNPs

```{r}
library(data.table)

df1 <- fread("/work/jyanglab/jyang21/sbhmp/SbDiv_imputed_Filter2.afreq", data.table=FALSE, header=T)
# 170608
range(df1$ALT_FREQS)
# 0.0504909 0.9495090; 

df2 <- fread("/work/jyanglab/jyang21/sbhmp/SbDiv_imputed_Filter2.vmiss", data.table=FALSE, header=T)
mean(df2$F_MISS)
# 0: imputed no missing data

df3 <- fread("/work/jyanglab/jyang21/sbhmp/SbDiv_imputed_Filter2.smiss", data.table=FALSE, header=T)
# 713 samples!


pdf("graphs/RNA-seq_frq_tem.pdf", width=6, height=5)
hist(df1$ALT_FREQS, main="Alt Allele Freq", xlab="Allele Freq", breaks = 100, col="lightblue")
abline(v=0.05, col="red", lty=2, lwd=3)
abline(v=0.95, col="red", lty=2, lwd=3)
hist(df3$F_MISS, main="Count by locus", xlab="Variant missingness", breaks = 100, col="lightgreen")
abline(v=0.1, col="red", lty=2, lwd=3)
dev.off()

```

# Genomics SNPs

```{r}
library(data.table)

df1 <- fread("/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.afreq", data.table=FALSE, header=T)
# 35935778
range(df1$ALT_FREQS)
# 0.00125 1.00000 
nrow(subset(df1, ALT_FREQS > 0.05 & ALT_FREQS < 0.95))
# 7889507


df2 <- fread("/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.smiss", data.table=FALSE, header=T)
mean(df2$F_MISS)
# 7.735799e-05: very low individual missing rate
# 400 samples

df3 <- fread("/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.vmiss", data.table=FALSE, header=T)
# 713 samples!

nrow(subset(df3, F_MISS < 0.1))


pdf("graphs/SAP_raw_frq.pdf", width=12, height=4)
par(mfrow = c(1, 3))  ## set the layout
hist(df1$ALT_FREQS, main="Alt Allele Freq", xlab="Allele Freq", breaks = 100, col="lightblue")
abline(v=0.05, col="red", lty=2, lwd=3)
abline(v=0.95, col="red", lty=2, lwd=3)
hist(df2$F_MISS, main="Count by sample", xlab="Sample missingness", breaks = 100, col="purple")
hist(df3$F_MISS, main="Count by locus", xlab="Variant missingness", breaks = 100, col="lightgreen")
abline(v=0.1, col="red", lty=2, lwd=3)
dev.off()
```


# Merge the two datasets together

filtered genomics variants:
/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/

```{bash}
plink2 --bfile RAW_SAP_BQSR.vcf_v5_060924 --maf 0.05 --geno 0.1 --make-bed --out /work/jyanglab/jyang21/sbhmp/filtered_RAW_SAP_BQSR.vcf_v5_060924

plink2 -bfile /work/jyanglab/jyang21/sbhmp/filtered_RAW_SAP_BQSR.vcf_v5_060924 --freq --missing --pca --out /work/jyanglab/jyang21/sbhmp/filtered_RAW_SAP_BQSR.vcf_v5_060924
```

--geno: 2188 variants removed due to missing genotype data.
27949948 variants removed due to minor allele threshold(s)
(--maf/--max-maf/--mac/--max-mac).
7983642 variants remaining after main filters.

### change RNA-seq .fam file to indicate its source

```{r}
fam <- read.delim(file="/work/jyanglab/jyang21/sbhmp/SbDiv_imputed_Filter2.recode.fam", header=FALSE)
fam$V1 <- paste0(fam$V1, "_R")
fam$V2 <- paste0(fam$V2, "_R")
write.table(fam, "/work/jyanglab/jyang21/SbDiv_imputed_Filter2.recode.fam", sep="\t", row.names = FALSE, col.names = FALSE, quote=FALSE)
```

### Keep only shared SNPs

```{r}
library(data.table)

df1 <- fread("/work/jyanglab/jyang21/sbhmp/SbDiv_imputed_Filter2.afreq", data.table=FALSE, header=T)

df2 <- fread("/work/jyanglab/jyang21/sbhmp/filtered_RAW_SAP_BQSR.vcf_v5_060924.afreq", data.table=FALSE, header=T)

df2a <- subset(df2, ID %in% df1$ID)
# 152512      6

df1a <- subset(df1, ID %in% df2$ID)

df1a <- subset(df1a, !(ID %in% c("Chr02_78582483", "Chr07_64458720", "Chr09_47854574")) )

write.table(df1a[,2], "/work/jyanglab/jyang21/sbhmp/shared_snpid.txt", sep="\t", col.names = F, row.names = F, quote=F)
```

# remove these three SNPs due to multi-allele after merging that cause error!

Chr02_78582483
Chr07_64458720
Chr09_47854574


### start to merge the data

filtered_RAW_SAP_BQSR.vcf_v5_06092
SbDiv_imputed_Filter2.recode

```{bash}
ml plink
cd /work/jyanglab/jyang21/sbhmp/

plink --bfile SbDiv_imputed_Filter2.recode --extract shared_snpid.txt --make-bed --out shared/RNA_seq
# 152509 variants and 713 people pass filters and QC.
plink --bfile filtered_RAW_SAP_BQSR.vcf_v5_060924 --extract shared_snpid.txt --make-bed --out shared/genomic
# 152509 variants and 400 people pass filters and QC.

plink --bfile shared/genomic --bmerge shared/RNA_seq --make-bed --out shared/merged 
```

152509 variants loaded from .bim file.
1113 people (0 males, 0 females, 1113 ambiguous) loaded from .fam.

```{bash}
plink -bfile shared/merged --genome 'full' --ppc-gap 10000 --out shared/dist
```


```{r}
library(data.table)
d <- fread("/work/jyanglab/jyang21/sbhmp/shared/dist.genome", header=T)
# keep only comparison between RNA-seq and genomic

d$s1 <- gsub(".*_", "", d$FID1)
d$s2 <- gsub(".*_", "", d$FID2)

d <- subset(d, s1 != s2)
d <- subset(d, s2 == "R")
d$IID2 <- gsub("_R", "", d$IID2)

d <- d[order(d$FID1, d$DST, decreasing = T), ]

write.table(d, "largedata/dist.txt", sep="\t", row.names = F, quote=FALSE)

```


