---
title: "Comparison results"
author: "Jinliang"
date: "2024-07-10"
output: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=normalizePath('..//..//')) 
```


Move `largedata/dist.txt` from HCC to local computer


```{r}
library("data.table")
d <- fread("largedata/dist.txt")

d0 <- subset(d, IID1 != IID2)
hist(d0$DST, main="distance using IBS", xlab="IBS distance", breaks = 100, col="lightblue")
```

First of all, for the label matched samples, let's take a look at their stats

```{r}
# only 295, all the others are different genotypes!
d1 <- subset(d, IID1 == IID2)

hist(d1$DST, main="distance using IBS", xlab="IBS distance", breaks = 100, col="lightblue")
nrow(subset(d1, DST > 0.85)) # 262
nrow(subset(d1, DST > 0.9)) # 246
nrow(subset(d1, DST > 0.95)) # 225


pdf("graphs/ibs_res.pdf", width = 5, height=5)
hist(d1$DST, main="distance using IBS", xlab="IBS distance", breaks = 100, col="lightblue")
abline(v=0.95, col="red", lty=2, lwd=3)
abline(v=0.9, col="blue", lty=2, lwd=3)
dev.off()
```

# how many of these matched labels showed the highest IBS

```{r}
# select the top IBS
d <- d[order(d$FID1, d$DST, decreasing = T),]
d2 <- d[!(duplicated(d$FID1)), ]

d3 <- merge(d2, d1[, c("FID1", "DST")], by="FID1")
d3$diff <- d3$DST.x - d3$DST.y

nrow(subset(d3, diff < 0.05)) #273
a <- subset(d3, diff > 0.05)
fwrite(a, "data/xlabels.csv", sep=",", row.names = FALSE, quote=FALSE)
#a <- subset(d3, diff > 0.01 & DST.x > 0.8 & DST.y < 0.85)

plot(d3$DST.y, d3$DST.x, xlab="Matched", ylab = "Highest IBS Score", ylim=c(0.6, 1))
abline(coef = c(0,1), lty=2)
points(a$DST.y, a$DST.x, pch=4, col="red")
```



```{r}
length(unique(d$FID1))
# select the top matching id
d1 <- d[!(duplicated(d$FID1)), ]

pdf("graphs/distance.pdf", width=6, height=5)
hist(d1$DST, main="distance using IBS", xlab="IBS distance", breaks = 100, col="lightblue")
dev.off()
```


```{r}
s1 <- subset(d1, DST > 0.90)
```


-------------

# Double-checking Hongyu's correction

I am attaching the name conversion I did as we discussed in the morning. In the file SAP.crazynames2.csv, the first column is the original name from the vcf file downloaded from the EVA, and the second column is the new name I assign. The file SAP.matchdiy.xlsx is the is the spreadsheet being screenshot in the slides.

```{r}
x <- read.delim("data/SAP.crazynames2.csv", sep=" ", header=FALSE)
```


let's see if any of the labels appear twice.
```{r}
b <- c(a$IID1, a$IID2)
b <- as.data.frame(table(b))
b <- b[order(b$Freq, decreasing=T),]
```

