---
title: "Comparison between RNA-seq and genomics SNP sets"
author: "Jinliang"
date: "2024-07-30"
output: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/work/jyanglab/subhash/variant_calling")
```


/work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode.vcf.gz

# 1. convert vcf to plink2 bed format 

```{bash}
ml bcftools
ml vcftools
ml plink2

#RNA seq SNPs
plink2 -vcf /work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode.vcf.gz --set-all-var-ids @_#_\$1_\$2 --make-bed --out /work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode
# WGS SNPs
plink2 -vcf /work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/SAP_V5_annotate.vcf --set-all-var-ids @_#_\$1_\$2 --make-bed --out /work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/RAW_SAP_BQSR.vcf_v5_060924

# --vcf: 35935778 variants scanned.

```

713 samples 
170608 variants

### 2. Get SNP stat

```{bash}
#RNA_SEQ
plink2 -bfile /work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode --freq --missing --pca --out /work/jyanglab/subhash/1.RNA_Seq/1.test/SbDiv_imputed_Filter2  

# --ibc (ported from GCTA) calculates three inbreeding coefficients for each sample
#WGS
plink2 -bfile /work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/RAW_SAP_BQSR.vcf_v5_060924 --freq --missing --pca --out /work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/3.test/RAW_SAP_BQSR.vcf_v5_060924 
```

--freq: Allele frequencies (founders only) written to
/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.afreq .
--missing: Sample missing data report written to
/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.smiss .
--missing: Variant missing data report written to
/work/jyanglab/jyang21/sbhmp/RAW_SAP_BQSR.vcf_v5_060924.vmiss .



# RNA-seq SNPs

```{r}
library(data.table)

df1 <- fread("SbDiv_imputed_Filter2.afreq", data.table=FALSE, header=T)
# 170608
range(df1$ALT_FREQS)
# 0.0504909 0.9495090; 

df2 <- fread("SbDiv_imputed_Filter2.vmiss", data.table=FALSE, header=T)
mean(df2$F_MISS)
# 0: imputed no missing data

df3 <- fread("SbDiv_imputed_Filter2.smiss", data.table=FALSE, header=T)
# 713 samples!


pdf("profiling/1.variant_calling/RNA-seq_frq_tem.pdf", width=6, height=5)
hist(df1$ALT_FREQS, main="Alt Allele Freq", xlab="Allele Freq", breaks = 100, col="lightblue")
abline(v=0.05, col="red", lty=2, lwd=3)
abline(v=0.95, col="red", lty=2, lwd=3)
hist(df3$F_MISS, main="Count by locus", xlab="Variant missingness", breaks = 100, col="lightgreen")
abline(v=0.1, col="red", lty=2, lwd=3)
dev.off()
```



# Genomics SNPs

```{r}
library(data.table)

df1 <- fread("RAW_SAP_BQSR.vcf_v5_060924.afreq", data.table=FALSE, header=T)
# 28,461,719
str(df1)
sum(is.na(df1$ALT_FREQS))
#  10114
df1$ALT_FREQS <- as.numeric(df1$ALT_FREQS)

# Remove rows with NA in ALT_FREQS into df1_clean
df1_clean <- df1[!is.na(df1$ALT_FREQS), ]
range(df1_clean$ALT_FREQS)
# 0.00125 1.00000 
# 0 1

nrow(subset(df1, ALT_FREQS > 0.05 & ALT_FREQS < 0.95))
# 7,889,507
# 6,968,602

df2 <- fread("RAW_SAP_BQSR.vcf_v5_060924.smiss", data.table=FALSE, header=T)
mean(df2$F_MISS)
# 7.735799e-05: very low individual missing rate
# 0.1083693: 0.1 missing rate
# 400 samples

df3 <- fread("RAW_SAP_BQSR.vcf_v5_060924.vmiss", data.table=FALSE, header=T)
# 713 samples!
mean(df3$F_MISS)
# 0.1083693
nrow(subset(df3, df3$F_MISS < 0.3))
# 35,933,485
# 24,567,887
nrow(subset(df3, df3$F_MISS > 0.1))
# 7802008

pdf("profiling/1.variant_calling/SAP_raw_frq.pdf", width=12, height=4)
par(mfrow = c(1, 3))  ## set the layout
hist(df1$ALT_FREQS, main="Alt Allele Freq", xlab="Allele Freq", breaks = 100, col="lightblue")
abline(v=0.05, col="red", lty=2, lwd=3)
abline(v=0.95, col="red", lty=2, lwd=3)
hist(df2$F_MISS, main="Count by sample", xlab="Sample missingness", breaks = 100, col="purple")
hist(df3$F_MISS, main="Count by locus", xlab="Variant missingness", breaks = 100, col="lightgreen")
abline(v=0.1, col="red", lty=2, lwd=3)
dev.off()
```


# Merge the two datasets together

filtered genomics variants:
/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/

```{bash}
plink2 --bfile /work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/RAW_SAP_BQSR.vcf_v5_060924 --maf 0.05 --geno 0.1 --make-bed --out /work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/3.test/1.filter/filtered_RAW_SAP_BQSR.vcf_v5_060924

plink2 -bfile /work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/3.test/1.filter/filtered_RAW_SAP_BQSR.vcf_v5_060924 --freq --missing --pca --out //work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/3.test/1.filter/filtered_RAW_SAP_BQSR.vcf_v5_060924
```

# subhash
--geno: 7802008 variants removed due to missing genotype data.
15087546 variants removed due to minor allele threshold(s)
(--maf/--max-maf/--mac/--max-mac).
5572165 variants remaining after main filters.


### change RNA-seq .fam file to indicate its source

```{r}
fam <- read.delim(file="/work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode.fam", header=FALSE)
fam$V1 <- paste0(fam$V1, "_R")
fam$V2 <- paste0(fam$V2, "_R")
write.table(fam, "/work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode.fam", sep="\t", row.names = FALSE, col.names = FALSE, quote=FALSE)
```

### Keep only shared SNPs

```{r}
library(data.table)

df1 <- fread("/work/jyanglab/subhash/1.RNA_Seq/1.test/SbDiv_imputed_Filter2.afreq", data.table=FALSE, header=T)

df2 <- fread("/work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/3.test/filtered_RAW_SAP_BQSR.vcf_v5_060924.afreq", data.table=FALSE, header=T)

df2a <- subset(df2, ID %in% df1$ID)
# 149,895      6

df1a <- subset(df1, ID %in% df2$ID)

df1a <- subset(df1a, !(ID %in% c("Chr02_78582483", "Chr07_64458720", "Chr09_47854574")) )

write.table(df1a[,2], "/work/jyanglab/subhash/1.RNA_Seq/shared_snpid.txt", sep="\t", col.names = F, row.names = F, quote=F)
```

# remove these three SNPs due to multi-allele after merging that cause error!

Chr02_78582483
Chr07_64458720
Chr09_47854574


### start to merge the data

filtered_RAW_SAP_BQSR.vcf_v5_06092
SbDiv_imputed_Filter2.recode

```{bash}
ml plink

plink --bfile /work/jyanglab/subhash/1.RNA_Seq/SbDiv_imputed_Filter2.recode --extract /work/jyanglab/subhash/1.RNA_Seq/shared_snpid.txt --make-bed --out /work/jyanglab/subhash/1.RNA_Seq/shared/RNA_seq
# 149895 variants and 713 people pass filters and QC.

plink --bfile /work/jyanglab/subhash/variant_calling/9.gvcf_to_vcf/4.filtering/1.annotate/3.test/1.filter/filtered_RAW_SAP_BQSR.vcf_v5_060924 --extract /work/jyanglab/subhash/1.RNA_Seq/shared_snpid.txt --make-bed --out /work/jyanglab/subhash/1.RNA_Seq/shared/genomic
# 149895 variants and 400 people pass filters and QC.

plink --bfile /work/jyanglab/subhash/1.RNA_Seq/shared/genomic --bmerge /work/jyanglab/subhash/1.RNA_Seq/shared/RNA_seq --make-bed --out /work/jyanglab/subhash/1.RNA_Seq/shared/merged
# 149895 variants and 1113 people pass filters and QC.
```

149895 variants loaded from .bim file.
1113 people (0 males, 0 females, 1113 ambiguous) loaded from .fam.

```{bash}
plink -bfile /work/jyanglab/subhash/1.RNA_Seq/shared/merged --genome 'full' --ppc-gap 10000 --out /work/jyanglab/subhash/1.RNA_Seq/shared/dist
```


```{r}
library(data.table)
d <- fread("/work/jyanglab/subhash/1.RNA_Seq/shared/dist.genome", header=T)
# keep only comparison between RNA-seq and genomic

d$s1 <- gsub(".*_", "", d$FID1)
d$s2 <- gsub(".*_", "", d$FID2)

d <- subset(d, s1 != s2)
d <- subset(d, s2 == "R")
d$IID2 <- gsub("_R", "", d$IID2)

d <- d[order(d$FID1, d$DST, decreasing = T), ]

write.table(d, "/work/jyanglab/subhash/1.RNA_Seq/shared/dist.txt", sep="\t", row.names = F, quote=FALSE)
```



### 1.4 update 

Move `largedata/dist.txt` from HCC to local computer


```{r}
library("data.table")
d <- fread("/work/jyanglab/subhash/1.RNA_Seq/shared/dist.txt")

d0 <- subset(d, IID1 != IID2)
hist(d0$DST, main="distance using IBS", xlab="IBS distance", breaks = 100, col="lightblue")
```

First of all, for the label matched samples, let's take a look at their stats

```{r}
# only 295, all the others are different genotypes!
d1 <- subset(d, IID1 == IID2)

hist(d1$DST, main="distance using IBS", xlab="IBS distance", breaks = 100, col="lightblue")
nrow(subset(d1, DST > 0.9))  # 246 249
nrow(subset(d1, DST > 0.95)) # 225 226
nrow(subset(d1, DST > 0.85)) # 225 262

pdf("graphs/ibs_res.pdf", width = 5, height=5)
hist(d1$DST, main="distance using IBS", xlab="IBS distance", breaks = 100, col="lightblue")
abline(v=0.95, col="red", lty=2, lwd=3)
abline(v=0.9, col="blue", lty=2, lwd=3)
dev.off()
```

# how many of these matched labels showed the highest IBS

```{r}
# select the top IBS
d <- d[order(d$FID1, d$DST, decreasing = T),]
d2 <- d[!(duplicated(d$FID1)), ]

d3 <- merge(d2, d1[, c("FID1", "DST")], by="FID1")
d3$diff <- d3$DST.x - d3$DST.y

nrow(subset(d3, diff < 0.05)) #273
a <- subset(d3, diff > 0.05)
fwrite(a, "data/xlabels.csv", sep=",", row.names = FALSE, quote=FALSE)
#a <- subset(d3, diff > 0.01 & DST.x > 0.8 & DST.y < 0.85)

plot(d3$DST.y, d3$DST.x, xlab="Matched", ylab = "Highest IBS Score", ylim=c(0.6, 1))
abline(coef = c(0,1), lty=2)
points(a$DST.y, a$DST.x, pch=4, col="red")
```



```{r}
length(unique(d$FID1))
# select the top matching id
d1 <- d[!(duplicated(d$FID1)), ]

pdf("graphs/distance.pdf", width=6, height=5)
hist(d1$DST, main="distance using IBS", xlab="IBS distance", breaks = 100, col="lightblue")
dev.off()
```


```{r}
s1 <- subset(d1, DST > 0.90)
```


-------------

# Double-checking Hongyu's correction

I am attaching the name conversion I did as we discussed in the morning. In the file SAP.crazynames2.csv, the first column is the original name from the vcf file downloaded from the EVA, and the second column is the new name I assign. The file SAP.matchdiy.xlsx is the is the spreadsheet being screenshot in the slides.

```{r}
x <- read.delim("data/SAP.crazynames2.csv", sep=" ", header=FALSE)
```


let's see if any of the labels appear twice.
```{r}
b <- c(a$IID1, a$IID2)
b <- as.data.frame(table(b))
b <- b[order(b$Freq, decreasing=T),]
```



