---
title: "Domestication_trajectory"
author: "subhash"
date: "2025-08-22"
output: html_document
---
After getting the output of B2 scores in all the 3 populaions
I have merged the each chromosome and merged all the chromosomes into one big file.

# landraces
creating the bed file for the top one perecent
filtered the files from top 1% B2 scores

```{bash}
#create top 1% threshold
#output of all the lines are available at largedata on HCC

cd /work/jyanglab/subhash/git/GWAS_sorghum_seedling/largedata/

sort -k4,4nr /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/Sorghum_landrace.AGPV5.B2_stat.txt | head -n $(($(wc -l < /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/Sorghum_landrace.AGPV5.B2_stat.txt )/100)) > /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.txt

#convert to bed file
awk -v OFS="\t" '{start=$2-40000; if(start<0) start=0; end=$2+40000; print $1, start, end, $4}' /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.txt > /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.40kb.bed

#sort the bed file according to the chromosomes
sort -k1,1V -k2,2n /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.40kb.bed > /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.40kb.sorted.bed

wc -l /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.40kb.sorted.bed
#147267

```

#wild population 
#improved lines
## Domestication trajectory
creating the bed files for the wild and improved population
creating the bed files from the balancing selection results
```{bash}
#wild population 
#generating the bed file and sorting based on chromomes
awk -v OFS="\t" '{start=$1-40000; if(start<0) start=0; end=$1+40000; print $7, start, end, $3}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/wild.top1.txt > wild.40kb.bed

sort -k1,1V -k2,2n wild.40kb.bed > wild.40kb.sorted.bed

wc -l wild.40kb.sorted.bed 
#213176

#improved lines
awk -v OFS="\t" '{start=$1-40000; if(start<0) start=0; end=$1+40000; print $7, start, end, $3}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.top1.txt > /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.40kb.bed

sort -k1,1V -k2,2n imp.40kb.bed > imp.40kb.sorted.bed

wc -l imp.40kb.sorted.bed 
#94048

```

# looking for the overlap in bed files within bal

overall, I have opbtained 60 loci that are overlapped between all these 3 populaions
(there are all under balancing selection throughout domestication and breeding)

```{bash}
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input
ml bedtools
#merging the nearby positions and kept the highest value
bedtools merge -i wild.40kb.sorted.bed -d 40000 -c 4 -o max > wild.40kb.merged.bed
bedtools merge -i land.40kb.sorted.bed -d 40000 -c 4 -o max > land.40kb.merged.bed
bedtools merge -i imp.40kb.sorted.bed -d 40000 -c 4 -o max > imp.40kb.merged.bed

cd /work/jyanglab/subhash/sorgsd/bal_s/3.input
bedtools intersect -a wild.40kb.merged.bed -b land.40kb.merged.bed -wa -wb > W_L_overlap.bed
wc -l W_L_overlap.bed 
#137
bedtools intersect -a W_L_overlap.bed -b imp.40kb.merged.bed -wa -wb > W_L_I_overlap.bed
wc -l W_L_I_overlap.bed 
#60

#These 60 loci are conserved balancing selection targets the selective pressure that maintains variation in wild populations has persisted through domestication and improvement.
#They are high confidence candidates for important functional genes or regulatory regions.
###These are loci where genetic diversity is actively maintained across evolutionary and breeding processes.
```

# to find unique peaks, I am looking at wild and landrace, improved individually


```{bash}
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input

ml bedtools

bedtools intersect -a /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild.40kb.merged.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/6.BT623_gene.V5_up_dw_2k.bed -wa -wb > Wild_only_with_gene_overlap.bed

bedtools intersect -a /work/jyanglab/subhash/sorgsd/bal_s/3.input/land.40kb.merged.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/6.BT623_gene.V5_up_dw_2k.bed -wa -wb > Landrace_only_with_gene_overlap.bed

bedtools intersect -a /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp.40kb.sorted.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/6.BT623_gene.V5_up_dw_2k.bed -wa -wb > Improved_only_with_gene_overlap.bed

```


# Finding the gene description for the overlapping loci 

```{r}
library(readxl)
library(data.table)
library(dplyr)
library(writexl)

#d = fread("/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/B2/Wild_only_with_gene_overlap.bed")

#d = fread("/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/B2/Landrace_only_with_gene_overlap.bed")

d = fread("/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/B2/Improved_only_with_gene_overlap.bed")

dim(d)
a = read_excel("/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/7.sorghum.annotations.xlsx", sheet = 2)
dim(a)
a_unique <- a %>%
  distinct(Gene, .keep_all = TRUE)

merged <- d %>%
  left_join(a_unique, by = c("V9" = "Gene"))
dim(merged)
writexl::write_xlsx(merged, "/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/B2/Improved_only_bal_description.xlsx")

#writexl::write_xlsx(merged, "/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/B2/Landrace_only_bal_description.xlsx")

#writexl::write_xlsx(merged, "/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/B2/Wild_only_bal_description.xlsx")

```


# seeing the overlap between the bal and pos

```{bash}
bedtools intersect -a wild.40kb.merged.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/positive/wild_vs_landrace.bed -wa -wb > pos_1_bal_wild_landrace.bed

bedtools intersect -a wild.40kb.merged.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/positive/landrace_vs_improved.bed -wa -wb > pos_2_bal_landrace_vs_improved.bed
```

"what is loci?
A locus is a genomic region defined by merged B2 windows (+,- 40kb from physPos).
Each locus can include multiple SNPs or variants, but it is considered one signal of balancing selection.
By merging nearby windows (±40 kb and adjacent regions), we avoid counting the same signal multiple times."

```{bash}
/work/jyanglab/subhash/sorgsd/bal_s/3.input

awk '{OFS="\t"; print $NF, $1-1, $1, $3}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/wild.top1.txt > 0.1wild.sites.bed

sort -k1,1 -k2,2n 0.1wild.sites.bed > 0.01wild.sites.sorted.bed

bedtools intersect -a 0.01wild.sites.sorted.bed -b W_L_I_overlap.bed -wa > 0.001wild_in_overlap.bed

bedtools intersect -a 0.01wild.sites.sorted.bed -b W_L_I_overlap.bed -wa -wb > 0.0001wild_overlap_with_locus.bed

awk '{
    key = $5":"$6; 
    if($4>max[key]){max[key]=$4; line[key]=$0}
} END{for(k in line) print line[k]}' 0.0001wild_overlap_with_locus.bed > wild_leading_SNPs.bed
#60
```

Improved 

```{bash}
/work/jyanglab/subhash/sorgsd/bal_s/3.input

awk '{OFS="\t"; print $NF, $1-1, $1, $3}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.top1.txt > 0.1imp.sites.bed
#94048 0.1imp.sites.bed
sort -k1,1 -k2,2n 0.1imp.sites.bed > 0.01imp.sites.sorted.bed
#94048 0.01imp.sites.sorted.bed
bedtools intersect -a 0.01imp.sites.sorted.bed -b W_L_I_overlap.bed -wa > 0.001imp_in_overlap.bed
#40620 0.001imp_in_overlap.bed
bedtools intersect -a 0.01imp.sites.sorted.bed -b W_L_I_overlap.bed -wa -wb > 0.0001imp_overlap_with_locus.bed
#40620 0.0001imp_overlap_with_locus.bed
awk '{
    key = $5":"$6; 
    if($4>max[key]){max[key]=$4; line[key]=$0}
} END{for(k in line) print line[k]}' 0.0001imp_overlap_with_locus.bed > imp_leading_SNPs.bed
#59 imp_leading_SNPs.bed
```

landraces

```{bash}
/work/jyanglab/subhash/sorgsd/bal_s/3.input

awk '{OFS="\t"; print $1, $2-1, $2, $4}' /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.txt > 0.1land.sites.bed
 wc -l 0.1land.sites.bed
#147267

sort -k1,1 -k2,2n 0.1land.sites.bed > 0.01land.sites.sorted.bed
 wc -l 0.01land.sites.sorted.bed
#147267

bedtools intersect -a 0.01land.sites.sorted.bed -b W_L_I_overlap.bed -wa > 0.001land_in_overlap.bed
#55858 
bedtools intersect -a 0.01land.sites.sorted.bed -b W_L_I_overlap.bed -wa -wb > 0.0001land_overlap_with_locus.bed

awk '{
    key = $5":"$6; 
    if($4>max[key]){max[key]=$4; line[key]=$0}
} END{for(k in line) print line[k]}' 0.0001land_overlap_with_locus.bed > land_leading_SNPs.bed
#57 
```

# calculating allele freequecies from all 3 sub-populations

```{bash}
#!/bin/sh
#SBATCH --job-name=freq
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/frq_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/frq_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

cd /work/jyanglab/subhash/sorgsd/bal_s

ml vcftools
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/wild_50_samples.recode.vcf --freq --out wild_freq
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/landraces.recode.vcf --freq --out landrace_freq
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/imp_129_samples.recode.vcf --freq --out improved_freq
```

taking the allele frequencies for the leading SNPs in all the 3 sub-populations
extracting the minor allele freequecy for those leading SNPs

```{bash}
# Example for wild
awk 'NR==FNR {pos[$1":"$2]=1; next} 
     ($1":"$2) in pos {split($5,a,":"); split($6,b,":"); 
     freq=(a[2]<=b[2]?a[2]:b[2]); 
     print $1,$2,freq}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild_leading_SNPs_positions.sorted.bed /work/jyanglab/subhash/sorgsd/bal_s/wild_freq.frq > wild_leading_SNPs.af
     
awk 'NR==FNR {pos[$1":"$2]=1; next} 
     ($1":"$2) in pos {split($5,a,":"); split($6,b,":"); 
     freq=(a[2]<=b[2]?a[2]:b[2]); 
     print $1,$2,freq}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild_leading_SNPs_positions.sorted.bed /work/jyanglab/subhash/sorgsd/bal_s/landrace_freq.frq > land_leading_SNPs.af
     
awk 'NR==FNR {pos[$1":"$2]=1; next} 
     ($1":"$2) in pos {split($5,a,":"); split($6,b,":"); 
     freq=(a[2]<=b[2]?a[2]:b[2]); 
     print $1,$2,freq}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild_leading_SNPs_positions.sorted.bed /work/jyanglab/subhash/sorgsd/bal_s/improved_freq.frq > imp_leading_SNPs.af     

```

Extracting minor allele frequencies for all the leading SNPs in all the 3 sub-popualtions in same file 

```{bash}
paste wild_leading_SNPs.af land_leading_SNPs.af imp_leading_SNPs.af | \
awk '{OFS="\t"; print $1":"$2, $3, $6, $9}' > trajectory_table.txt
wc -l trajectory_table.txt
#60
```

I thought to see the leading SNPs allele frequencies pattern of these 3 sub-popualtions
I ploted both the average MAF accross these leading SNPs in each sub-population and 
also plotted the individual leading SNP MAF in each of these 3 sub-pop

```{r}
library(tidyverse)

# Load trajectory table
traj <- read.table("trajectory_table.txt", header = FALSE, sep = "\t",
                   col.names = c("SNP", "Wild", "Landrace", "Improved"))

traj_long <- traj %>%
  pivot_longer(cols = c("Wild", "Landrace", "Improved"),
               names_to = "Population",
               values_to = "MAF")

# Individual SNP trajectories
p1 <- ggplot(traj_long, aes(x = Population, y = MAF, group = SNP, color = SNP)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Allele Frequency Trajectories of Leading SNPs",
       x = "Population",
       y = "Minor Allele Frequency") +
  theme(legend.position = "none")  # hide legend if too many SNPs

#PDF
#ggsave("wild_SNP_trajectories.pdf", plot = p1, width = 8, height = 6)

#PNG
ggsave("wild_SNP_trajectories.png", plot = p1, width = 8, height = 6, dpi = 300)

#Average MAF per population
avg_traj <- traj_long %>%
  group_by(Population) %>%
  summarise(mean_MAF = mean(MAF))

#Plot2
p2 <- ggplot(avg_traj, aes(x = Population, y = mean_MAF, group = 1)) +
  geom_line(size = 1.2, color = "blue") +
  geom_point(size = 3, color = "red") +
  theme_minimal() +
  labs(title = "Average Allele Frequency Trajectory",
       x = "Population",
       y = "Average Minor Allele Frequency")

#ggsave("average_SNP_trajectory.pdf", plot = p2, width = 6, height = 5)
ggsave("average_SNP_trajectory.png", plot = p2, width = 6, height = 5, dpi = 300)

```

#identifying the common genes

```{bash}

cd /work/jyanglab/subhash/sorgsd/bal_s/3.input

ml bedtools

bedtools intersect -a /work/jyanglab/subhash/sorgsd/bal_s/3.input/W_L_I_overlap.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/6.BT623_gene.V5_up_dw_2k.bed -wa -wb > W_L_I_with_gene_overlap.bed

```

```{r}
library(readxl)
library(data.table)
library(dplyr)
library(writexl)

d = fread("/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/positive_selection/W_L_I_with_gene_overlap.bed")

d = fread("/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/B2/W_L_I_with_gene_5kb.bed")

dim(d)
a = read_excel("/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/7.sorghum.annotations.xlsx", sheet = 2)
dim(a)
a_unique <- a %>%
  distinct(Gene, .keep_all = TRUE)

merged <- d %>%
  left_join(a_unique, by = c("V20" = "Gene"))
dim(merged)
writexl::write_xlsx(merged, "/Users/subhashmahamkali/Downloads/1.miscellaneous/sorghum_project/B2/W_L_I_5kb_description.xlsx")
```
# Finding genes overlapping with top 1% B2 scores
date: 2025-09-08


```{bash}
# 1. SNPs to 1-bp BED intervals (chr, start=pos-1, end=pos, score=LR)
awk 'NR>1{print $7,$1-1,$1,$3}' OFS="\t" wild.top1.txt \
  | sort -k1,1V -k2,2n > wild.top1.sites.bed
  
# 2. Merge nearby sites within 10 kb, keep SNP count per region
#bedtools merge -i wild.top1.sites.bed -d 10000 -c 4 -o count > wild.top1.merged.bed
bedtools merge -i wild.top1.sites.bed -d 10000 -c 4 -o count,max > wild.top1.merged.bed


# 3. Keep only regions with ≥2 significant SNPs
awk '$4>=2' wild.top1.merged.bed > wild.top1.min2.bed

# 4. Extend each region ±5 kb
awk 'BEGIN{OFS="\t"}{start=$2-5000; if(start<0) start=0; end=$3+5000; print $1,start,end,$4,$5}' \
  wild.top1.min2.bed > wild.top1.ext5k.bed

#5. Intersect with genes (needs genes.bed from your GFF3)
bedtools intersect -a wild.top1.ext5k.bed -b /work/jyanglab/subhash/sorgsd/bal_s/3.input/6.BT623_gene.V5_up_dw_2k.bed -wa -wb > wild.top1.genehits.bed
```

```{bash}
awk 'NR>1{print $7,$1-1,$1,$3}' OFS="\t" imp.top1.txt \
  | sort -k1,1V -k2,2n > imp.top1.sites.bed
  
bedtools merge -i imp.top1.sites.bed -d 10000 -c 4 -o count,max > imp.top1.merged.bed

awk '$4>=2' imp.top1.merged.bed > imp.top1.min2.bed

awk 'BEGIN{OFS="\t"}{start=$2-5000; if(start<0) start=0; end=$3+5000; print $1,start,end,$4,$5}' \
  imp.top1.min2.bed > imp.top1.ext5k.bed

bedtools intersect -a imp.top1.ext5k.bed -b /work/jyanglab/subhash/sorgsd/bal_s/3.input/6.BT623_gene.V5_up_dw_2k.bed -wa -wb > imp.top1.genehits.bed


#landraces gene overlap.
cd /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/
awk 'NR>1{print $1,$2-1,$2,$4}' OFS="\t" land.top1.txt \
  | sort -k1,1V -k2,2n > land.top1.sites.bed
  
bedtools merge -i land.top1.sites.bed -d 10000 -c 4 -o count,max > land.top1.merged.bed

awk '$4>=2' land.top1.merged.bed > land.top1.min2.bed

awk 'BEGIN{OFS="\t"}{start=$2-5000; if(start<0) start=0; end=$3+5000; print $1,start,end,$4,$5}' \
  land.top1.min2.bed > land.top1.ext5k.bed

bedtools intersect -a land.top1.ext5k.bed -b /work/jyanglab/subhash/sorgsd/bal_s/3.input/6.BT623_gene.V5_up_dw_2k.bed -wa -wb > land.top1.genehits.bed
#/mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.genehits.bed

cd /work/jyanglab/subhash/sorgsd/bal_s/3.input

bedtools intersect -a /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/wild.top1.ext5k.bed -b /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.top1.ext5k.bed -wa -wb > W_I_5kb.bed

wc -l W_I_5kb.bed 
#74
bedtools intersect -a /work/jyanglab/subhash/sorgsd/bal_s/3.input/W_I_5kb.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.ext5k.bed -wa -wb > W_L_I_5kb.bed
# df = one of dd_filt (wild), la_filt (land), ii_filt (improved)
# Assumes df has columns: physPos (Mb) and B2 in column 3

peak_y <- function(df, pos_mb, window_mb = 0.20) {   # 0.20 Mb = ±200 kb (tune as needed)
  lo <- pos_mb - window_mb
  hi <- pos_mb + window_mb
  idx <- df$physPos >= lo & df$physPos <= hi
  if (!any(idx)) return(NA_real_)
  max(df[idx, 3], na.rm = TRUE)
}

# For your selected points:
red_points1$y_peak <- vapply(red_points1$pos, peak_y, numeric(1), df = dd_filt)  # wild
red_points2$y_peak <- vapply(red_points2$pos, peak_y, numeric(1), df = la_filt)  # land
red_points3$y_peak <- vapply(red_points3$pos, peak_y, numeric(1), df = ii_filt)  # improved

wc -l W_L_I_5kb.bed 
#57
bedtools intersect -a /work/jyanglab/subhash/sorgsd/bal_s/3.input/W_L_I_5kb.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/6.BT623_gene.V5_up_dw_2k.bed -wa -wb > W_L_I_with_gene_5kb.bed

wc -l W_L_I_with_gene_5kb.bed
#159
```






