---
title: "1.Balancing selection ananlysis (B2 score)"
output: html_notebook
author: "subhash"
date: "2025-08-13"
---
# Filtering the vcf files for the wild and improved lines along with S.propinqaum (ancestor allele)
using vcftools subset the 3 .vcf files for wild, improved and ancestor
also calculated the allele freequecny in ancestor(#3) samples

```{bash}
#!/bin/sh
#SBATCH --job-name=filter_missing
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/filter_missing_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/filter_missing_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml vcftools

vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/propinqaum_sample.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample

vcftools --vcf propinqaum_3sample.recode.vcf --freq --out /work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample

vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/wild.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/wild_50_samples

vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/improved.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/imp_129_samples
```


submitting below R code as slurm job on HCC unl

```{bash}
#!/bin/sh
#SBATCH --job-name=ans
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml R
Rscript /work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans.R
```

## Calculating ancestor allele freq and count for all the chromosomes at once from the outgroup

```{r}
library(data.table)

d = fread("/work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample.frq", header=F, data.table=TRUE)
a1 = unlist(strsplit(d[[5]], ":"))
allele1 = a1[seq(1, length(a1), by=2)]
freq1 = as.numeric(a1[seq(2, length(a1), by=2)])
a2 = unlist(strsplit(d[[6]], ":"))
allele2 = a2[seq(1, length(a2), by=2)]
freq2 = as.numeric(a2[seq(2, length(a2), by=2)])

#after spliting giving the allele 1 and freq 1 and same for allele 2 and freq2

d1 = data.table(d[, 1:2], allele1, freq1, allele2, freq2)
setnames(d1, c("Chr", "Pos", "allele1", "Freq1", "allele2", "Freq2"))

#consider whichever allele is higher is considered to be ancestral allele
d1[, ancestral_allele := ifelse(Freq1 >= Freq2, allele1, allele2)]
fwrite(d1, "/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/Sorghum_ancestral_allele_V3.1.txt", sep = "\t")
```

### Calculating allele counts for both the wild and improved lines
To caculate how many derived alleles at each site

```{bash}
#!/bin/sh
#SBATCH --job-name=frq
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/2.frq/frq_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/2.frq/frq_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml vcftools
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/wild_50_samples.recode.vcf --counts --out /work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/imp_129_samples.recode.vcf --counts --out /work/jyanglab/subhash/sorgsd/bal_s/2.frq/imp_counts
```



```{bash}
#!/bin/sh
#SBATCH --job-name=inpu_w
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/3.input/inpW_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/3.input/inpW_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml R
Rscript /work/jyanglab/subhash/sorgsd/bal_s/3.input/in_w.R
```


here I did subset to work on small sample set to test the pipeline once, it sucessfully generated the output files I have run the whole pipeline on large dataset

```{bash}
head -n 101 /work/jyanglab/subhash/sorgsd/bal_s/Sorghum_ancestral_allele_V3.1.txt > /work/jyanglab/subhash/sorgsd/bal_s/subset/anc_subset.txt
head -n 101 /work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts.frq.count > /work/jyanglab/subhash/sorgsd/bal_s/subset/wild_counts_subset.frq.count
head -n 101 imp_counts.frq.count > /work/jyanglab/subhash/sorgsd/bal_s/subset/imp_counts_subset.frq.count
head -n 1000 /work/jyanglab/subhash/sorgsd/bal_s/subset/BalLeRMix_input_chr1.filtered.txt > 1000.txt
```

#### Generating the input file for the BalLeRiMix 

This below script helps to calculate derived allele count (which is opposite of ancester allele)
```{r}
library(data.table)

anc <- fread("/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/Sorghum_ancestral_allele_V3.1.txt")

wild <- fread("/work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts.frq.count", skip = 1,
              col.names = c("Chr", "Pos", "N_alleles", "n", "allele1_count", "allele2_count"))

setnames(wild, c("Chr", "Pos", "N_alleles", "n", "count1", "count2"))

wild[, count1_num := as.integer(sub("^[A-Z]:", "", count1))]
wild[, count2_num := as.integer(sub("^[A-Z]:", "", count2))]

#chromsome by chromsome loop here to calculate derived allele count in each site ( opposite of ancester allel) 

for (chr in unique(anc$Chr)) {
  anc_chr <- anc[Chr == chr]
  wild_chr <- wild[Chr == chr]
  wild_merged_chr <- merge(anc_chr, wild_chr, by = c("Chr", "Pos"))
  wild_merged_chr[, x := ifelse(ancestral_allele == allele1, count2_num, count1_num)]
  wild_input_chr <- wild_merged_chr[, .(physPos = Pos, 
                                        genePos = Pos / 1e6, 
                                        x = x, 
                                        n = n)]
#writingg the input file for BalLeRMix
  fwrite(wild_input_chr, 
         paste0("/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild_BalLeRMix_input_chr", chr, ".txt"), 
         sep = "\t", col.names = FALSE)
}
```



```{bash}
#copying the Ballerimix input files (for each chromosome) in to different directories (by creating new ones) to keep the files organized and then filtereing the empty rows
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild
for chr in {1..10}; do
    mkdir -p chr_${chr}
    # Copy original input
    cp imp_BalLeRMix_input_chr${chr}.txt chr_${chr}/
    # Filter out monomorphic sites
    awk '$3>0 && $3<$4' chr_${chr}/imp_BalLeRMix_input_chr${chr}.txt \
        > chr_${chr}/imp_BalLeRMix_input_chr${chr}.filtered.txt
done


#splitting the each chromosome file into 5k varinats for easy calculation of B2 score
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild
for chr in {1..10}; do
    cd chr_${chr}
    split -l 5000 --numeric-suffixes=1 --suffix-length=4 \
        imp_BalLeRMix_input_chr${chr}.filtered.txt chunk_
    cd ..
done

#generatig the spectrun DAF file for each chrmosome 
#SFS calculation from the BalLeriMix to use to calculate the B2 score
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild
for chr in {1..10}; do
    python /mnt/nrdstor/jyanglab/subhash/software/BallerMix/software/BalLeRMix_v2.5.py \
        -i chr_${chr}/imp_BalLeRMix_input_chr${chr}.filtered.txt \
        --getSpect \
        --spect chr_${chr}/chr${chr}.spect.txt
done


#counting number of chunks in each chromosome (just for teh ease of jobsubimiting on HCC unl)
for chr in {1..10}; do
    echo "chr_${chr}: $(ls chr_${chr}/chunk_* | wc -l) chunks"
done
#chr_1: 207 chunks
#chr_2: 226 chunks
#chr_3: 202 chunks
#chr_4: 214 chunks
#chr_5: 176 chunks
#chr_6: 167 chunks
#chr_7: 156 chunks
#chr_8: 181 chunks
#chr_9: 175 chunks
#chr_10: 182 chunks

#creating log files to get the out and error files for each job
for chr in {1..10}; do
    mkdir -p /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/chr_${chr}/logs
done
```


The below is the template I used to submit the BalLeriMix model to compute B2 score on each variant. using this template I can run many jobs to get the B2 stat score

```{bash}
#!/bin/bash
#SBATCH --job-name=chr${CHR}_wild
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=5:00:00
#SBATCH --mem=4G
#SBATCH --partition=batch,guest,jclarke
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/chr_${CHR}/logs/%A_%a.out.txt
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/chr_${CHR}/logs/%A_%a.err.txt
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu
#SBATCH --mail-type=END,FAIL

# Inputs
CHR=${CHR}   # passed via sbatch --export
CHUNK_DIR=/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/chr_${CHR}
SOFTWARE=/mnt/nrdstor/jyanglab/subhash/software/BallerMix/software/BalLeRMix_v2.5.py

# Format ID like 0001, 0002...
ID=$(printf "%04d" ${SLURM_ARRAY_TASK_ID})
FILE=$CHUNK_DIR/chunk_${ID}

echo ">>> Processing chromosome $CHR, chunk $ID"

python $SOFTWARE \
   -i "$FILE" \
   --spect $CHUNK_DIR/chr${CHR}.spect.txt \
   --physPos \
   -o "${FILE}.B2.txt"

echo ">>> Finished chromosome $CHR, chunk $ID$ CHUNK_DIR/chr${CHR}.spect.txt"

```

job submission by chromosome to chromsome
```{bash}
sbatch --export=CHR=1 --array=1-207 chr_array.slurm
sbatch --export=CHR=2 --array=1-226 chr_array.slurm
sbatch --export=CHR=3 --array=1-202 chr_array.slurm
sbatch --export=CHR=4 --array=1-214 chr_array.slurm
sbatch --export=CHR=5 --array=1-176 chr_array.slurm
sbatch --export=CHR=6 --array=1-167 chr_array.slurm
sbatch --export=CHR=7 --array=1-156 chr_array.slurm
sbatch --export=CHR=8 --array=1-181 chr_array.slurm
sbatch --export=CHR=9 --array=1-175 chr_array.slurm
sbatch --export=CHR=10 --array=1-182 chr_array.slurm
```


# these are mischlenus code from the begginer 
I use them to track my running jobs
since I submitted toomany jobs to see and track using naked eye 
the below code lines will help to 
```{bash}
squeue -j 11541491 | wc -l
sacct -j 11545344 --format=JobID,State
squeue -j 11545344 -o "%.18i %.9T" | wc -l
```


# merging all the chunk output files into one big file

```{bash}
#merging all the chunks in each chromosome as one file
for chr in {1..10}; do     out="chr_${chr}/chr${chr}_merged.B2.txt";     echo "Merging chromosome $chr → $out";     cat chr_${chr}/chunk_*.B2.txt > "$out"; done

#merging all the chromsomes in to one big file buy adding an extra column called chromosme
out="imp_chr_merged_with_chr.B2.txt"
for chr in {1..10}; do     file="chr_${chr}/chr${chr}_merged.B2.txt";     echo "Adding $file to $out with chr=$chr";          if [ $chr -eq 1 ]; then         awk -v c=$chr 'NR==1{print $0 "\tchr"; next} {print $0 "\t" c}' "$file" > "$out";     else         awk -v c=$chr 'NR>1{print $0 "\t" c}' "$file" >> "$out";     fi; done


# zipping the file for ease of loading in R studio
gzip -c imp_chr_merged_with_chr.B2.txt > imp_chr_merged_with_chr.B2.txt.gz

#filtering the top1% B2 scores and then sorting based on chromosome and position
sort -k3,3nr all_chr_merged_with_chr.B2.txt | head -n $(($(wc -l < all_chr_merged_with_chr.B2.txt )/100)) > wild.top1.txt
```


# landraces
creating the bed file for the top one perecent

```{bash}
#create top 1% threshold
sort -k4,4nr /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/Sorghum_landrace.AGPV5.B2_stat.txt | head -n $(($(wc -l < /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/Sorghum_landrace.AGPV5.B2_stat.txt )/100)) > /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.txt

#convert to bed file
awk -v OFS="\t" '{start=$2-40000; if(start<0) start=0; end=$2+40000; print $1, start, end, $4}' /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.txt > /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.40kb.bed

#sort the bed file according to the chromosomes
sort -k1,1V -k2,2n /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.40kb.bed > /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.40kb.sorted.bed

wc -l /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.40kb.sorted.bed
#147267

```


# Domestication trajectory
creating the bed files from the balancing selection results
```{bash}
#wild population 
#generating the bed file and sorting based on chromomes
awk -v OFS="\t" '{start=$1-40000; if(start<0) start=0; end=$1+40000; print $7, start, end, $3}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/wild.top1.txt > wild.40kb.bed

sort -k1,1V -k2,2n wild.40kb.bed > wild.40kb.sorted.bed

wc -l wild.40kb.sorted.bed 
#213176

#improved lines
awk -v OFS="\t" '{start=$1-40000; if(start<0) start=0; end=$1+40000; print $7, start, end, $3}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.top1.txt > /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.40kb.bed

sort -k1,1V -k2,2n imp.40kb.bed > imp.40kb.sorted.bed

wc -l imp.40kb.sorted.bed 
#94048
```

# looking for the overlap in bed files within bal

```{bash}
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input
ml bedtools
#merging the nearby positions and kept the highest value
bedtools merge -i wild.40kb.sorted.bed -d 40000 -c 4 -o max > wild.40kb.merged.bed
bedtools merge -i land.40kb.sorted.bed -d 40000 -c 4 -o max > land.40kb.merged.bed
bedtools merge -i imp.40kb.sorted.bed -d 40000 -c 4 -o max > imp.40kb.merged.bed

cd /work/jyanglab/subhash/sorgsd/bal_s/3.input
bedtools intersect -a wild.40kb.merged.bed -b land.40kb.merged.bed -wa -wb > W_L_overlap.bed
wc -l W_L_overlap.bed 
#137
bedtools intersect -a W_L_overlap.bed -b imp.40kb.merged.bed -wa -wb > W_L_I_overlap.bed
wc -l W_L_I_overlap.bed 
#60

#These 60 loci are conserved balancing selection targets the selective pressure that maintains variation in wild populations has persisted through domestication and improvement.
#They are high confidence candidates for important functional genes or regulatory regions.
###These are loci where genetic diversity is actively maintained across evolutionary and breeding processes.
```


# seeing the overlap between the bal and pos
```{bash}
bedtools intersect -a wild.merged.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/positive/wild_vs_landrace.bed -wa -wb > bal_wild_landrace.bed

bedtools intersect -a wild.merged.bed -b /mnt/nrdstor/jyanglab/subhash/datasets/3.results/positive/landrace_vs_improved.bed -wa -wb > 
bal_landrace_vs_improved.bed
```

what is loci?
A locus is a genomic region defined by merged B2 windows (+,- 40kb from physPos).
Each locus can include multiple SNPs or variants, but it is considered one signal of balancing selection.
By merging nearby windows (±40 kb and adjacent regions), we avoid counting the same signal multiple times.






```{bash}
/work/jyanglab/subhash/sorgsd/bal_s/3.input

awk '{OFS="\t"; print $NF, $1-1, $1, $3}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/wild.top1.txt > 0.1wild.sites.bed

sort -k1,1 -k2,2n 0.1wild.sites.bed > 0.01wild.sites.sorted.bed

bedtools intersect -a 0.01wild.sites.sorted.bed -b W_L_I_overlap.bed -wa > 0.001wild_in_overlap.bed

bedtools intersect -a 0.01wild.sites.sorted.bed -b W_L_I_overlap.bed -wa -wb > 0.0001wild_overlap_with_locus.bed

awk '{
    key = $5":"$6; 
    if($4>max[key]){max[key]=$4; line[key]=$0}
} END{for(k in line) print line[k]}' 0.0001wild_overlap_with_locus.bed > wild_leading_SNPs.bed
#60
```


Improved 

```{bash}
/work/jyanglab/subhash/sorgsd/bal_s/3.input

awk '{OFS="\t"; print $NF, $1-1, $1, $3}' /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/imp.top1.txt > 0.1imp.sites.bed
#94048 0.1imp.sites.bed
sort -k1,1 -k2,2n 0.1imp.sites.bed > 0.01imp.sites.sorted.bed
#94048 0.01imp.sites.sorted.bed
bedtools intersect -a 0.01imp.sites.sorted.bed -b W_L_I_overlap.bed -wa > 0.001imp_in_overlap.bed
#40620 0.001imp_in_overlap.bed
bedtools intersect -a 0.01imp.sites.sorted.bed -b W_L_I_overlap.bed -wa -wb > 0.0001imp_overlap_with_locus.bed
#40620 0.0001imp_overlap_with_locus.bed
awk '{
    key = $5":"$6; 
    if($4>max[key]){max[key]=$4; line[key]=$0}
} END{for(k in line) print line[k]}' 0.0001imp_overlap_with_locus.bed > imp_leading_SNPs.bed
#59 imp_leading_SNPs.bed
```

landraces

```{bash}
/work/jyanglab/subhash/sorgsd/bal_s/3.input

awk '{OFS="\t"; print $1, $2-1, $2, $4}' /mnt/nrdstor/jyanglab/subhash/datasets/3.results/B2/land.top1.txt > 0.1land.sites.bed
 wc -l 0.1land.sites.bed
#147267

sort -k1,1 -k2,2n 0.1land.sites.bed > 0.01land.sites.sorted.bed
 wc -l 0.01land.sites.sorted.bed
#147267

bedtools intersect -a 0.01land.sites.sorted.bed -b W_L_I_overlap.bed -wa > 0.001land_in_overlap.bed
#55858 
bedtools intersect -a 0.01land.sites.sorted.bed -b W_L_I_overlap.bed -wa -wb > 0.0001land_overlap_with_locus.bed

awk '{
    key = $5":"$6; 
    if($4>max[key]){max[key]=$4; line[key]=$0}
} END{for(k in line) print line[k]}' 0.0001land_overlap_with_locus.bed > land_leading_SNPs.bed
#57 
```


```{bash}
#!/bin/sh
#SBATCH --job-name=freq
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/frq_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/frq_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

cd /work/jyanglab/subhash/sorgsd/bal_s

ml vcftools
#vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/wild_50_samples.recode.vcf --freq --out wild_freq
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/landraces.recode.vcf --freq --out landrace_freq
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/imp_129_samples.recode.vcf --freq --out improved_freq
```

