---
title: "1.Balancing selection ananlysis (B2 score)"
output: html_notebook
author: "subhash"
date: "2025-08-13"
---
#make sure to load the modules on HCC unl

# Filtering the vcf files for the wild and improved lines along with S.propinqaum (ancestor allele)
using vcftools subset the 3 .vcf files for wild, improved and ancestor
also calculated the allele freequecny in ancestor(#3) samples

```{bash}
#!/bin/sh
#SBATCH --job-name=filter_missing
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/filter_missing_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/filter_missing_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml vcftools
#preparing the outgroup file
vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/propinqaum_sample.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample
#getting the frequencies
vcftools --vcf propinqaum_3sample.recode.vcf --freq --out /work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample
#preparing the wild lines file
vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/wild.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/wild_50_samples
#preparing the improved lines
vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/improved.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/imp_129_samples
```


submitting below R code as slurm job on HCC unl

```{bash}
#!/bin/sh
#SBATCH --job-name=ans
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml R
Rscript /work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans.R
```

## Calculating ancestor allele freq from the outgroup

```{r}
library(data.table)

d = fread("/work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample.frq", header=F, data.table=TRUE)
a1 = unlist(strsplit(d[[5]], ":"))
allele1 = a1[seq(1, length(a1), by=2)]
freq1 = as.numeric(a1[seq(2, length(a1), by=2)])
a2 = unlist(strsplit(d[[6]], ":"))
allele2 = a2[seq(1, length(a2), by=2)]
freq2 = as.numeric(a2[seq(2, length(a2), by=2)])

#after spliting giving the allele 1 and freq 1 and same for allele 2 and freq2

d1 = data.table(d[, 1:2], allele1, freq1, allele2, freq2)
setnames(d1, c("Chr", "Pos", "allele1", "Freq1", "allele2", "Freq2"))

#consider whichever allele is higher is considered to be ancestral allele
d1[, ancestral_allele := ifelse(Freq1 >= Freq2, allele1, allele2)]
fwrite(d1, "/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/Sorghum_ancestral_allele_V3.1.txt", sep = "\t")
```

### Calculating allele counts for both the wild and improved lines
To caculate how many derived alleles at each site

```{bash}
#!/bin/sh
#SBATCH --job-name=frq
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/2.frq/frq_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/2.frq/frq_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml vcftools
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/wild_50_samples.recode.vcf --counts --out /work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts
vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/imp_129_samples.recode.vcf --counts --out /work/jyanglab/subhash/sorgsd/bal_s/2.frq/imp_counts
```



```{bash}
#!/bin/sh
#SBATCH --job-name=inpu_w
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/3.input/inpW_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/3.input/inpW_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml R
Rscript /work/jyanglab/subhash/sorgsd/bal_s/3.input/in_w.R
```

#### Generating the input file for the BalLeRiMix 

This below script helps to calculate derived allele count (which is opposite of ancester allele)
```{r}
library(data.table)

anc <- fread("/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/Sorghum_ancestral_allele_V3.1.txt")

wild <- fread("/work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts.frq.count", skip = 1,
              col.names = c("Chr", "Pos", "N_alleles", "n", "allele1_count", "allele2_count"))

setnames(wild, c("Chr", "Pos", "N_alleles", "n", "count1", "count2"))

wild[, count1_num := as.integer(sub("^[A-Z]:", "", count1))]
wild[, count2_num := as.integer(sub("^[A-Z]:", "", count2))]

#chromsome by chromsome loop here to calculate derived allele count in each site ( opposite of ancester allel) 

for (chr in unique(anc$Chr)) {
  anc_chr <- anc[Chr == chr]
  wild_chr <- wild[Chr == chr]
  wild_merged_chr <- merge(anc_chr, wild_chr, by = c("Chr", "Pos"))
  wild_merged_chr[, x := ifelse(ancestral_allele == allele1, count2_num, count1_num)]
  wild_input_chr <- wild_merged_chr[, .(physPos = Pos, 
                                        genePos = Pos / 1e6, 
                                        x = x, 
                                        n = n)]
#writingg the input file for BalLeRMix
  fwrite(wild_input_chr, 
         paste0("/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild_BalLeRMix_input_chr", chr, ".txt"), 
         sep = "\t", col.names = FALSE)
}
```



```{bash}
#copying the Ballerimix input files (for each chromosome) in to different directories (by creating new ones) to keep the files organized and then filtering the empty rows
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild
for chr in {1..10}; do
    mkdir -p chr_${chr}
    # Copy original input
    cp imp_BalLeRMix_input_chr${chr}.txt chr_${chr}/
    # Filter out monomorphic sites
    awk '$3>0 && $3<$4' chr_${chr}/imp_BalLeRMix_input_chr${chr}.txt \
        > chr_${chr}/imp_BalLeRMix_input_chr${chr}.filtered.txt
done


#splitting the each chromosome file into 5k varinats for easy calculation of B2 score
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild
for chr in {1..10}; do
    cd chr_${chr}
    split -l 5000 --numeric-suffixes=1 --suffix-length=4 \
        imp_BalLeRMix_input_chr${chr}.filtered.txt chunk_
    cd ..
done

#generatig the spectrun DAF file for each chrmosome 
#SFS calculation from the BalLeriMix to use to calculate the B2 score
cd /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild
for chr in {1..10}; do
    python /mnt/nrdstor/jyanglab/subhash/software/BallerMix/software/BalLeRMix_v2.5.py \
        -i chr_${chr}/imp_BalLeRMix_input_chr${chr}.filtered.txt \
        --getSpect \
        --spect chr_${chr}/chr${chr}.spect.txt
done


#counting number of chunks in each chromosome (just for teh ease of jobsubimiting on HCC unl)
for chr in {1..10}; do
    echo "chr_${chr}: $(ls chr_${chr}/chunk_* | wc -l) chunks"
done
#number of chunk files in each chromosome folder in wild samples
#chr_1: 878 chunks
#chr_2: 1114 chunks
#chr_3: 1010 chunks
#chr_4: 906 chunks
#chr_5: 684 chunks
#chr_6: 730 chunks
#chr_7: 786 chunks
#chr_8: 810 chunks
#chr_9: 798 chunks
#chr_10: 822 chunks

#creating log files to get the out and error files for each job
for chr in {1..10}; do
    mkdir -p /work/jyanglab/subhash/sorgsd/bal_s/3.input/imp/chr_${chr}/logs
done
```

The below is the template I used to submit the BalLeriMix model to compute B2 score on each variant. using this template I can run many jobs to get the B2 stat score

```{bash}
#!/bin/bash
#SBATCH --job-name=chr${CHR}_wild
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=5:00:00
#SBATCH --mem=4G
#SBATCH --partition=batch,guest,jclarke
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/chr_${CHR}/logs/%A_%a.out.txt
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/chr_${CHR}/logs/%A_%a.err.txt
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu
#SBATCH --mail-type=END,FAIL

# Inputs
CHR=${CHR}   # passed via sbatch --export
CHUNK_DIR=/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild/chr_${CHR}
SOFTWARE=/mnt/nrdstor/jyanglab/subhash/software/BallerMix/software/BalLeRMix_v2.5.py

# Format ID like 0001, 0002...
ID=$(printf "%04d" ${SLURM_ARRAY_TASK_ID})
FILE=$CHUNK_DIR/chunk_${ID}

echo ">>> Processing chromosome $CHR, chunk $ID"

python $SOFTWARE \
   -i "$FILE" \
   --spect $CHUNK_DIR/chr${CHR}.spect.txt \
   --physPos \
   -o "${FILE}.B2.txt"

echo ">>> Finished chromosome $CHR, chunk $ID$ CHUNK_DIR/chr${CHR}.spect.txt"

```

job submission by chromosome to chromsome

```{bash}
sbatch --export=CHR=1 --array=1-878 chr_array.slurm
sbatch --export=CHR=2 --array=1-1114 chr_array.slurm
sbatch --export=CHR=3 --array=1-1010 chr_array.slurm
sbatch --export=CHR=4 --array=1-906 chr_array.slurm
sbatch --export=CHR=5 --array=1-684 chr_array.slurm
sbatch --export=CHR=6 --array=1-730 chr_array.slurm
sbatch --export=CHR=7 --array=1-786 chr_array.slurm
sbatch --export=CHR=8 --array=1-810 chr_array.slurm
sbatch --export=CHR=9 --array=1-798 chr_array.slurm
sbatch --export=CHR=10 --array=1-822 chr_array.slurm
```


# these are mischlenus code from the begginer 
I use them to track my running jobs
since I submitted toomany jobs to see and track using naked eye 
the below code lines will help to 

#misch
here I did subset to work on small sample set to test the pipeline once, it sucessfully generated the output files I have run the whole pipeline on large dataset

```{bash}
head -n 101 /work/jyanglab/subhash/sorgsd/bal_s/Sorghum_ancestral_allele_V3.1.txt > /work/jyanglab/subhash/sorgsd/bal_s/subset/anc_subset.txt
head -n 101 /work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts.frq.count > /work/jyanglab/subhash/sorgsd/bal_s/subset/wild_counts_subset.frq.count
head -n 101 imp_counts.frq.count > /work/jyanglab/subhash/sorgsd/bal_s/subset/imp_counts_subset.frq.count
head -n 1000 /work/jyanglab/subhash/sorgsd/bal_s/subset/BalLeRMix_input_chr1.filtered.txt > 1000.txt
```

```{bash}
squeue -j 11541491 | wc -l
sacct -j 11545344 --format=JobID,State
squeue -j 11545344 -o "%.18i %.9T" | wc -l
```

# merging all the chunk output files into one big file

```{bash}
#merging all the chunks in each chromosome as one file

#for chr in {1..10}; do     out="chr_${chr}/chr${chr}_merged.B2.txt";     echo "Merging chromosome $chr → $out";     cat chr_${chr}/chunk_*.B2.txt > "$out"; done
#bug fix: getting all the headers from each chunk file and then getting the unique one and adding it to the merged file
#cd /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild
#grep -n "^physPos" all_chr_merged_with_chr.B2.txt

for chr in {1..10}; do
  in_dir="chr_${chr}"
  out="${in_dir}/chr${chr}_merged.B2.txt"
  echo "Merging chromosome ${chr} -> ${out}"

  { 
    # header once
    head -n 1 ${in_dir}/chunk_0001.B2.txt | awk -v c=$chr 'BEGIN{OFS="\t"} {print $0,"chr"}'
    # all data lines from all chunks (skip each header), append chr
    for f in ${in_dir}/chunk_*.B2.txt; do
      awk -v c=$chr 'NR>1{print $0 "\t" c}' "$f"
    done
  } > "$out"
done
```




```{bash}
#merging all the chromsomes in to one big file buy adding an extra column called chromosme
out="all_chr_merged_with_chr.B2.txt"

for chr in {1..10}; do file="chr_${chr}/chr${chr}_merged.B2.txt" echo "Adding $file -> $out" if [ "$chr" -eq 1 ]; then cat "$file" > "$out"          # keep header
  else
    tail -n +2 "$file" >> "$out"  # skip header, append data
  fi
done

# zipping the file for ease of loading in R studio
gzip -c all_chr_merged_with_chr.B2.txt > all_chr_merged_with_chr.B2.txt.gz

#filtering the top1% B2 scores and then sorting based on chromosome and position
(head -n 1 all_chr_merged_with_chr.B2.txt && tail -n +2 all_chr_merged_with_chr.B2.txt | sort -k3,3nr | head -n $(( ($(wc -l < all_chr_merged_with_chr.B2.txt) - 1) / 100 )) ) > wild.top1.txt

wc -l all_chr_merged_with_chr.B2.txt
#21313439
wc -l wild.top1.txt 
#213135 

```

