---
title: "R Notebook"
output: html_notebook
---

# Filtering the vcf files for the wild and improved lines along with S.propinqaum (ancestor allele)
using vcftools subset the 3 .vcf files for wild, improved and ancestor
also calculated the allele freequecny in ancestor(#3) samples

```{bash}
#!/bin/sh
#SBATCH --job-name=filter_missing
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/filter_missing_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/filter_missing_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml vcftools

vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/propinqaum_sample.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample

vcftools --vcf propinqaum_3sample.recode.vcf --freq --out /work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample

vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/wild.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/wild_50_samples

vcftools --vcf /work/jyanglab/subhash/sorgsd/gen_v5/SorGSD.289snp.imputed.V5.vcf --keep /work/jyanglab/subhash/sorgsd/filterfed_vcf/improved.txt --recode --out /work/jyanglab/subhash/sorgsd/bal_s/imp_129_samples
```


submitting below R code as slurm job on HCC unl
```{bash}
#!/bin/sh
#SBATCH --job-name=ans
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml R
Rscript /work/jyanglab/subhash/sorgsd/bal_s/ans_freq/ans.R
```

## Calculating ancestor allele freq and count for all the chromosomes at once from the outgroup

```{r}
library(data.table)

d = fread("/work/jyanglab/subhash/sorgsd/bal_s/propinqaum_3sample.frq", header=F, data.table=TRUE)
a1 = unlist(strsplit(d[[5]], ":"))
allele1 = a1[seq(1, length(a1), by=2)]
freq1 = as.numeric(a1[seq(2, length(a1), by=2)])
a2 = unlist(strsplit(d[[6]], ":"))
allele2 = a2[seq(1, length(a2), by=2)]
freq2 = as.numeric(a2[seq(2, length(a2), by=2)])

#after spliting giving the allele 1 and freq 1 and same for allele 2 and freq2

d1 = data.table(d[, 1:2], allele1, freq1, allele2, freq2)
setnames(d1, c("Chr", "Pos", "allele1", "Freq1", "allele2", "Freq2"))

#consider whichever allele is higher is considered to be ancestral allele
d1[, ancestral_allele := ifelse(Freq1 >= Freq2, allele1, allele2)]
fwrite(d1, "/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/Sorghum_ancestral_allele_V3.1.txt", sep = "\t")
```

### Calculating allele counts for both the wild and improved lines
To caculate how many derived alleles at each site

```{r}
#!/bin/sh
#SBATCH --job-name=frq
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/2.frq/frq_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/2.frq/frq_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu
ml vcftools

vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/wild_50_samples.recode.vcf --counts --out /work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts

vcftools --vcf /work/jyanglab/subhash/sorgsd/bal_s/imp_129_samples.recode.vcf --counts --out /work/jyanglab/subhash/sorgsd/bal_s/2.frq/imp_counts
```



```{bash}
#!/bin/sh
#SBATCH --job-name=inpu_w
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=120:00:00
#SBATCH --mem=10G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/3.input/inpW_err.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/3.input/inpW_out.txt
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

ml R
Rscript /work/jyanglab/subhash/sorgsd/bal_s/3.input/in_w.R
```


here I did subset to work on small sample set to test the pipeline once, it sucessfully generated the output files I have run the whole pipeline on large dataset

```{bash}
head -n 101 /work/jyanglab/subhash/sorgsd/bal_s/Sorghum_ancestral_allele_V3.1.txt > /work/jyanglab/subhash/sorgsd/bal_s/subset/anc_subset.txt
head -n 101 /work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts.frq.count > /work/jyanglab/subhash/sorgsd/bal_s/subset/wild_counts_subset.frq.count
head -n 101 imp_counts.frq.count > /work/jyanglab/subhash/sorgsd/bal_s/subset/imp_counts_subset.frq.count


head -n 1000 /work/jyanglab/subhash/sorgsd/bal_s/subset/BalLeRMix_input_chr1.filtered.txt > 1000.txt
```

#### Generating the input file for the BalLeRiMix 

This below script helps to calculate derived allele count (which is opposite of ancester allele)
```{r}
library(data.table)

anc <- fread("/work/jyanglab/subhash/sorgsd/bal_s/ans_freq/Sorghum_ancestral_allele_V3.1.txt")

wild <- fread("/work/jyanglab/subhash/sorgsd/bal_s/2.frq/wild_counts.frq.count", skip = 1,
              col.names = c("Chr", "Pos", "N_alleles", "n", "allele1_count", "allele2_count"))

setnames(wild, c("Chr", "Pos", "N_alleles", "n", "count1", "count2"))

wild[, count1_num := as.integer(sub("^[A-Z]:", "", count1))]
wild[, count2_num := as.integer(sub("^[A-Z]:", "", count2))]

#chromsome by chromsome loop here to calculate derived allele count in each site ( opposite of ancester allel) 

for (chr in unique(anc$Chr)) {
  anc_chr <- anc[Chr == chr]
  wild_chr <- wild[Chr == chr]
  wild_merged_chr <- merge(anc_chr, wild_chr, by = c("Chr", "Pos"))
  wild_merged_chr[, x := ifelse(ancestral_allele == allele1, count2_num, count1_num)]
  wild_input_chr <- wild_merged_chr[, .(physPos = Pos, 
                                        genePos = Pos / 1e6, 
                                        x = x, 
                                        n = n)]
#writingg the input file for BalLeRMix
  fwrite(wild_input_chr, 
         paste0("/work/jyanglab/subhash/sorgsd/bal_s/3.input/wild_BalLeRMix_input_chr", chr, ".txt"), 
         sep = "\t", col.names = FALSE)
}
```

filtering the missing count values 
```{bash}
for f in /work/jyanglab/subhash/sorgsd/bal_s/3.input/wild_BalLeRMix_input_chr*.txt; do
  awk '$3 > 0' "$f" > "${f}.filtered"
done
```

##### Running the BalLeRMix on each chromsome  to generate B2 score at each site

```{r}
#!/bin/sh
#SBATCH --job-name=wild_array
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --time=160:00:00
#SBATCH --mem=20G
#SBATCH --partition=jclarke,batch,guest
#SBATCH --error=/work/jyanglab/subhash/sorgsd/bal_s/4.b2/wild/W_err_%A_%a.txt
#SBATCH --output=/work/jyanglab/subhash/sorgsd/bal_s/4.b2/wild/W_out_%A_%a.txt
#SBATCH --array=1-10
#SBATCH --mail-type=ALL
#SBATCH --mail-user=smahamkalivenkatas2@huskers.unl.edu

# Load variables
CHR=${SLURM_ARRAY_TASK_ID}
# Paths
INPUT_DIR=/work/jyanglab/subhash/sorgsd/bal_s/3.input
OUT_DIR=/work/jyanglab/subhash/sorgsd/bal_s/4.b2/wild
SOFT_DIR=/mnt/nrdstor/jyanglab/subhash/software/BallerMix/software
# Filenames
INPUT_FILE=${INPUT_DIR}/wild_BalLeRMix_input_chr${CHR}.txt.filtered
DAF_FILE=${OUT_DIR}/Chr${CHR}.wild_49_samples.propinquum.DAF.txt
B2_FILE=${OUT_DIR}/Chr${CHR}.wild_49samples.propinquum.B2_stat.txt
# Step 1: Get spectrum
python ${SOFT_DIR}/BalLeRMix_v2.5.py -i ${INPUT_FILE} --getSpect --spect ${DAF_FILE}
# Step 2: Run B2 statistic
python ${SOFT_DIR}/BalLeRMix_v2.5.py -i ${INPUT_FILE} --spect ${DAF_FILE} --physPos -o ${B2_FILE}
```


