---
title: "2020 and 2021 phenotype distribution"
author: "subhash"
date: "2024-11-13"
output: html_document
---



#pathsetup

```{r setup}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('/Users/subhashmahamkali/Documents/gwas_sap')) 
```

#loading all the necessary libraries

```{r}
#loading libraries
library(dplyr)
library(tidyverse)
library(data.table)
library(ggridges)
library(patchwork)
```

#loading the phenotype dataset and sub-population of SAP

```{r}
#sub-populations in sorghum association panel (SAP)
cluster = read.table("data/0.SAP/sorted_cluster_data.txt", header = T)
cluster <- cluster %>% rename(genotype = PI_numbers)
cluster$Cluster <- as.numeric(cluster$Cluster)

cluster <- cluster %>%
  mutate(Cluster = case_when(
    Cluster == 1 ~ "durra",
    Cluster == 2 ~ "kafir",
    Cluster == 3 ~ "caudatum",
    Cluster == 4 ~ "guinea",
    Cluster == 5 ~ "milo/durra-bicolor",
    Cluster == 6 ~ "mixed/bicolor",
    TRUE ~ as.character(Cluster)  # Keep original value for unmatched cases
  ))
write.table(cluster, "data/0.SAP/cluster.txt")

k = read.table("data/0.SAP/cluster.txt", header = T)
```


```{r}
#2020 and 2021 phenotype dataset

data = read.csv("data/1.Phenotype_data/1.2020_2021_SAP/1.BLUEs_SAP_2020_2021.csv", check.names = F)
str(data)

d2021 = data[,c (1,47:66,87:91)]
d2020 = data[,c (1:46, 67:86)]
d = data[,c(1,92:119)]
write.table(d, file = "data/1.Phenotype_data/1.2020_2021_SAP/1.B.28_traits_B.txt", sep = "\t", row.names = TRUE, col.names = TRUE)


#making seprate dataset for the 2020
data$genotype <- gsub(" ", "", data$genotype) #2020 dataset
num_2020_LN <- length(grep("^2020:LN:", names(data))) #20
num_2020_HN <- length(grep("^2020:HN:", names(data))) #25
num_2020_NR <- length(grep("2020:.*:NR", names(data))) #20

#match("2021:stemDiameterUpper:NR", names(data))#To count the column number.
d2020 = data[,c (1:46, 67:86)]
#names(d2020)
column_names <- names(d2020)
```

###matching all the HN and LN traits to compare and calculating nitrogen response

```{r}
#2020

#identifying the commmon traits to do phenotype distribution 
# Extract the traits by removing the prefixes "2020:LN:", "2020:HN:", and suffix ":NR"
traits_LN <- sub("^2020:LN:", "", grep("^2020:LN:", column_names, value = TRUE))
traits_HN <- sub("^2020:HN:", "", grep("^2020:HN:", column_names, value = TRUE))
traits_NR <- sub("^2020:", "", sub(":NR$", "", grep(":NR$", column_names, value = TRUE)))

# Find the common traits that are present in all three conditions
common_traits <- intersect(intersect(traits_LN, traits_HN), traits_NR)

# Generate the full column names for each common trait under all three conditions
matched_columns <- unlist(lapply(common_traits, function(trait) {
  c(paste0("2020:LN:", trait), paste0("2020:HN:", trait), paste0("2020:", trait, ":NR"))
}))

# Include the first column (genotype) in the final selection
columns_to_keep <- c("genotype", matched_columns)

# Subset d2020 to include only the selected columns
d2020 <- d2020[, columns_to_keep]
# Display the structure of the filtered data frame
str(d2020)
names(d2020) <- sub("2020:(.*):NR", "2020:NR:\\1", names(d2020)) #matching the same format

d2020_long <- d2020 %>%
  left_join(cluster, by = "genotype") %>%
  pivot_longer(cols = -c(genotype, Cluster), 
               names_to = c("Year", "Condition", "Trait"), 
               names_sep = ":", 
               values_to = "Value") %>%
  filter(!is.na(Cluster)) %>%  # Remove rows with NA in Cluster
  arrange(Trait, Condition)

```




```{r}
# List of traits in the desired order
desired_traits <- c("flagLeafLength", "flagLeafWidth", "leafAngleStandardDeviation", "thirdLeafLength", "thirdLeafWidth",
                    "extantLeafNumber","medianLeafAngle",
                    "panicleGrainWeight","paniclesPerPlot","rachisDiameterUpper","rachisDiameterLower","rachisLength",
                    "stemDiameterLower", "stemDiameterUpper",
                    "daysToFlower", "plantHeight", "estimatedPlotYield", "tillersPerPlant", "primaryBranchNumber", "branchInternodeLength")

# Generate the desired column order
desired_order <- c("genotype", 
                   unlist(lapply(desired_traits, function(trait) {
                     c(paste0("2020:LN:", trait), paste0("2020:HN:", trait), paste0("2020:NR:", trait))
                   })))

# Rearrange the data frame to match the desired order
d2020 <- d2020[, desired_order]

# Verify the new column order
names(d2020)

# Convert the Trait column to a factor with levels in the desired order
d2020_long <- d2020_long %>%
  mutate(Trait = factor(Trait, levels = desired_traits)) %>%  # Arrange traits in desired order
  arrange(Trait, Condition)  # Arrange rows by Trait and Condition


```

#function to create Phenotype distribution (densityplots) with HN, LN and NR.

```{r}
# Function to create density plots for each trait with HN/LN in one panel and NR in another
create_density_plot <- function(data, trait) {
  ggplot(data %>% filter(Trait == trait), aes(x = Value, fill = Condition)) +
    geom_density(alpha = 0.6) +
    labs(title = paste(unique(data$Year), "-", trait),
         x = "BLUEs Value", y = "Density") +
    theme_minimal() +
    scale_fill_manual(values = c("LN" = "green", "HN" = "red", "NR" = "yellow")) +
    facet_wrap(~ Condition %in% c("NR"), scales = "free", labeller = as_labeller(c(`FALSE` = "HN & LN", `TRUE` = "NR"))) +
    theme(
      plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
      legend.position = "right",
      legend.title = element_text(size = 18, face = "bold"),
      legend.text = element_text(size = 14, face = "bold"),
      axis.title.x = element_text(size = 18, face = "bold"),
      axis.title.y = element_text(size = 18, face = "bold"),
      axis.text.x = element_text(size = 18, face = "bold"),
      axis.text.y = element_text(size = 18, face = "bold")
    ) +
    guides(fill = guide_legend(title = "Condition"))
}
```


#fucntion to create sub-population level phenotypic distribution for the HN and LN 
```{r}
create_grouped_ridge_plot_with_corrected_vlines <- function(data, trait) {
  # Perform t-tests and calculate means
  t_test_results <- data %>%
    group_by(Cluster) %>%
    summarise(
      mean_HN = mean(Value[Condition == "HN"], na.rm = TRUE),
      mean_LN = mean(Value[Condition == "LN"], na.rm = TRUE),
      p_value = t.test(Value[Condition == "HN"], Value[Condition == "LN"])$p.value
    )
  
  # Add significance stars
  t_test_results <- t_test_results %>%
    mutate(
      significance = case_when(
        p_value < 0.001 ~ "***",
        p_value < 0.01 ~ "**",
        p_value < 0.05 ~ "*",
        TRUE ~ ""
      )
    )
  
  # Merge means into the original data for vline
  data <- data %>%
    left_join(t_test_results %>% select(Cluster, mean_HN, mean_LN), by = "Cluster")
  
  # Create the ridge plot
  ggplot(data %>% filter(Trait == trait), aes(x = Value, y = factor(Cluster), fill = Condition)) +
    geom_density_ridges(scale = 1, alpha = 0.6) +
    # Add mean lines restricted to each ridge and correct alignment
    geom_segment(
      data = t_test_results,
      aes(x = mean_HN, xend = mean_HN, y = as.numeric(factor(Cluster)) - 0.2, 
          yend = as.numeric(factor(Cluster)) + 0.2),
      linetype = "dotted", color = "red", size = 1, inherit.aes = FALSE
    ) +
    geom_segment(
      data = t_test_results,
      aes(x = mean_LN, xend = mean_LN, y = as.numeric(factor(Cluster)) - 0.2, 
          yend = as.numeric(factor(Cluster)) + 0.2),
      linetype = "dotted", color = "green", size = 1, inherit.aes = FALSE
    ) +
    geom_text(
      data = t_test_results,
      aes(x = max(data$Value, na.rm = TRUE) * 0.95, y = factor(Cluster), label = significance),
      color = "black", size = 5, hjust = 1, inherit.aes = FALSE
    ) +
    labs(
      title = paste(unique(data$Year), "-", trait),
      x = "BLUEs Value", y = "Clusters"
    ) +
    theme_minimal() +
    scale_fill_manual(values = c("LN" = "green", "HN" = "red")) +
    theme(
      plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
      legend.position = "right",
      legend.title = element_text(size = 18, face = "bold"),
      legend.text = element_text(size = 14, face = "bold"),
      axis.title.x = element_text(size = 18, face = "bold"),
      axis.title.y = element_text(size = 18, face = "bold"),
      axis.text.x = element_text(size = 18, face = "bold"),
      axis.text.y = element_text(size = 18, face = "bold")
    ) +
    guides(fill = guide_legend(title = "Condition"))
}
```


#plots with HN, LN and NR (individual plots) fro the year 2020
```{r}
# Define the directory where you want to save the plots
save_path <- "graphs/1.phenotype_distribution/2020/1.individual_trait_pheno_dist/"

# Loop through each trait, create the density plot, and save it
traits <- unique(d2020_long$Trait)

for (trait in traits) {
  # Filter for the current trait and get the year for the filename
  plot_data <- d2020_long %>% filter(Trait == trait)
  year <- unique(plot_data$Year)
  # Generate the density plot
  plot <- create_density_plot(plot_data, trait)
  # Define the filename with the specified path
  filename <- paste0(save_path, year, "_", trait, ".png")
  # Save the plot
  ggsave(filename = filename, plot = plot, width = 10, height = 6)
  print(paste("Saved:", filename))  # Print confirmation of saved file
}

```



```{r}
ridge_plots <- list()
save_path = "graphs/1.phenotype_distribution/2020/2.clustered_ridges_individual/"

for (trait in traits) {
  # Filter data for the current trait
  plot_data <- d2020_long %>% filter(Trait == trait, Condition %in% c("HN", "LN"))
  year <- unique(plot_data$Year)
  # Generate the ridge plot
  ridge_plot <- create_grouped_ridge_plot_with_corrected_vlines(plot_data, trait)
  # Save the individual plot
  filename <- paste0(save_path, "/", year, "_ind_ridge_", trait, ".png")
  ggsave(filename = filename, plot = ridge_plot, width = 10, height = 6)
  # Print confirmation for individual plot
  print(paste("Saved:", filename))
  # Add the plot to the list for multipanel
  ridge_plots[[trait]] <- ridge_plot
}
combined_plot <- wrap_plots(ridge_plots, ncol = 5) +
plot_layout(guides = "collect")
filename_combined <- paste0("graphs/1.phenotype_distribution/2020/", year, "_combined_ridge_plots.pdf")
ggsave(filename = filename_combined, plot = combined_plot, width = 40, height =20 )
# Print confirmation for multipanel plot
print(paste("Saved combined multipanel plot:", filename_combined))

```


#for the year 2021
```{r}
d2021 = data[,c (1,47:66,87:91)]
names(d2021)

#2021 datasets
num_2021_LN <- length(grep("^2021:LN:", names(data))) #5
num_2021_HN <- length(grep("^2021:HN:", names(data))) #15
num_2021_NR <- length(grep("2021:.*:NR", names(data))) #5
column_names <- names(d2021)
```

```{r}
# Extract the traits by removing the prefixes "2020:LN:", "2020:HN:", and suffix ":NR"
traits_LN <- sub("^2021:LN:", "", grep("^2021:LN:", column_names, value = TRUE))
traits_HN <- sub("^2021:HN:", "", grep("^2021:HN:", column_names, value = TRUE))
traits_NR <- sub("^2021:", "", sub(":NR$", "", grep(":NR$", column_names, value = TRUE)))

# Find the common traits that are present in all three conditions
common_traits <- intersect(intersect(traits_LN, traits_HN), traits_NR)

# Generate the full column names for each common trait under all three conditions
matched_columns <- unlist(lapply(common_traits, function(trait) {
  c(paste0("2021:LN:", trait), paste0("2021:HN:", trait), paste0("2021:", trait, ":NR"))
}))

# Include the first column (genotype) in the final selection
columns_to_keep <- c("genotype", matched_columns)

# Subset d2020 to include only the selected columns
d2021 <- d2021[, columns_to_keep]
# Display the structure of the filtered data frame
str(d2021)
names(d2021) <- sub("2021:(.*):NR", "2021:NR:\\1", names(d2021)) #matching the same format

d2021_long <- d2021 %>%
  left_join(cluster, by = "genotype") %>%
  pivot_longer(cols = -c(genotype, Cluster), 
               names_to = c("Year", "Condition", "Trait"), 
               names_sep = ":", 
               values_to = "Value") %>%
  filter(!is.na(Cluster)) %>%  # Remove rows with NA in Cluster
  arrange(Trait, Condition)
```



```{r}
#2021
# Define the directory where you want to save the plots
save_path <- "graphs/1.phenotype_distribution/2021/1.individual_trait_pheno_dist/"

# Loop through each trait, create the density plot, and save it
traits <- unique(d2021_long$Trait)

for (trait in traits) {
  # Filter for the current trait and get the year for the filename
  plot_data <- d2021_long %>% filter(Trait == trait)
  year <- unique(plot_data$Year)
  
  # Generate the density plot
  plot <- create_density_plot(plot_data, trait)
  
  # Define the filename with the specified path
  filename <- paste0(save_path, year, "_", trait, ".png")
  
  # Save the plot
  ggsave(filename = filename, plot = plot, width = 10, height = 8)
  
  print(paste("Saved:", filename))  # Print confirmation of saved file
}

```



```{r}
ridge_plots <- list()
save_path = "graphs/1.phenotype_distribution/2021/2.clustered_ridges_individual/"

for (trait in traits) {
  # Filter data for the current trait
  plot_data <- d2021_long %>% filter(Trait == trait, Condition %in% c("HN", "LN"))
  year <- unique(plot_data$Year)
  # Generate the ridge plot
  ridge_plot <- create_grouped_ridge_plot_with_corrected_vlines(plot_data, trait)
  # Save the individual plot
  filename <- paste0(save_path, "/", year, "_ind_ridge_", trait, ".png")
  ggsave(filename = filename, plot = ridge_plot, width = 10, height = 6)
  # Print confirmation for individual plot
  print(paste("Saved:", filename))
  # Add the plot to the list for multipanel
  ridge_plots[[trait]] <- ridge_plot
}
combined_plot <- wrap_plots(ridge_plots, ncol = 3) +
plot_layout(guides = "collect")
filename_combined <- paste0("graphs/1.phenotype_distribution/2021/", year, "_combined_ridge_plots.pdf")
ggsave(filename = filename_combined, plot = combined_plot, width = 20, height =12)
# Print confirmation for multipanel plot
print(paste("Saved combined multipanel plot:", filename_combined))

```
